### 2015-09-04 write_up.Rmd
want original causality mindmap to work off of

---
title: "Distilling Causality from Correlation in Large-scale Omics Datasets"
author: "Robert Ness"
date: "September 3, 2015"
output: html_document
---

Our goal in molecular cell biology is causal inference -- meaning uncovering the regulatory relationships between components of the cell that determine phenotype and disease states.  Generally observing statistical association between such component does not demonstrate causality.  The exception is in the context of a sound experimental design, where we can safely use statistical association as evidence for the presence of a causal relation.

High-throughput omics technologies enable us to conduct large-scale experiments, where we simultaneously measure multiple cellular components in a sample, and in many cases acquire a large amount of data for each component.  However, despite generating much more data, these large-scale experiments make causal inference much more difficult.  Large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.  We can address this problem by improving our experimental design.

## How we infer causality from experiments

The signaling proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.

Assuming this causal relation were not known, the question of whether Mek has a causal effect on Erk is evaluated with a designed experiment, where Mek is the treatment and Erk is the response.  Parts of the experimental design include intervention and randomization.  The intervention targets the potential cause, Mek, and fixes it at different levels, for example 'high activity' and 'baseline activity'.  Randomization refers to random assignment of samples or subjects to each level, helping to alleviate bias introduced by unseen factors.  The experiment generates measurements that will show a strong statistical association between Mek and Erk.  Intervention and randomization enable us to interpret that statistical association as evidence for Mek's causal influence on Erk.

With omics technologies, instead of measuring one response, we simultaneous measure many.  Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each measured features.  As with the simple one-response experimental design, we seek to use these associations to evaluate the presence of causal relations.  The challenge with large datasets is that many of these associations will be spurious.  The more spurious associations, the more likely we are to incorrectly conclude a causal relation exists based on those associations.

We can remedy this problem in two ways; introducing more interventions, and introducing more prior biological knowledge.

## How spurious statistical associations increase with the number of features measured

The following simulation illustrates how, as the number of things we measure increases, the number of spurious associations increase.  In each instance of the simulation, we simulate Gaussian measurements for 10 features and 100 features, and each set of feature measurements is completely independent of all others.  Using Pearson correlation to quantify association, since all the features are independent, any correlation we find between a pair of features is completely spurious.  We record the maximum correlation between the 10 feature set and the 100 feature set. We repeat this 500 times.

```{r, echo=FALSE, eval=FALSE}
set_20 <- rep(0, 500)
set_500 <- rep(0, 500)
for(i in 1:500){
  x <- matrix(rnorm(100 * 20), ncol = 20)
  cor_x <- cor(x)
  diag(cor_x) <- 0
  set_20[i] <- max(cor_x)
  y <- matrix(rnorm(100 * 500), ncol = 500)
  cor_y <- cor(y)
  diag(cor_y) <- 0
  set_500[i] <- max(cor_y)
}
hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL, xlim = c(min(set_20) , max(set_500) + .1 ), 
     main = "Highest Spurious Correlation")
hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
legend("topr", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
```

<figure>
  <img src="http://i.imgur.com/CSuQdePl.png" alt="spurious1">
  <figcaption>Fig1. Increasing the number of measured features from 20 to 100 makes it much more likely you will find correlations that are high but spurious. </figcaption>
</figure>

Another way of saying two objects have a statistical association is to say they are not statistically independent.  However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing.  A good example is genetics. President Barak Obama and his father Obama Sr.'s genetic profile both have a strong association with that of his daughter Malia Obama, but if you already know Barak Obama's profile, knowing Obama Sr.'s profile will give you no additional information about Malia's profile -- Malia's genetic profile is conditionally independent of Obama Sr.'s profile, given you know the Barak Obama's profile. Figure 2 continues with the MAPK example;

<figure>
  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
  <figcaption>Fig2. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
  Top; the lines represent correlation - the activity of the kinases on this pathway are all correlation. Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk. They are conditionally independent.</figcaption>
</figure>

Though approaches vary, state-of-the-art algorithms for determining causal relationships between several simultaneously measured features all attempt to reduce the set of associations by searching for cases of conditional independence, resulting in a far sparser set of conditional dependence relationships.  However, large-scale experiments make this search task more difficult.  To illustrate, we repeat the previous simulation, and instead of finding the max Pearson correlation in each instance, we apply a search algorithm that iteratively performs conditional independence tests, returning a sparse set of conditional dependence relationships.

```{r, echo=FALSE, eval = FALSE}
library(bnlearn)
set_20 <- rep(0, 10)
set_500 <- rep(0, 10)
for(i in 1:length(set_10)){
  print(i)
  x <- as.data.frame(matrix(rnorm(500 * 20), ncol = 20))
  set_10[i] <- narcs(gs(x, undirected = TRUE))
  y <- as.data.frame(matrix(rnorm(500 * 500), ncol = 500))
  set_100[i] <- narcs(gs(y, undirected = TRUE))
}
hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
     main = "Counts of Conditional Dependence Relationships")
hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
legend("top", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
```

Simulation running, image placeholder.

## Use of prior knowledge and interventions

### 2015-09-04 write_up.Rmd
finished core argument

@@
 Our goal in molecular cell biology is causal inference -- meaning uncovering the regulatory relationships between components of the cell that determine phenotype and disease states.
-Generally observing statistical association between such component does not demonstrate causality.
-The exception is in the context of a sound experimental design, where we can safely use statistical association as evidence for the presence of a causal relation.
+Generally observing statistical association between such components does not demonstrate causality.
+However, we can safely use statistical association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
 
 High-throughput omics technologies enable us to conduct large-scale experiments, where we simultaneously measure multiple cellular components in a sample, and in many cases acquire a large amount of data for each component.
-However, despite generating much more data, these large-scale experiments make causal inference much more difficult.
+However, despite generating much more data, these large-scale experiments make the task of causal inference much more difficult.
 Large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
@@
 
-## How we infer causality from experiments
+## How we infer causality from omics experiments
 
-The signaling proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.
+The MAPK signaling pathway proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.
 
 Assuming this causal relation were not known, the question of whether Mek has a causal effect on Erk is evaluated with a designed experiment, where Mek is the treatment and Erk is the response.
-Parts of the experimental design include intervention and randomization.
-The intervention targets the potential cause, Mek, and fixes it at different levels, for example 'high activity' and 'baseline activity'.
-Randomization refers to random assignment of samples or subjects to each level, helping to alleviate bias introduced by unseen factors.
+Intervention and randomization are key parts of the experimental design.
+Intervention means we target the potential cause, Mek, and fix it at different levels, for example 'high activity' and 'baseline activity'.
+Randomization refers to random assignment of samples or subjects to each of these levels, helping to alleviate bias introduced by unseen factors.
 The experiment generates measurements that will show a strong statistical association between Mek and Erk.
@@
 With omics technologies, instead of measuring one response, we simultaneous measure many.
-Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each measured features.
+Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each of the measured features.
+
 As with the simple one-response experimental design, we seek to use these associations to evaluate the presence of causal relations.
-The challenge with large datasets is that many of these associations will be spurious.
-The more spurious associations, the more likely we are to incorrectly conclude a causal relation exists based on those associations.
+Saying two objects have a statistical association is the same as saying they are not statistically independent.
+However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing.
+Figure 1 illustrates this principle with the MAPK proteins.
 
-We can remedy this problem in two ways; introducing more interventions, and introducing more prior biological knowledge.
+<figure>
+  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
+  <figcaption>Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
+  Top; the lines represent associate - the activity of the kinases on this pathway are all correlated.
+Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk.
+They are conditionally independent.</figcaption>
+</figure>
 
-## How spurious statistical associations increase with the number of features measured
+We use algorithms borrowed from the causal Bayesian network modeling domain to do causal inference in the context of an experiment with multiple intercorrelated response variables.
+These  algorithms' approaches vary but they essentially do two things; (1) find cases of conditional independence and reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use the experiment's intervention information to evaluate those conditional dependencies as evidence for potential causal relationtions.
 
-The following simulation illustrates how, as the number of things we measure increases, the number of spurious associations increase.
-In each instance of the simulation, we simulate Gaussian measurements for 10 features and 100 features, and each set of feature measurements is completely independent of all others.
-Using Pearson correlation to quantify association, since all the features are independent, any correlation we find between a pair of features is completely spurious.
-We record the maximum correlation between the 10 feature set and the 100 feature set.
-We repeat this 500 times.
+The challenge with large datasets is that by measuring more things, we increase the number of spurious statistical associations, and thus the number of spurious cases of conditional dependence, and thus the number of incorrect conclusions of the presence of a causal relation.
+To illustrate, we ran a simulation where first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.
+In each case the features are completely independent.
+This means if we quantify association using Pearson correlation, any correlation we find will be completely spurious.
+We record the maximum correlation between the 20 feature set and the 500 feature set.
+We repeat this simulation 500 times.
 
@@
   <img src="http://i.imgur.com/CSuQdePl.png" alt="spurious1">
-  <figcaption>Fig1. Increasing the number of measured features from 20 to 100 makes it much more likely you will find correlations that are high but spurious. </figcaption>
+  <figcaption>Fig2. As you increase the number of features, you get more high scoring spurious correlations. </figcaption>
 </figure>
 
-Another way of saying two objects have a statistical association is to say they are not statistically independent.
-However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing.
-A good example is genetics.
-President Barak Obama and his father Obama Sr.'s genetic profile both have a strong association with that of his daughter Malia Obama, but if you already know Barak Obama's profile, knowing Obama Sr.'s profile will give you no additional information about Malia's profile -- Malia's genetic profile is conditionally independent of Obama Sr.'s profile, given you know the Barak Obama's profile.
-Figure 2 continues with the MAPK example;
-
-<figure>
-  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
-  <figcaption>Fig2. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
-  Top; the lines represent correlation - the activity of the kinases on this pathway are all correlation.
-Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk.
-They are conditionally independent.</figcaption>
-</figure>
-
-Though approaches vary, state-of-the-art algorithms for determining causal relationships between several simultaneously measured features all attempt to reduce the set of associations by searching for cases of conditional independence, resulting in a far sparser set of conditional dependence relationships.
-However, large-scale experiments make this search task more difficult.
-To illustrate, we repeat the previous simulation, and instead of finding the max Pearson correlation in each instance, we apply a search algorithm that iteratively performs conditional independence tests, returning a sparse set of conditional dependence relationships.
+The increased incidence of spurious correlation means increased false positives in detecting conditional dependence relationships.
+To illustrate, we repeat the previous simulation, except now expanding from 20 to only 100 features.
+In each of the 500 instances, we apply a search algorithm that iteratively performs conditional independence tests, returning a count of detected conditional dependence relationships.
+As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is false.
 
@@
 library(bnlearn)
-set_20 <- rep(0, 10)
-set_500 <- rep(0, 10)
-for(i in 1:length(set_10)){
+set_20 <- rep(0, 500)
+set_100 <- rep(0, 500)
+for(i in 1:length(set_20)){
   print(i)
-  x <- as.data.frame(matrix(rnorm(500 * 20), ncol = 20))
-  set_10[i] <- narcs(gs(x, undirected = TRUE))
-  y <- as.data.frame(matrix(rnorm(500 * 500), ncol = 500))
+  x <- as.data.frame(matrix(rnorm(100 * 20), ncol = 20))
+  set_20[i] <- narcs(gs(x, undirected = TRUE))
+  y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
   set_100[i] <- narcs(gs(y, undirected = TRUE))
 }
-hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
-     main = "Counts of Conditional Dependence Relationships")
+hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
+     main = "Spurious Conditional Dependence Relationships")
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("top", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+legend("top", legend=c("20 features", "100 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
 ```
 
-Simulation running, image placeholder.
+<figure>
+  <img src="http://i.imgur.com/GzfGjio.png" alt="spurious conditional dependence" width="304" height="228">
+  <figcaption>Fig3. Increasing the number of measured features means increasing the false positive detection of conditional dependence.</figcaption>
+</figure>
 
-## Use of prior knowledge and interventions
+Lastly, measuring more things also adversely effects the selection of interventions in the design.
+The more components of the cell you measure, the more interventions you need to target those components.
+This, of course, increases the cost of the experiment.
+Under constrained resources, we can use a limited set of interventions that will partially elucidate causality in the system.
+We use prior biological knowledge to prioritize what to target with that limited set of interventions.
+For example, if we know what proteins a candidate drug is likely to impact, then interventions should target those proteins and those known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
+
+## Use of prior biological knowledge to address the spurious association problem
+
+Similarly, we can use to prior biological knowledge to address the spurious association problem and improve detection of conditional dependence relationships.
+Specifically, this approach uses pathway databases such as KEGG to answer the questions;
+
+1) What causal relations are we sure exist? 
+2) What causal relations are we sure do not exist? 
+3) What causal relations do we expect to exist/not to exist?
+
+Answering these questions enables us to improve the search for conditional dependence relations, minimizing false positives do to spurious associations.

### 2015-09-06 write_up.Rmd
incorporated Karens changes

@@
 
-Our goal in molecular cell biology is causal inference -- meaning uncovering the regulatory relationships between components of the cell that determine phenotype and disease states.
-Generally observing statistical association between such components does not demonstrate causality.
+Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies.
+Generally, observing statistical association between network components does not demonstrate causality.
 However, we can safely use statistical association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
 
-High-throughput omics technologies enable us to conduct large-scale experiments, where we simultaneously measure multiple cellular components in a sample, and in many cases acquire a large amount of data for each component.
-However, despite generating much more data, these large-scale experiments make the task of causal inference much more difficult.
-Large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-We can address this problem by improving our experimental design.
+High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
+The huge amount of data generated paradoxically make the task of causal inference much more difficult, because, when many associations are examined, some will appear significant just by chance.
+In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
+This problem can be addressed by improving experimental design and/or refining the biological question.
+Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
 
-## How we infer causality from omics experiments
+Statistical inference
+In MAPK signaling, Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
+Imagine this relationship was unknown: could it be detected from measurements of these phosphoproteins?
+Given sufficient data, a correlation will be detected between each pair of proteins, due to (direct and indirect) causal relationships.
+Figure 1, top panel, depicts these correlation as black lines, causal relationships are depicted as blue arrows.
+The edge linking Raf to Erk is not causal, because Raf’s effect on Erk occurs only via Mek (and that affect is already represented, rendering the Raf-Erk edge superfluous).
+Can we use a statistical trick to eliminate this noncausal edge?
 
-The MAPK signaling pathway proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.
+First, some terminology.
+A statistical association (such as a correlation) between two variables indicates that they are dependent.
+Sometimes we choose to assess a dependence in the context of the value of a third variable, in this case the dependence is called conditional dependence, ie because it is conditioned on the third variable.
+Let’s see how this applies here, by examining the dependence between Raf and Erk.
+If we condition on Mek, then we assume that we know the value of Mek: perhaps we select only samples which have a high level of Mek, so Mek is fixed at level “high”.
+Can the dependence between Raf and Erk still be detected?
+If, due to noise, some of these Mek=high samples have an intermediate or even low level of Raf, will Erk be correspondingly low?
+Based on the causal mechanism, Erk will more closely follow Mek, since Mek is its causal parent.
+Because of that, the Raf-Erk dependence, once conditioned on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.
+Using this approach, it is possible to keep only the causal edges as depicted by the black lines in Figure 1, bottom panel.
 
-Assuming this causal relation were not known, the question of whether Mek has a causal effect on Erk is evaluated with a designed experiment, where Mek is the treatment and Erk is the response.
-Intervention and randomization are key parts of the experimental design.
-Intervention means we target the potential cause, Mek, and fix it at different levels, for example 'high activity' and 'baseline activity'.
-Randomization refers to random assignment of samples or subjects to each of these levels, helping to alleviate bias introduced by unseen factors.
-The experiment generates measurements that will show a strong statistical association between Mek and Erk.
-Intervention and randomization enable us to interpret that statistical association as evidence for Mek's causal influence on Erk.
-
-With omics technologies, instead of measuring one response, we simultaneous measure many.
-Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each of the measured features.
-
-As with the simple one-response experimental design, we seek to use these associations to evaluate the presence of causal relations.
-Saying two objects have a statistical association is the same as saying they are not statistically independent.
-However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing.
-Figure 1 illustrates this principle with the MAPK proteins.
+So far in this toy example, the causal edges have been found, but their causal direction has not been revealed.
+The most powerful technique for finding edge directionality is, quite naturally, perturbations, such as small molecule inhibitors.
+An intervention on Mek for instance reveals the causal edge direction from Mek to Erk.
+The Raf-->Mek edge is automatically revealed for two reasons: First, because Raf is not affected (or not affected strongly) by Mek inhibition, and secondly, because once one edge in this triplet is oriented, the second is automatically oriented by statistical properties of the structure, given certain assumptions.
+These properties are beyond the scope of this discussion (see … my thesis or a thousand other places) for a description.
+Suffice to say that while perturbations are very useful for orienting edges, it is happily not necessary to perform exhaustive perturbations in order to assess causal directions.
 
@@
 
+Fig1.
+In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
+Top; the lines represent associate - the activity of the kinases on this pathway are all correlated.
+Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk.
+They are conditionally independent.
+
 We use algorithms borrowed from the causal Bayesian network modeling domain to do causal inference in the context of an experiment with multiple intercorrelated response variables.
-These  algorithms' approaches vary but they essentially do two things; (1) find cases of conditional independence and reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use the experiment's intervention information to evaluate those conditional dependencies as evidence for potential causal relationtions.
+These algorithms’ approaches vary but they essentially do two things; (1) find cases of conditional independence and reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use the experiment’s intervention information to evaluate those conditional dependencies as evidence for potential causal relations.
 
-The challenge with large datasets is that by measuring more things, we increase the number of spurious statistical associations, and thus the number of spurious cases of conditional dependence, and thus the number of incorrect conclusions of the presence of a causal relation.
+II. Inference with large numbers of variables
+
+The challenge with large datasets is the large number of variables, which increase the number of spurious statistical associations, and thus the number of spurious cases of conditional dependence, and the number of incorrect conclusions of the presence of a causal relation.
 To illustrate, we ran a simulation where first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.
@@
 
-Lastly, measuring more things also adversely effects the selection of interventions in the design.
-The more components of the cell you measure, the more interventions you need to target those components.
-This, of course, increases the cost of the experiment.
-Under constrained resources, we can use a limited set of interventions that will partially elucidate causality in the system.
-We use prior biological knowledge to prioritize what to target with that limited set of interventions.
-For example, if we know what proteins a candidate drug is likely to impact, then interventions should target those proteins and those known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
+III. Inferring causality from omics experiments
 
-## Use of prior biological knowledge to address the spurious association problem
+These simulations demonstrate that many variables (may) give many false results, rendering the problem of finding true connections relatively difficult.
+Fortunately, there are ways to address this problem, these Causality Inference in Omics tools are listed below:
+ 1) Measure many samples.
+If feasible, high throughput measurements that measure many samples (not just many variables) provide the statistical power to tell spurious associations from true associations.
+Think of a correlation plot showing just 4 dots.
+Even if they line up fairly well, they will not convincingly show a correlation.
+In contrast, 400 samples showing a correlation provide ample evidence.
+This intuition underlies the use of single cell data for causal modeling, as we described previously (self citation !
+:-P
 
-Similarly, we can use to prior biological knowledge to address the spurious association problem and improve detection of conditional dependence relationships.
-Specifically, this approach uses pathway databases such as KEGG to answer the questions;
+2) Refine the problem, thus limiting the variable number.
+If the biological system in question is sufficiently well studied, it might be possible to ask specific questions of the data, such as ascertaining the presence or absence of a particular edge or pathway, or assessing connections to a particular protein.
+The more specific the question, the less data is needed overall to make solid statistical inferences.
 
-1) What causal relations are we sure exist? 
-2) What causal relations are we sure do not exist? 
-3) What causal relations do we expect to exist/not to exist?
+3) Employ perturbations, especially ones relevant to the biological question under study.
+Perturbations are invaluable in finding causal directions but also help to determine which variables may be important to include in statistical analyses.
 
-Answering these questions enables us to improve the search for conditional dependence relations, minimizing false positives do to spurious associations.
+4) Use prior knowledge, not just from experts but also from noisier sources.  This approach uses pathway databases such as KEGG to answer the questions;
+
+i) What causal relations are we sure exist? 
+ii) What causal relations are we sure do not exist? 
+iii) What causal relations do we expect to exist/not to exist?
+
+In the context of this ‘weak’ evidence, it becomes much easier to constrain the problem of causal inference and learn more reliable relationships.
+
+IV. Winning workflows
+
+The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, rendering it much smaller, and thus allowing for accurate statistical inferences.
+For instance, the task of assessing which of all the possible KEGG pathways is present in my dataset is a far smaller (thus easier) task statistically speaking than the task of assessing which random paths among any measured variables exist in my dataset.
+
+How should the tools listed be used?
+They are most powerful when used in combination, and in fact the lines between them are somewhat arbitrary and frequently blurred.
+For instance, tool #3 calls for use of perturbations, but this task itself is complicated by large scale data, as the number of possible perturbations grows and experiments quickly become infeasible.
+Tool 4, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.
+For example, if proteins a candidate drug is likely to impact are known, then interventions could target those proteins and those known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
+
+Effective workflows will resemble Mr Potato Head with these tools as components, and may be used iteratively.  [OK I just wanted to entertain you guys with that image.
+Here I think we can wrap up, or we can list a few more tantalizing workflow suggestions with accompanying examples. Ie profer some workflow templates.  I’m not sure if that’s valuable or not]
+
+signing off..
+Karen :)

### 2015-09-07 write_up.Rmd
incorporated Olgas comments into Karens

--- 2015-09-04 write_up.Rmd
@@
 ---
-title: "Distilling Causality from Correlation in Large-scale Omics Datasets"
-author: "Robert Ness"
+title: "Large-scale Omics Experiments are not Conducive to Causal Inference"
 date: "September 3, 2015"
@@
 
-Our goal in molecular cell biology is causal inference -- meaning uncovering the regulatory relationships between components of the cell that determine phenotype and disease states.
-Generally observing statistical association between such components does not demonstrate causality.
-However, we can safely use statistical association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
+Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies.
+Generally, observing statistical association between network components does not demonstrate causality.
+However, we can safely use association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
 
-High-throughput omics technologies enable us to conduct large-scale experiments, where we simultaneously measure multiple cellular components in a sample, and in many cases acquire a large amount of data for each component.
-However, despite generating much more data, these large-scale experiments make the task of causal inference much more difficult.
-Large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-We can address this problem by improving our experimental design.
+High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
+The huge amount of data generated paradoxically make the task of causal inference much more difficult because, when many examining pairwise associations between measured features, some will appear significant just by chance.
+In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
+This problem can be addressed by improving experimental design and refining the biological question.
+Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
 
-## How we infer causality from omics experiments
+## Increasing the number of measured features decreases the ability to infer causality
 
-The MAPK signaling pathway proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.
+In MAPK signaling, Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
+Imagine these relationships were unknown: could it be detected from measurements of these phosphoproteins?
+If we took an arbitrary biological sample and simultaneously quantified Raf, Mek, and Erk, we'd find Raf-Mek, Mek-Erk, and Raf-Erk correlations, due their direct and indirect causal relationships.
 
-Assuming this causal relation were not known, the question of whether Mek has a causal effect on Erk is evaluated with a designed experiment, where Mek is the treatment and Erk is the response.
-Intervention and randomization are key parts of the experimental design.
-Intervention means we target the potential cause, Mek, and fix it at different levels, for example 'high activity' and 'baseline activity'.
-Randomization refers to random assignment of samples or subjects to each of these levels, helping to alleviate bias introduced by unseen factors.
-The experiment generates measurements that will show a strong statistical association between Mek and Erk.
-Intervention and randomization enable us to interpret that statistical association as evidence for Mek's causal influence on Erk.
+To get from correlation to causality, we need a controlled experiment.
+Based on the Mek-Erk correlation, we can design an experiment based on the hypothesis that Mek has a causal influence on Erk.
+In the experiment, we'd *intervene* in Mek, meaning we'd artificially control Mek at different levels of phospho-activity (for example with a small molecule treatment that inhibits Mek), then randomly assign cell lines to each of these levels.
+In doing so we'd control for confounding factors and free ourselves to interpret the correlation between Mek levels and Erk measurements as evidence for the presence of causal influence of Mek on Erk.
+Indeed, if our intervention targeted Erk and we observed Mek, the measured correlation would be very weak since the Erk->Mek causal relationship doesn't exist.
 
-With omics technologies, instead of measuring one response, we simultaneous measure many.
-Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each of the measured features.
+Now, suppose you have access to a technology that can simultaneously quantify several thousands of phosphoproteins, including Raf, Mek and Erk.
+As is typical with such large-scale experiments, you only have a very small number of interventions, and maybe 3 replicates.
+Even if those interventions targeted Raf, Mek, or Erk, there would be so much spurious correlation between these phosphoproteins and all the others you mentioned that causal inference would be infeasible.
 
-As with the simple one-response experimental design, we seek to use these associations to evaluate the presence of causal relations.
-Saying two objects have a statistical association is the same as saying they are not statistically independent.
-However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing.
-Figure 1 illustrates this principle with the MAPK proteins.
+To explain why, we need to introduce some terminology.
+A statistical association (such as Pearson correlation) between two measured entities indicates that they are dependent.
+Sometimes we choose to assess the dependence in the context of already knowing what other key components of the system are doing.
+In this case the dependence is called *conditional dependence*, because we evaluate the association conditional on the states of other components in the system.
+
+We can see how this applies here by examining the dependence between Raf and Erk.
+Figure 1, top panel, depicts the correlation between Raf, Mek, and Erk as black lines, causal relationships are depicted as blue arrows.
+If we condition on Mek, then we assume that we know what Mek is doing.
+In all the measurements where Mek is high, can the dependence between Raf and Erk still be detected?
+Since Mek is directly upstream of Erk, knowing Mek tells us all there is to know about Erk, and knowing Raf tells us nothing.
+Because of that, the Raf-Erk dependence, once we condition on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.
+Using this approach, it is possible to keep only the conditional dependence edges as depicted by the black lines in Figure 1, bottom panel.
+Conditional dependence is important because, under certain assumptions, we can move from conditional dependence to causal influence by applying interventions.
 
@@
   <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
-  <figcaption>Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
-  Top; the lines represent associate - the activity of the kinases on this pathway are all correlated.
+  <figcaption>
+  Fig1.
+In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
+Top; the lines represent association - the activity of the kinases on this pathway are all correlated.
 Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk.
-They are conditionally independent.</figcaption>
+They are conditionally independent.
+  </figcaption>
 </figure>
 
-We use algorithms borrowed from the causal Bayesian network modeling domain to do causal inference in the context of an experiment with multiple intercorrelated response variables.
-These  algorithms' approaches vary but they essentially do two things; (1) find cases of conditional independence and reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use the experiment's intervention information to evaluate those conditional dependencies as evidence for potential causal relationtions.
+Computational methods that do causal inference in the context of an experiment that simultaneously measures multiple intercorrelated response variables all essentially have the same two step process; (1) reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
 
-The challenge with large datasets is that by measuring more things, we increase the number of spurious statistical associations, and thus the number of spurious cases of conditional dependence, and thus the number of incorrect conclusions of the presence of a causal relation.
-To illustrate, we ran a simulation where first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.
-In each case the features are completely independent.
+The problem of doing causal inference with large-scale experiments is two-fold.
+Firstly, sample size is typically far too small for the large number of features we measure.
+With the large-scale experiment described above, it is clear that viewing a scatter-plot between a pair of features that has only 3 points is uninformative -- even if the points correlated perfectly, 3 points is not enough to conclude anything about the relationship with any confidence.
+But even with hundreds of samples, we are still plagued by spurious associations when the number of features is so large, since more features means greater chance associations will occur at random.
+
+To illustrate, we ran a simulation where we first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.
+In both cases the features are completely independent.
 This means if we quantify association using Pearson correlation, any correlation we find will be completely spurious.
@@
 In each of the 500 instances, we apply a search algorithm that iteratively performs conditional independence tests, returning a count of detected conditional dependence relationships.
-As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is false.
+As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is a false positive, i.e.
+it doesn't actually exist in the mechanism that generated the data.
 
@@
 
-Lastly, measuring more things also adversely effects the selection of interventions in the design.
-The more components of the cell you measure, the more interventions you need to target those components.
-This, of course, increases the cost of the experiment.
-Under constrained resources, we can use a limited set of interventions that will partially elucidate causality in the system.
-We use prior biological knowledge to prioritize what to target with that limited set of interventions.
-For example, if we know what proteins a candidate drug is likely to impact, then interventions should target those proteins and those known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
+As you increase the number of features you measure, the number of false positive detection of conditional dependence explodes.
+What this means is that the computational methods for causal inference will fail for the typical large-scale experiment, because when their input data contains a large amount of features relative to sample size, they cannot reliably detect the conditional dependence relationships they need to infer causality.
 
-## Use of prior biological knowledge to address the spurious association problem
+The second problem is that the aforementioned computational methods require interventions to to infer causality.
+To infer the causal relationships between many measured features, you need many interventions targeting those many features.
+Thousands of interventions targeting thousands of features is infeasible.
 
-Similarly, we can use to prior biological knowledge to address the spurious association problem and improve detection of conditional dependence relationships.
-Specifically, this approach uses pathway databases such as KEGG to answer the questions;
+## Experimentalists can apply heuristics for causal inference in large-scale omics data
 
-1) What causal relations are we sure exist? 
-2) What causal relations are we sure do not exist? 
-3) What causal relations do we expect to exist/not to exist?
+Fortunately, there are ways to address this problem, these *Causality Inference in Omics* heuristics are listed below:
 
-Answering these questions enables us to improve the search for conditional dependence relations, minimizing false positives do to spurious associations.
+1) Measure many samples.
+If feasible, high throughput measurements that measure many samples (not just many features) provide the statistical power to tell true associations from spurious associations.
+This intuition motivates the use of single cell data for causal modeling, where having many thousands of cells per sample provides statistical power.
+
+2) Refine the problem, thus limiting the number of features.
+If the biological system in question is sufficiently well-studied, it might be possible to ask specific questions of the data, such as ascertaining the presence or absence of a particular edge or pathway, or assessing connections to a particular protein.
+The more specific the question, the less data is needed overall to make solid statistical inferences.
+
+3) Use prior knowledge, not just from experts but also from noisier sources.  Based on information in  pathway databases such as KEGG, answer the questions;
+
+i) What causal relations are we sure exist? 
+ii) What causal relations are we sure do not exist? 
+iii) What causal relations do we expect to exist/not to exist?
+
+Such questions can be answered with a variety of biological knowledge.
+For example, we know more about causal relations between phosphoproteins known to cluster together spatially within the cell, and be more interested in learning about signaling mechanisms that cross compartments inside the cell.
+By answering these questions, we can bias the application of computational causal inference questions such that they focus less on conditional dependence and causal relations we know to exist and more towards the novel causal information hidden in the data.
+
+4) Employ interventions selectively.
+Some interventions are 'fat-handed', in that they perturb many features in the system, providing more causal inference bang for your intervention buck.
+At the same time, targeted interventions are more useful for gaining mechanistic insight into the biological question under study.
+For example, if proteins a candidate drug is likely to impact are known, then interventions could target those proteins and other proteins known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
+
+The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, rendering it much smaller, and thus allowing for accurate statistical inferences.
+For instance, the task of assessing which of all the possible KEGG pathways is present in my dataset is a far smaller (thus easier) task statistically speaking than the task of assessing which of all possible combinations of my measured features might form a biological pathway.
+
+How should the tools listed be used?
+They are most powerful when used in combination, and in fact the lines between them are somewhat arbitrary and frequently blurred.
+For instance, tool #4 calls for use of interventions, but this task itself is complicated by large scale data, as the number of possible perturbations grows and experiments quickly become infeasible.
+Tool 3, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.

### 2015-09-07 write_up.Rmd
olgas intro

@@
 Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies.
-Generally, observing statistical association between network components does not demonstrate causality.
-However, we can safely use association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
+Large-scale experiments are fishing expeditions, where we look for associations.
+However, the associations provided by this type of experiment are not conducive to finding causal relationships.
 
-High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
-The huge amount of data generated paradoxically make the task of causal inference much more difficult because, when many examining pairwise associations between measured features, some will appear significant just by chance.
-In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-This problem can be addressed by improving experimental design and refining the biological question.
-Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
+The goal of this manuscript is to highlight how large-scale experiments can impede the quest to uncovering the causal mechanism that determines the biology, and provide suggestions for addressing this problem through better experimental design and refining the biological question.
+We illustrate this on the MAPK pathway.
 

### 2015-09-14 write_up.Rmd
iteration of draft after week of google doc editing

--- 2015-09-07 write_up.Rmd
@@
 ---
-title: "Large-scale Omics Experiments are not Conducive to Causal Inference"
-date: "September 3, 2015"
+title: "From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments"
 output: html_document
@@
 Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies.
-Generally, observing statistical association between network components does not demonstrate causality.
-However, we can safely use association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.
+Statistical associations can reveal an alluring set of putative causal interactions, but when are such associations significant and reliable?
+And when can they be interpreted as causal?
+Our goal below is to provide suggestions for causal inference in large scale experiments, such as those resulting from high throughput omics technologies.
+We describe the pitfalls of large datasets, and suggest methods to reliably find causal associations despite these pitfalls.
 
 High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
-The huge amount of data generated paradoxically make the task of causal inference much more difficult because, when many examining pairwise associations between measured features, some will appear significant just by chance.
+The huge amount of data generated paradoxically make the task of causal inference much more difficult, because, when many associations are examined, some will appear significant just by chance.
 In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-This problem can be addressed by improving experimental design and refining the biological question.
+This problem can be addressed by improving experimental design and/or refining the biological question.
 Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
 
-## Increasing the number of measured features decreases the ability to infer causality
+## I. Statistical inference of causal relationships
 
 In MAPK signaling, Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
-Imagine these relationships were unknown: could it be detected from measurements of these phosphoproteins?
-If we took an arbitrary biological sample and simultaneously quantified Raf, Mek, and Erk, we'd find Raf-Mek, Mek-Erk, and Raf-Erk correlations, due their direct and indirect causal relationships.
+Imagine this relationship was unknown: could it be detected from measurements of these phosphoproteins?
+Given sufficient data, a correlation will be detected between each pair of proteins, due to (direct and indirect) causal relationships.
+Figure 1, top panel, depicts these correlation as black lines, causal relationships are depicted as blue arrows.
 
-To get from correlation to causality, we need a controlled experiment.
-Based on the Mek-Erk correlation, we can design an experiment based on the hypothesis that Mek has a causal influence on Erk.
-In the experiment, we'd *intervene* in Mek, meaning we'd artificially control Mek at different levels of phospho-activity (for example with a small molecule treatment that inhibits Mek), then randomly assign cell lines to each of these levels.
-In doing so we'd control for confounding factors and free ourselves to interpret the correlation between Mek levels and Erk measurements as evidence for the presence of causal influence of Mek on Erk.
-Indeed, if our intervention targeted Erk and we observed Mek, the measured correlation would be very weak since the Erk->Mek causal relationship doesn't exist.
+To go from correlation to causality, we need to perform an intervention: for instance, we can use a small molecule inhibitor to block Mek activity.
+Resulting measurements show an impact on Erk but not on Raf, correctly orienting the two edges.
+Note that it was not necessary to intervene on each node in order to get the correct causal directions, from Raf to Mek and Mek to Erk.
+What about the noncausal edge?
+The edge linking Raf to Erk is not causal, because Raf’s effect on Erk occurs only via Mek (and that affect is already represented, rendering the Raf-Erk edge superfluous).
+In statistical inference, this edge can be eliminated without the need of further interventions.
+How is this done?
 
-Now, suppose you have access to a technology that can simultaneously quantify several thousands of phosphoproteins, including Raf, Mek and Erk.
-As is typical with such large-scale experiments, you only have a very small number of interventions, and maybe 3 replicates.
-Even if those interventions targeted Raf, Mek, or Erk, there would be so much spurious correlation between these phosphoproteins and all the others you mentioned that causal inference would be infeasible.
+First, some terminology.
+A statistical association (such as a correlation) between two variables indicates that they are dependent.
+Sometimes we choose to assess a dependence in the context of the value of other system components, such as a third variable.
+In this case the dependence is called conditional dependence, because it is evaluated conditional on the state of other system components (i.e., the third variable).
+Let’s see how this applies here, by examining the dependence between Raf and Erk.
+If we condition on Mek, then we assume that we know the value of Mek: perhaps we select only samples which have a high level of Mek, so Mek is fixed at level “high”.
+Can the dependence between Raf and Erk still be detected?
+If, due to noise, some of these Mek=high samples have an intermediate or even low level of Raf, will Erk be correspondingly low?
+Based on the causal mechanism, Erk will more closely follow Mek, since Mek is its causal parent.
+Because of that, the Raf-Erk dependence, once conditioned on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.
+Using this approach, it is possible to keep only the causal edges as depicted by the black lines in Figure 1, bottom panel.
 
-To explain why, we need to introduce some terminology.
-A statistical association (such as Pearson correlation) between two measured entities indicates that they are dependent.
-Sometimes we choose to assess the dependence in the context of already knowing what other key components of the system are doing.
-In this case the dependence is called *conditional dependence*, because we evaluate the association conditional on the states of other components in the system.
+Computational methods that do causal inference in the context of an experiment that simultaneously measures multiple intercorrelated response variables all essentially have the same two step process; (1) reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
 
-We can see how this applies here by examining the dependence between Raf and Erk.
-Figure 1, top panel, depicts the correlation between Raf, Mek, and Erk as black lines, causal relationships are depicted as blue arrows.
-If we condition on Mek, then we assume that we know what Mek is doing.
-In all the measurements where Mek is high, can the dependence between Raf and Erk still be detected?
-Since Mek is directly upstream of Erk, knowing Mek tells us all there is to know about Erk, and knowing Raf tells us nothing.
-Because of that, the Raf-Erk dependence, once we condition on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.
-Using this approach, it is possible to keep only the conditional dependence edges as depicted by the black lines in Figure 1, bottom panel.
-Conditional dependence is important because, under certain assumptions, we can move from conditional dependence to causal influence by applying interventions.
+## II. Inference with large numbers of variables
+Suppose we perform a large scale experiment, quantifying several thousands of phosphoproteins, including Raf, Mek and Erk.
+A typical large-scale experiments includes a very small number of interventions, and around 3 replicates.
+Statistical inference fails in this case, because the large number of variables measured result in a large number of spurious statistical associations, which appear as strong as true associations.
+The relationship between Raf, Mek and Erk, for instance, will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.
 
@@
 </figure>
-
-Computational methods that do causal inference in the context of an experiment that simultaneously measures multiple intercorrelated response variables all essentially have the same two step process; (1) reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
 
@@
 <figure>
-  <img src="http://i.imgur.com/CSuQdePl.png" alt="spurious1">
+  <img src="http://i.imgur.com/CSuQdePl.png" alt="spurious1", width="304" height="228">
   <figcaption>Fig2. As you increase the number of features, you get more high scoring spurious correlations. </figcaption>
@@
 
-The second problem is that the aforementioned computational methods require interventions to to infer causality.
-To infer the causal relationships between many measured features, you need many interventions targeting those many features.
-Thousands of interventions targeting thousands of features is infeasible.
+The second problem is the requirement for interventions.
+As the number of features grows, the number of interventions needed to infer causality grows, and performing sufficient perturbation experiments becomes infeasible.
 
-## Experimentalists can apply heuristics for causal inference in large-scale omics data
+## III. Inferring causality from omics experiments
 
-Fortunately, there are ways to address this problem, these *Causality Inference in Omics* heuristics are listed below:
+The problems outlined in section II paint a grim picture for causal learning in large datasets.
+Fortunately, these can be overcome, and effective causal learning can be a reality for large scale datasets.
+We proffer the approaches listed in the Causality Inference in Omics tools below.
 
-1) Measure many samples.
+1.
+*Refine the biological problem*, thus limiting the number of features.
+If the broader biological system is well-studied, it may be possible design an experiment that focuses on a specific part of the system of interest, then ask more specific questions of the data, such as whether a particular edge or pathway is present.
+The more specific the question, the less data is needed overall to make solid statistical inferences.
+2.
+*Measure more  samples*.
 If feasible, high throughput measurements that measure many samples (not just many features) provide the statistical power to tell true associations from spurious associations.
-This intuition motivates the use of single cell data for causal modeling, where having many thousands of cells per sample provides statistical power.
+This intuition motivates the use technologies that have lower throughput but enable measurement of more samples (or gather more data points per sample), for causal modeling.
+An example is  single cell  mass cytometry, where  many thousands of cells per sample provide ample statistical power.
+3.
+*Use prior knowledge*, not just from experts but also from noisier sources, to improve the process of searching for conditional independence.
+The MAPK pathway for instance, mentioned earlier, is well established and canonical.
+Assuming it, and other well established connections a priori can help reduce the number of possible associations, and enable more effective use of data.
+Knowledge about canonical causal relationships can be sourced from pathway databases such as KEGG. 
+ Another example is contextual information, such as spatial relations between phosphoproteins in the cell -- causal inference algorithms can be formulated to weigh evidence of conditional dependence differently depending on whether phosphoproteins are from the same or different spatial compartments.
+Employing prior knowledge that provides evidence for the regulatory network reduces the number of connections and causal arcs under consideration, which allows the available data to make reliable statistical inferences.
+4.
+*Employ targeted interventions selectively*.
+Targeted interventions perturb individual components of the system.
+An example is small molecule inhibitors which block the causal influence of a specific phoshoprotein on downstream components.
+Since it is usually prohibitive to perturb every target component of a system, it is strategic to prioritize application of targeted interventions to parts of the system that have the most potential for new discoveries.
+Selection of these perturbation targets can be based on prior knowledge - e.g.
+knowledge of which components are crucial players in the system of interest - or they can be applied iteratively, after an initial statistical analysis has revealed areas of the network in which causal inferences are not possible based on existing data.
+For instance, a resulting model graph with undirected edges can be inspected to reveal which nodes have potential to reveal the most causality if perturbed.
+5.
+*Consider broad-scale interventions*.
+Traditionally, experiments consider a biological response to a particular stimulus, possibly in the context of an inhibitor.
+In contrast, broad-scale interventions sacrifice specificity to perturb many features in the system simultaneously.
+The advantage of this approach is that it can enable elucidation of  causality across the entire system.
+Just as one intervention was sufficient to infer causality between 3 proteins Raf, Mek, and Erk, perturbing many things at once creates cascades of causal direction orientation across the broader network.
+This includes varying experimental conditions to activate multiple pathways.
+For example, signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling.
+Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.
+In contrast to targeted interventions, while it is difficult to know in advance what information will provided by broad-scale interventions, they provide more causal inference bang for your intervention buck.
 
-2) Refine the problem, thus limiting the number of features.
-If the biological system in question is sufficiently well-studied, it might be possible to ask specific questions of the data, such as ascertaining the presence or absence of a particular edge or pathway, or assessing connections to a particular protein.
-The more specific the question, the less data is needed overall to make solid statistical inferences.
-
-3) Use prior knowledge, not just from experts but also from noisier sources.  Based on information in  pathway databases such as KEGG, answer the questions;
-
-i) What causal relations are we sure exist? 
-ii) What causal relations are we sure do not exist? 
-iii) What causal relations do we expect to exist/not to exist?
-
-Such questions can be answered with a variety of biological knowledge.
-For example, we know more about causal relations between phosphoproteins known to cluster together spatially within the cell, and be more interested in learning about signaling mechanisms that cross compartments inside the cell.
-By answering these questions, we can bias the application of computational causal inference questions such that they focus less on conditional dependence and causal relations we know to exist and more towards the novel causal information hidden in the data.
-
-4) Employ interventions selectively.
-Some interventions are 'fat-handed', in that they perturb many features in the system, providing more causal inference bang for your intervention buck.
-At the same time, targeted interventions are more useful for gaining mechanistic insight into the biological question under study.
-For example, if proteins a candidate drug is likely to impact are known, then interventions could target those proteins and other proteins known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.
-
-The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, rendering it much smaller, and thus allowing for accurate statistical inferences.
-For instance, the task of assessing which of all the possible KEGG pathways is present in my dataset is a far smaller (thus easier) task statistically speaking than the task of assessing which of all possible combinations of my measured features might form a biological pathway.
+The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, and thus allowing for accurate statistical inferences.
+For instance, the task of assessing which of all the possible KEGG pathways is present in a dataset will be far less error-prone  than the task of assessing which of all possible combinations of my measured features might form a biological pathway.
 
@@
 They are most powerful when used in combination, and in fact the lines between them are somewhat arbitrary and frequently blurred.
-For instance, tool #4 calls for use of interventions, but this task itself is complicated by large scale data, as the number of possible perturbations grows and experiments quickly become infeasible.
-Tool 3, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.
+For instance, using tool #1 and tool #2 in concert can be thought of as reducing the breadth and increasing the depth of the investigation.
+Tools #4 and #5 call for use of interventions, but this task itself is complicated by measuring many things, since we have more features to expose to intervention.
+.
+Tool #3, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.

### 2015-09-14 write_up.Rmd
pdf doc format

@@
 title: "From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments"
-output: html_document
+output: pdf_document
 ---

### 2015-09-14 write_up.tex
used pandoc to create a tex

--- 2015-09-14 write_up.Rmd
@@
----
-title: "From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments"
-output: pdf_document
----
+\documentclass[]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies.
-Statistical associations can reveal an alluring set of putative causal interactions, but when are such associations significant and reliable?
-And when can they be interpreted as causal?
-Our goal below is to provide suggestions for causal inference in large scale experiments, such as those resulting from high throughput omics technologies.
-We describe the pitfalls of large datasets, and suggest methods to reliably find causal associations despite these pitfalls.
+\date{}
 
-High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
-The huge amount of data generated paradoxically make the task of causal inference much more difficult, because, when many associations are examined, some will appear significant just by chance.
-In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-This problem can be addressed by improving experimental design and/or refining the biological question.
-Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
+\begin{document}
 
-## I. Statistical inference of causal relationships
+Causal inference -- the task of uncovering regulatory relationships
+between components of biomolecular pathways and networks - is a primary
+goal of omics datasets in molecular biology studies. Statistical
+associations can reveal an alluring set of putative causal interactions,
+but when are such associations significant and reliable? And when can
+they be interpreted as causal? Our goal below is to provide suggestions
+for causal inference in large scale experiments, such as those resulting
+from high throughput omics technologies. We describe the pitfalls of
+large datasets, and suggest methods to reliably find causal associations
+despite these pitfalls.
 
-In MAPK signaling, Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
-Imagine this relationship was unknown: could it be detected from measurements of these phosphoproteins?
-Given sufficient data, a correlation will be detected between each pair of proteins, due to (direct and indirect) causal relationships.
-Figure 1, top panel, depicts these correlation as black lines, causal relationships are depicted as blue arrows.
+High-throughput omics technologies enable large-scale experiments, with
+simultaneous measurements of multiple cellular components in a sample.
+The huge amount of data generated paradoxically make the task of causal
+inference much more difficult, because, when many associations are
+examined, some will appear significant just by chance. In other words,
+large-scale experiments are more prone to hiding the true signal and
+generating spurious associations, leading to increased false positive
+conclusions that causal relationships are present. This problem can be
+addressed by improving experimental design and/or refining the
+biological question. Below, we describe the process of elucidating
+statistical associations from data and suggest practical approaches for
+analysis of large scale datasets.
 
-To go from correlation to causality, we need to perform an intervention: for instance, we can use a small molecule inhibitor to block Mek activity.
-Resulting measurements show an impact on Erk but not on Raf, correctly orienting the two edges.
-Note that it was not necessary to intervene on each node in order to get the correct causal directions, from Raf to Mek and Mek to Erk.
-What about the noncausal edge?
-The edge linking Raf to Erk is not causal, because Raf’s effect on Erk occurs only via Mek (and that affect is already represented, rendering the Raf-Erk edge superfluous).
-In statistical inference, this edge can be eliminated without the need of further interventions.
-How is this done?
+I. Statistical inference of causal relationships
 
-First, some terminology.
-A statistical association (such as a correlation) between two variables indicates that they are dependent.
-Sometimes we choose to assess a dependence in the context of the value of other system components, such as a third variable.
-In this case the dependence is called conditional dependence, because it is evaluated conditional on the state of other system components (i.e., the third variable).
-Let’s see how this applies here, by examining the dependence between Raf and Erk.
-If we condition on Mek, then we assume that we know the value of Mek: perhaps we select only samples which have a high level of Mek, so Mek is fixed at level “high”.
-Can the dependence between Raf and Erk still be detected?
-If, due to noise, some of these Mek=high samples have an intermediate or even low level of Raf, will Erk be correspondingly low?
-Based on the causal mechanism, Erk will more closely follow Mek, since Mek is its causal parent.
-Because of that, the Raf-Erk dependence, once conditioned on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.
-Using this approach, it is possible to keep only the causal edges as depicted by the black lines in Figure 1, bottom panel.
+In MAPK signaling, Raf causally affects active Mek levels (via
+phosphorylation), while Mek causally affects Erk. Imagine this
+relationship was unknown: could it be detected from measurements of
+these phosphoproteins? Given sufficient data, a correlation will be
+detected between each pair of proteins, due to (direct and indirect)
+causal relationships. Figure 1, top panel, depicts these correlation as
+black lines, causal relationships are depicted as blue arrows.
 
-Computational methods that do causal inference in the context of an experiment that simultaneously measures multiple intercorrelated response variables all essentially have the same two step process; (1) reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
+To go from correlation to causality, we need to perform an intervention:
+for instance, we can use a small molecule inhibitor to block Mek
+activity. Resulting measurements show an impact on Erk but not on Raf,
+correctly orienting the two edges. Note that it was not necessary to
+intervene on each node in order to get the correct causal directions,
+from Raf to Mek and Mek to Erk. What about the noncausal edge? The edge
+linking Raf to Erk is not causal, because Raf's effect on Erk occurs
+only via Mek (and that affect is already represented, rendering the
+Raf-Erk edge superfluous). In statistical inference, this edge can be
+eliminated without the need of further interventions. How is this done?
 
-## II. Inference with large numbers of variables
-Suppose we perform a large scale experiment, quantifying several thousands of phosphoproteins, including Raf, Mek and Erk.
-A typical large-scale experiments includes a very small number of interventions, and around 3 replicates.
-Statistical inference fails in this case, because the large number of variables measured result in a large number of spurious statistical associations, which appear as strong as true associations.
-The relationship between Raf, Mek and Erk, for instance, will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.
+First, some terminology. A statistical association (such as a
+correlation) between two variables indicates that they are dependent.
+Sometimes we choose to assess a dependence in the context of the value
+of other system components, such as a third variable. In this case the
+dependence is called conditional dependence, because it is evaluated
+conditional on the state of other system components (i.e., the third
+variable). Let's see how this applies here, by examining the dependence
+between Raf and Erk. If we condition on Mek, then we assume that we know
+the value of Mek: perhaps we select only samples which have a high level
+of Mek, so Mek is fixed at level ``high''. Can the dependence between
+Raf and Erk still be detected? If, due to noise, some of these Mek=high
+samples have an intermediate or even low level of Raf, will Erk be
+correspondingly low? Based on the causal mechanism, Erk will more
+closely follow Mek, since Mek is its causal parent. Because of that, the
+Raf-Erk dependence, once conditioned on Mek, will be reduced or
+eliminated, thus indicating that the Raf-Erk edge can be removed. Using
+this approach, it is possible to keep only the causal edges as depicted
+by the black lines in Figure 1, bottom panel.
 
-<figure>
-  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
-  <figcaption>
-Fig1.
-In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
-Top; the lines represent association - the activity of the kinases on this pathway are all correlated.
-Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk.
-They are conditionally independent.
-  </figcaption>
-</figure>
+Computational methods that do causal inference in the context of an
+experiment that simultaneously measures multiple intercorrelated
+response variables all essentially have the same two step process; (1)
+reduce the dense set of pairwise associations to a sparse set of
+conditional dependencies, (2) use information about the interventions
+included in the experiment to evaluate those conditional dependencies as
+evidence for potential causal relations.
 
-The problem of doing causal inference with large-scale experiments is two-fold.
-Firstly, sample size is typically far too small for the large number of features we measure.
-With the large-scale experiment described above, it is clear that viewing a scatter-plot between a pair of features that has only 3 points is uninformative -- even if the points correlated perfectly, 3 points is not enough to conclude anything about the relationship with any confidence.
-But even with hundreds of samples, we are still plagued by spurious associations when the number of features is so large, since more features means greater chance associations will occur at random.
+\begin{enumerate}
+\def\labelenumi{\Roman{enumi}.}
+\setcounter{enumi}{1}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Inference with large numbers of variables
+\end{enumerate}
 
-To illustrate, we ran a simulation where we first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.
-In both cases the features are completely independent.
-This means if we quantify association using Pearson correlation, any correlation we find will be completely spurious.
-We record the maximum correlation between the 20 feature set and the 500 feature set.
-We repeat this simulation 500 times.
+Suppose we perform a large scale experiment, quantifying several
+thousands of phosphoproteins, including Raf, Mek and Erk. A typical
+large-scale experiments includes a very small number of interventions,
+and around 3 replicates. Statistical inference fails in this case,
+because the large number of variables measured result in a large number
+of spurious statistical associations, which appear as strong as true
+associations. The relationship between Raf, Mek and Erk, for instance,
+will be obscured by the many spurious relationships that they will each
+form with causally unrelated proteins.
 
-```{r, echo=FALSE, eval=FALSE}
-set_20 <- rep(0, 500)
-set_500 <- rep(0, 500)
-for(i in 1:500){
-  x <- matrix(rnorm(100 * 20), ncol = 20)
-  cor_x <- cor(x)
-  diag(cor_x) <- 0
-  set_20[i] <- max(cor_x)
-  y <- matrix(rnorm(100 * 500), ncol = 500)
-  cor_y <- cor(y)
-  diag(cor_y) <- 0
-  set_500[i] <- max(cor_y)
-}
-hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL, xlim = c(min(set_20) , max(set_500) + .1 ), 
-     main = "Highest Spurious Correlation")
-hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("topr", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
-```
+Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates
+Erk. Top; the lines represent association - the activity of the kinases
+on this pathway are all correlated. Bottom: the lines represent
+conditional dependence, there is no conditional dependence edge from Raf
+to Erk because given you know Mek, knowing Raf tells you nothing about
+Erk. They are conditionally independent.
 
-<figure>
-  <img src="http://i.imgur.com/CSuQdePl.png" alt="spurious1", width="304" height="228">
-  <figcaption>Fig2. As you increase the number of features, you get more high scoring spurious correlations. </figcaption>
-</figure>
+The problem of doing causal inference with large-scale experiments is
+two-fold. Firstly, sample size is typically far too small for the large
+number of features we measure. With the large-scale experiment described
+above, it is clear that viewing a scatter-plot between a pair of
+features that has only 3 points is uninformative -- even if the points
+correlated perfectly, 3 points is not enough to conclude anything about
+the relationship with any confidence. But even with hundreds of samples,
+we are still plagued by spurious associations when the number of
+features is so large, since more features means greater chance
+associations will occur at random.
 
-The increased incidence of spurious correlation means increased false positives in detecting conditional dependence relationships.
-To illustrate, we repeat the previous simulation, except now expanding from 20 to only 100 features.
-In each of the 500 instances, we apply a search algorithm that iteratively performs conditional independence tests, returning a count of detected conditional dependence relationships.
-As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is a false positive, i.e.
-it doesn't actually exist in the mechanism that generated the data.
+To illustrate, we ran a simulation where we first simulate a 20 features
+dataset each with a 100 Gaussian random measurements, then increase the
+number of features to 500. In both cases the features are completely
+independent. This means if we quantify association using Pearson
+correlation, any correlation we find will be completely spurious. We
+record the maximum correlation between the 20 feature set and the 500
+feature set. We repeat this simulation 500 times.
 
-```{r, echo=FALSE, eval = FALSE}
-library(bnlearn)
-set_20 <- rep(0, 500)
-set_100 <- rep(0, 500)
-for(i in 1:length(set_20)){
-  print(i)
-  x <- as.data.frame(matrix(rnorm(100 * 20), ncol = 20))
-  set_20[i] <- narcs(gs(x, undirected = TRUE))
-  y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
-  set_100[i] <- narcs(gs(y, undirected = TRUE))
-}
-hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
-     main = "Spurious Conditional Dependence Relationships")
-hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("top", legend=c("20 features", "100 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
-```
+Fig2. As you increase the number of features, you get more high scoring
+spurious correlations.
 
-<figure>
-  <img src="http://i.imgur.com/GzfGjio.png" alt="spurious conditional dependence" width="304" height="228">
-  <figcaption>Fig3. Increasing the number of measured features means increasing the false positive detection of conditional dependence.</figcaption>
-</figure>
+The increased incidence of spurious correlation means increased false
+positives in detecting conditional dependence relationships. To
+illustrate, we repeat the previous simulation, except now expanding from
+20 to only 100 features. In each of the 500 instances, we apply a search
+algorithm that iteratively performs conditional independence tests,
+returning a count of detected conditional dependence relationships. As
+before, since we simulated independent features, any conditional
+dependence relationship reported by this algorithm is a false positive,
+i.e.~it doesn't actually exist in the mechanism that generated the data.
 
-As you increase the number of features you measure, the number of false positive detection of conditional dependence explodes.
-What this means is that the computational methods for causal inference will fail for the typical large-scale experiment, because when their input data contains a large amount of features relative to sample size, they cannot reliably detect the conditional dependence relationships they need to infer causality.
+Fig3. Increasing the number of measured features means increasing the
+false positive detection of conditional dependence.
 
-The second problem is the requirement for interventions.
-As the number of features grows, the number of interventions needed to infer causality grows, and performing sufficient perturbation experiments becomes infeasible.
+As you increase the number of features you measure, the number of false
+positive detection of conditional dependence explodes. What this means
+is that the computational methods for causal inference will fail for the
+typical large-scale experiment, because when their input data contains a
+large amount of features relative to sample size, they cannot reliably
+detect the conditional dependence relationships they need to infer
+causality.
 
-## III. Inferring causality from omics experiments
+The second problem is the requirement for interventions. As the number
+of features grows, the number of interventions needed to infer causality
+grows, and performing sufficient perturbation experiments becomes
+infeasible.
 
-The problems outlined in section II paint a grim picture for causal learning in large datasets.
-Fortunately, these can be overcome, and effective causal learning can be a reality for large scale datasets.
-We proffer the approaches listed in the Causality Inference in Omics tools below.
+\begin{enumerate}
+\def\labelenumi{\Roman{enumi}.}
+\setcounter{enumi}{2}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Inferring causality from omics experiments
+\end{enumerate}
 
-1.
-*Refine the biological problem*, thus limiting the number of features.
-If the broader biological system is well-studied, it may be possible design an experiment that focuses on a specific part of the system of interest, then ask more specific questions of the data, such as whether a particular edge or pathway is present.
-The more specific the question, the less data is needed overall to make solid statistical inferences.
-2.
-*Measure more  samples*.
-If feasible, high throughput measurements that measure many samples (not just many features) provide the statistical power to tell true associations from spurious associations.
-This intuition motivates the use technologies that have lower throughput but enable measurement of more samples (or gather more data points per sample), for causal modeling.
-An example is  single cell  mass cytometry, where  many thousands of cells per sample provide ample statistical power.
-3.
-*Use prior knowledge*, not just from experts but also from noisier sources, to improve the process of searching for conditional independence.
-The MAPK pathway for instance, mentioned earlier, is well established and canonical.
-Assuming it, and other well established connections a priori can help reduce the number of possible associations, and enable more effective use of data.
-Knowledge about canonical causal relationships can be sourced from pathway databases such as KEGG. 
- Another example is contextual information, such as spatial relations between phosphoproteins in the cell -- causal inference algorithms can be formulated to weigh evidence of conditional dependence differently depending on whether phosphoproteins are from the same or different spatial compartments.
-Employing prior knowledge that provides evidence for the regulatory network reduces the number of connections and causal arcs under consideration, which allows the available data to make reliable statistical inferences.
-4.
-*Employ targeted interventions selectively*.
-Targeted interventions perturb individual components of the system.
-An example is small molecule inhibitors which block the causal influence of a specific phoshoprotein on downstream components.
-Since it is usually prohibitive to perturb every target component of a system, it is strategic to prioritize application of targeted interventions to parts of the system that have the most potential for new discoveries.
-Selection of these perturbation targets can be based on prior knowledge - e.g.
-knowledge of which components are crucial players in the system of interest - or they can be applied iteratively, after an initial statistical analysis has revealed areas of the network in which causal inferences are not possible based on existing data.
-For instance, a resulting model graph with undirected edges can be inspected to reveal which nodes have potential to reveal the most causality if perturbed.
-5.
-*Consider broad-scale interventions*.
-Traditionally, experiments consider a biological response to a particular stimulus, possibly in the context of an inhibitor.
-In contrast, broad-scale interventions sacrifice specificity to perturb many features in the system simultaneously.
-The advantage of this approach is that it can enable elucidation of  causality across the entire system.
-Just as one intervention was sufficient to infer causality between 3 proteins Raf, Mek, and Erk, perturbing many things at once creates cascades of causal direction orientation across the broader network.
-This includes varying experimental conditions to activate multiple pathways.
-For example, signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling.
-Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.
-In contrast to targeted interventions, while it is difficult to know in advance what information will provided by broad-scale interventions, they provide more causal inference bang for your intervention buck.
+The problems outlined in section II paint a grim picture for causal
+learning in large datasets. Fortunately, these can be overcome, and
+effective causal learning can be a reality for large scale datasets. We
+proffer the approaches listed in the Causality Inference in Omics tools
+below.
 
-The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, and thus allowing for accurate statistical inferences.
-For instance, the task of assessing which of all the possible KEGG pathways is present in a dataset will be far less error-prone  than the task of assessing which of all possible combinations of my measured features might form a biological pathway.
+Refine the biological problem, thus limiting the number of features. If
+the broader biological system is well-studied, it may be possible design
+an experiment that focuses on a specific part of the system of interest,
+then ask more specific questions of the data, such as whether a
+particular edge or pathway is present. The more specific the question,
+the less data is needed overall to make solid statistical inferences.
 
-How should the tools listed be used?
-They are most powerful when used in combination, and in fact the lines between them are somewhat arbitrary and frequently blurred.
-For instance, using tool #1 and tool #2 in concert can be thought of as reducing the breadth and increasing the depth of the investigation.
-Tools #4 and #5 call for use of interventions, but this task itself is complicated by measuring many things, since we have more features to expose to intervention.
-.
-Tool #3, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.
+Measure more samples. If feasible, high throughput measurements that
+measure many samples (not just many features) provide the statistical
+power to tell true associations from spurious associations. This
+intuition motivates the use technologies that have lower throughput but
+enable measurement of more samples (or gather more data points per
+sample), for causal modeling. An example is single cell mass cytometry,
+where many thousands of cells per sample provide ample statistical
+power.
+
+Use prior knowledge, not just from experts but also from noisier
+sources, to improve the process of searching for conditional
+independence. The MAPK pathway for instance, mentioned earlier, is well
+established and canonical. Assuming it, and other well established
+connections a priori can help reduce the number of possible
+associations, and enable more effective use of data. Knowledge about
+canonical causal relationships can be sourced from pathway databases
+such as KEGG. Another example is contextual information, such as spatial
+relations between phosphoproteins in the cell -- causal inference
+algorithms can be formulated to weigh evidence of conditional dependence
+differently depending on whether phosphoproteins are from the same or
+different spatial compartments. Employing prior knowledge that provides
+evidence for the regulatory network reduces the number of connections
+and causal arcs under consideration, which allows the available data to
+make reliable statistical inferences.
+
+Employ targeted interventions selectively. Targeted interventions
+perturb individual components of the system. An example is small
+molecule inhibitors which block the causal influence of a specific
+phoshoprotein on downstream components. Since it is usually prohibitive
+to perturb every target component of a system, it is strategic to
+prioritize application of targeted interventions to parts of the system
+that have the most potential for new discoveries. Selection of these
+perturbation targets can be based on prior knowledge - e.g.~knowledge of
+which components are crucial players in the system of interest - or they
+can be applied iteratively, after an initial statistical analysis has
+revealed areas of the network in which causal inferences are not
+possible based on existing data. For instance, a resulting model graph
+with undirected edges can be inspected to reveal which nodes have
+potential to reveal the most causality if perturbed.
+
+Consider broad-scale interventions. Traditionally, experiments consider
+a biological response to a particular stimulus, possibly in the context
+of an inhibitor. In contrast, broad-scale interventions sacrifice
+specificity to perturb many features in the system simultaneously. The
+advantage of this approach is that it can enable elucidation of
+causality across the entire system. Just as one intervention was
+sufficient to infer causality between 3 proteins Raf, Mek, and Erk,
+perturbing many things at once creates cascades of causal direction
+orientation across the broader network. This includes varying
+experimental conditions to activate multiple pathways. For example,
+signals from endocrine, paracrine, and autocrine ligands elicit various
+signaling responses in hepatocytes, thus interventions that cover this
+range of signals gives the best picture of the broader causal network of
+hepatocyte signaling. Similarly, interventions that go beyond
+receptor-level and perturb multiple components of the system bring
+cascading causal direct orientation deeper into the network. In contrast
+to targeted interventions, while it is difficult to know in advance what
+information will provided by broad-scale interventions, they provide
+more causal inference bang for your intervention buck.
+
+The Causality in Omics tool list above provides impactful approaches
+that can drastically improve causal inference from omics datasets, by
+constraining the inference task, and thus allowing for accurate
+statistical inferences. For instance, the task of assessing which of all
+the possible KEGG pathways is present in a dataset will be far less
+error-prone than the task of assessing which of all possible
+combinations of my measured features might form a biological pathway.
+
+How should the tools listed be used? They are most powerful when used in
+combination, and in fact the lines between them are somewhat arbitrary
+and frequently blurred. For instance, using tool \#1 and tool \#2 in
+concert can be thought of as reducing the breadth and increasing the
+depth of the investigation. Tools \#4 and \#5 call for use of
+interventions, but this task itself is complicated by measuring many
+things, since we have more features to expose to intervention. . Tool
+\#3, prior biological knowledge, can be used to prioritize what to
+target with that limited set of interventions.
+
+\end{document}

### 2015-09-14 write_up.tex
working from tex now

@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{0}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% This is a (brief) model paper using the achemso class
+%% The document class accepts keyval options, which should include
+%% the target journal and optionally the manuscript type.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[journal=jacsat,manuscript=article]{achemso}
 
-\date{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Place any additional packages needed here.  Only include packages
+%% which are essential, to avoid problems later. Do NOT use any
+%% packages which require e-TeX (for example etoolbox): the e-TeX
+%% extensions are not currently available on the ACS conversion
+%% servers.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
+\usepackage[T1]{fontenc}       % Use modern font encodings
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% If issues arise when submitting your manuscript, you may want to
+%% un-comment the next line.  This provides information on the
+%% version of every file you have used.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%\listfiles
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Place any additional macros here.  Please use \newcommand* where
+%% possible, and avoid layout-changing macros (which are not used
+%% when typesetting).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Meta-data block
+%% ---------------
+%% Each author should be given as a separate \author command.
+%%
+%% Corresponding authors should have an e-mail given after the author
+%% name as an \email command. Phone and fax numbers can be given
+%% using \phone and \fax, respectively; this information is optional.
+%%
+%% The affiliation of authors is given after the authors; each
+%% \affiliation command applies to all preceding authors not already
+%% assigned an affiliation.
+%%
+%% The affiliation takes an option argument for the short name.  This
+%% will typically be something like "University of Somewhere".
+%%
+%% The \altaffiliation macro should be used for new address, etc.
+%% On the other hand, \alsoaffiliation is used on a per author basis
+%% when authors are associated with multiple institutions.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{Robert O. Ness}
+\email{nessr@purdue.edu}
+\affiliation[Purdue University]{Department of Statistics, Purdue University, West Lafayette}
+\author{Karen Sachs}
+\affiliation[Stanford University]{School of Medicine, Stanford University, Palo Alto}
+\email{karens1@stanford.edu}
+\author{Olga Vitek}
+\email{o.vitek@neu.edu}
+\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Palo Alto}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The document title should be given as usual. Some journals require
+%% a running title from the author: this should be supplied as an
+%% optional argument to \title.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title[]
+  {From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Some journals require a list of abbreviations or keywords to be
+%% supplied. These should be set up here, and will be printed after
+%% the title and author information, if needed.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\abbreviations{IR,NMR,UV}
+\keywords{American Chemical Society, \LaTeX}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The manuscript does not need to include \maketitle, which is
+%% executed automatically.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Start the main part of the manuscript here.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
 Causal inference -- the task of uncovering regulatory relationships
@@
 analysis of large scale datasets.
-
-I. Statistical inference of causal relationships
+\section{Discussion}
+\subsection{I. Statistical inference of causal relationships}
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\Roman{enumi}.}
-\setcounter{enumi}{1}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Inference with large numbers of variables
-\end{enumerate}
+\subsection{II. Inference with large numbers of variables}
 
@@
 
-Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=0.5\textwidth]{figs/mapk.jpg}}
+\caption{In the MAPK signaling pathway, Raf regulates Mek, which regulates
 Erk. Top; the lines represent association - the activity of the kinases
@@
 to Erk because given you know Mek, knowing Raf tells you nothing about
-Erk. They are conditionally independent.
+Erk. They are conditionally independent.}
+\label{mapk}
+\end{figure}
 
@@
 
-Fig2. As you increase the number of features, you get more high scoring
-spurious correlations.
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
+\caption{As you increase the number of features, you get more high scoring
+spurious correlations.}
+\label{spur_corr}
+\end{figure}
 
@@
 
-Fig3. Increasing the number of measured features means increasing the
-false positive detection of conditional dependence.
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_dep.png}}
+\caption{Fig3. Increasing the number of measured features means increasing the
+false positive detection of conditional dependence.}
+\label{spur_dep}
+\end{figure}
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\Roman{enumi}.}
-\setcounter{enumi}{2}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Inferring causality from omics experiments
-\end{enumerate}
+\subsection{III. Inferring causality from omics experiments}
 
@@
 effective causal learning can be a reality for large scale datasets. We
-proffer the approaches listed in the Causality Inference in Omics tools
+proffer the approaches listed in \textit{The Causality Inference in Omics Tool List}
 below.
 
-Refine the biological problem, thus limiting the number of features. If
+\begin{enumerate}
+\item \textit{Refine the biological problem}, thus limiting the number of features. If
 the broader biological system is well-studied, it may be possible design
@@
 the less data is needed overall to make solid statistical inferences.
-
-Measure more samples. If feasible, high throughput measurements that
+\item \textit{Measure more samples}. If feasible, high throughput measurements that
 measure many samples (not just many features) provide the statistical
@@
 power.
-
-Use prior knowledge, not just from experts but also from noisier
+\item \textit{Use prior knowledge}, not just from experts but also from noisier
 sources, to improve the process of searching for conditional
@@
 make reliable statistical inferences.
-
-Employ targeted interventions selectively. Targeted interventions
+\item \textit{Employ targeted interventions selectively}. Targeted interventions
 perturb individual components of the system. An example is small
@@
 potential to reveal the most causality if perturbed.
-
-Consider broad-scale interventions. Traditionally, experiments consider
+\item \textit{Consider broad-scale interventions}. Traditionally, experiments consider
 a biological response to a particular stimulus, possibly in the context
@@
 more causal inference bang for your intervention buck.
+\end{enumerate}
 
-The Causality in Omics tool list above provides impactful approaches
+\textit{The Causality in Omics Tool List} above provides impactful approaches
 that can drastically improve causal inference from omics datasets, by
@@
 interventions, but this task itself is complicated by measuring many
-things, since we have more features to expose to intervention. . Tool
+things, since we have more features to expose to intervention. Tool
 \#3, prior biological knowledge, can be used to prioritize what to

### 2015-09-14 scripts_for_figures.R

@@ -1,0 +1,34 @@
+# R scripts for generating spurious correlation and conditional dependence histograms.
+## correlation figure
+set_20 <- rep(0, 500)
+set_500 <- rep(0, 500)
+for(i in 1:500){
+  x <- matrix(rnorm(100 * 20), ncol = 20)
+  cor_x <- cor(x)
+  diag(cor_x) <- 0
+  set_20[i] <- max(cor_x)
+  y <- matrix(rnorm(100 * 500), ncol = 500)
+  cor_y <- cor(y)
+  diag(cor_y) <- 0
+  set_500[i] <- max(cor_y)
+}
+hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL, xlim = c(min(set_20) , max(set_500) + .1 ), 
+     main = "Highest Spurious Correlation")
+hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
+legend("topr", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+
+## conditional dependence figure
+library(bnlearn)
+set_20 <- rep(0, 500)
+set_100 <- rep(0, 500)
+for(i in 1:length(set_20)){
+  print(i)
+  x <- as.data.frame(matrix(rnorm(100 * 20), ncol = 20))
+  set_20[i] <- narcs(gs(x, undirected = TRUE))
+  y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
+  set_100[i] <- narcs(gs(y, undirected = TRUE))
+}
+hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
+     main = "Spurious Conditional Dependence Relationships")
+hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
+legend("top", legend=c("20 features", "100 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))

### 2015-09-14 write_up.tex
edit before olga edit

@@
 analysis of large scale datasets.
+
 \section{Discussion}
@@
 detected between each pair of proteins, due to (direct and indirect)
-causal relationships. Figure 1, top panel, depicts these correlation as
+causal relationships. Figure 1, top panel, depicts these correlations as
 black lines, causal relationships are depicted as blue arrows.
@@
 between Raf and Erk. If we condition on Mek, then we assume that we know
-the value of Mek: perhaps we select only samples which have a high level
-of Mek, so Mek is fixed at level ``high''. Can the dependence between
+the value of Mek: perhaps we limit our analysis to samples which have a high level
+of Mek, so Mek is fixed at level "high". Can the dependence between
 Raf and Erk still be detected? If, due to noise, some of these Mek=high
@@
 correspondingly low? Based on the causal mechanism, Erk will more
-closely follow Mek, since Mek is its causal parent. Because of that, the
-Raf-Erk dependence, once conditioned on Mek, will be reduced or
-eliminated, thus indicating that the Raf-Erk edge can be removed. Using
-this approach, it is possible to keep only the causal edges as depicted
-by the black lines in Figure 1, bottom panel.
+closely follow Mek (i.e. its values will tend to be high), since Mek is 
+its causal parent. Because of that, the Raf-Erk dependence, once conditioned 
+on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge 
+can be removed. Using this approach, it is possible to keep only the causal edges 
+as depicted by the black lines in Figure 1, bottom panel.
 
-Computational methods that do causal inference in the context of an
-experiment that simultaneously measures multiple intercorrelated
-response variables all essentially have the same two step process; (1)
+Computational methods for causal inference in the context of
+experiments that simultaneously measure multiple intercorrelated
+features all rely on finding conditional dependence.  All of these algorithms
+essentially have the same two step process; (1)
 reduce the dense set of pairwise associations to a sparse set of
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
-\caption{As you increase the number of features, you get more high scoring
+\caption{Increasing the number of features results in more high scoring
 spurious correlations.}
@@
 
-As you increase the number of features you measure, the number of false
-positive detection of conditional dependence explodes. What this means
-is that the computational methods for causal inference will fail for the
+As we increased the number of features we measured, we saw a comparable increase in 
+the number of false positive detection of conditional dependence relationships. What this
+means is that the computational methods for causal inference will fail for the
 typical large-scale experiment, because when their input data contains a
@@
 
-The second problem is the requirement for interventions. As the number
+The second problem is the requirement for interventions. This is just as important as the 
+problem of spurious conditional dependence, but much easier to explain -- as the number
 of features grows, the number of interventions needed to infer causality
@@
 established and canonical. Assuming it, and other well established
-connections a priori can help reduce the number of possible
+connections are present in the data a priori can help reduce the number of possible
 associations, and enable more effective use of data. Knowledge about

### 2015-09-23 write_up.tex
presubmission draft

--- 2015-09-14 write_up.tex
@@
 \abbreviations{IR,NMR,UV}
-\keywords{American Chemical Society, \LaTeX}
+\keywords{causal inference, big data, causal networks, Bayesian networks}
 
@@
 
+\begin{abstract}
+  Causal inference -- the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics experiments in molecular biology studies.
+Statistical associations between quantitative measurements can reveal an enticing number of putative causal interactions, but when do such associations reflect the underlying causal biomolecular mechanisms?
+The goal of this perspective is to provide suggestions in non-technical terms for causal inference in large scale experiments, resulting from high throughput omics technologies such as MS-based proteomics.
+We describe the pitfalls of inference in large datasets, and suggest methods to reliably find causal associations despite these pitfalls.
+\end{abstract}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Introduction}
-Causal inference -- the task of uncovering regulatory relationships
-between components of biomolecular pathways and networks - is a primary
-goal of omics datasets in molecular biology studies. Statistical
-associations can reveal an alluring set of putative causal interactions,
-but when are such associations significant and reliable? And when can
-they be interpreted as causal? Our goal below is to provide suggestions
-for causal inference in large scale experiments, such as those resulting
-from high throughput omics technologies. We describe the pitfalls of
-large datasets, and suggest methods to reliably find causal associations
-despite these pitfalls.
+One of the tasks of causal inference methods is to distinguish statistical association resulting from causal relationships from association resulting from indirect relationships.
+High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
+Paradoxically, the large amount of data generated by these experiments  make this separation task more difficult, because, when many associations are examined, spurious associations -- associations arising purely from random chance -- may appear as strong as the true associations that result from actual indirect and causal relationships.
+In other words, large-scale experiments are more prone to hiding true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
+This problem can be addressed by improving experimental design and/or refining the biological question.
+Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
 
-High-throughput omics technologies enable large-scale experiments, with
-simultaneous measurements of multiple cellular components in a sample.
-The huge amount of data generated paradoxically make the task of causal
-inference much more difficult, because, when many associations are
-examined, some will appear significant just by chance. In other words,
-large-scale experiments are more prone to hiding the true signal and
-generating spurious associations, leading to increased false positive
-conclusions that causal relationships are present. This problem can be
-addressed by improving experimental design and/or refining the
-biological question. Below, we describe the process of elucidating
-statistical associations from data and suggest practical approaches for
-analysis of large scale datasets.
-\section{Discussion}
-\subsection{I. Statistical inference of causal relationships}
+\section{I. Statistical inference of causal relationships: Conditional independence and Interventions}
 
-In MAPK signaling, Raf causally affects active Mek levels (via
-phosphorylation), while Mek causally affects Erk. Imagine this
-relationship was unknown: could it be detected from measurements of
-these phosphoproteins? Given sufficient data, a correlation will be
-detected between each pair of proteins, due to (direct and indirect)
-causal relationships. Figure 1, top panel, depicts these correlation as
-black lines, causal relationships are depicted as blue arrows.
-
-To go from correlation to causality, we need to perform an intervention:
-for instance, we can use a small molecule inhibitor to block Mek
-activity. Resulting measurements show an impact on Erk but not on Raf,
-correctly orienting the two edges. Note that it was not necessary to
-intervene on each node in order to get the correct causal directions,
-from Raf to Mek and Mek to Erk. What about the noncausal edge? The edge
-linking Raf to Erk is not causal, because Raf's effect on Erk occurs
-only via Mek (and that affect is already represented, rendering the
-Raf-Erk edge superfluous). In statistical inference, this edge can be
-eliminated without the need of further interventions. How is this done?
-
-First, some terminology. A statistical association (such as a
-correlation) between two variables indicates that they are dependent.
-Sometimes we choose to assess a dependence in the context of the value
-of other system components, such as a third variable. In this case the
-dependence is called conditional dependence, because it is evaluated
-conditional on the state of other system components (i.e., the third
-variable). Let's see how this applies here, by examining the dependence
-between Raf and Erk. If we condition on Mek, then we assume that we know
-the value of Mek: perhaps we select only samples which have a high level
-of Mek, so Mek is fixed at level ``high''. Can the dependence between
-Raf and Erk still be detected? If, due to noise, some of these Mek=high
-samples have an intermediate or even low level of Raf, will Erk be
-correspondingly low? Based on the causal mechanism, Erk will more
-closely follow Mek, since Mek is its causal parent. Because of that, the
-Raf-Erk dependence, once conditioned on Mek, will be reduced or
-eliminated, thus indicating that the Raf-Erk edge can be removed. Using
-this approach, it is possible to keep only the causal edges as depicted
-by the black lines in Figure 1, bottom panel.
-
-Computational methods that do causal inference in the context of an
-experiment that simultaneously measures multiple intercorrelated
-response variables all essentially have the same two step process; (1)
-reduce the dense set of pairwise associations to a sparse set of
-conditional dependencies, (2) use information about the interventions
-included in the experiment to evaluate those conditional dependencies as
-evidence for potential causal relations.
-
-\subsection{II. Inference with large numbers of variables}
-
-Suppose we perform a large scale experiment, quantifying several
-thousands of phosphoproteins, including Raf, Mek and Erk. A typical
-large-scale experiments includes a very small number of interventions,
-and around 3 replicates. Statistical inference fails in this case,
-because the large number of variables measured result in a large number
-of spurious statistical associations, which appear as strong as true
-associations. The relationship between Raf, Mek and Erk, for instance,
-will be obscured by the many spurious relationships that they will each
-form with causally unrelated proteins.
+A biomolecular mechanism creates a causal relationship between molecular species, which may manifest as statistical association between omics measurements of these components.
+The goal of causal inference performed on these measurements is to elucidate these associations, and determine which of these reflect causal relationships.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/mapk.jpg}}
-\caption{In the MAPK signaling pathway, Raf regulates Mek, which regulates
-Erk. Top; the lines represent association - the activity of the kinases
-on this pathway are all correlated. Bottom: the lines represent
-conditional dependence, there is no conditional dependence edge from Raf
-to Erk because given you know Mek, knowing Raf tells you nothing about
-Erk. They are conditionally independent.}
+\centerline{\includegraphics[width=0.5\textwidth]{figs/egfr.png}}
+\caption{EGFR MAPK signaling pathway, an example of a pathway containing the phosphorylation cascade from Raf to Mek to Erk.
+The binding of ligand EGF to EGFR initiates a signal that leads to the cascade, which leads to gene transcription.
+This cascade, a component of several pathways, implies two causal relationships Raf --> MEK, Mek --> Erk.
+Raf and Erk have an indirect relationship, through Mek.}
 \label{mapk}
@@
 
-The problem of doing causal inference with large-scale experiments is
-two-fold. Firstly, sample size is typically far too small for the large
-number of features we measure. With the large-scale experiment described
-above, it is clear that viewing a scatter-plot between a pair of
-features that has only 3 points is uninformative -- even if the points
-correlated perfectly, 3 points is not enough to conclude anything about
-the relationship with any confidence. But even with hundreds of samples,
-we are still plagued by spurious associations when the number of
-features is so large, since more features means greater chance
-associations will occur at random.
+Take for example the MAPK signaling cascade, part of several signaling pathways such as the EGFR MAPK pathway in figure 1.
+Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
+Imagine these causal relationships were unknown: could they be detected from measurements of these proteins?
 
-To illustrate, we ran a simulation where we first simulate a 20 features
-dataset each with a 100 Gaussian random measurements, then increase the
-number of features to 500. In both cases the features are completely
-independent. This means if we quantify association using Pearson
-correlation, any correlation we find will be completely spurious. We
-record the maximum correlation between the 20 feature set and the 500
-feature set. We repeat this simulation 500 times.
+To illustrate causal inference in this context, we use a computational model of the phosphorylation cascade mechanism behind the causal and indirect relationships between Raf, Mek, and Erk.
+The Huang-Ferrell computational model of MAPK signaling models the key binding, phosphorylation, and dephosphorylation reactions of the cascade with mass action kinetics, and replicates MAPK key signaling behavior observed in nature.
+With this model we simulated an experiment with 100 simultaneous measurements of concentrations of phosphorylated Raf, and doubly phosphorylated Mek and Erk.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
-\caption{As you increase the number of features, you get more high scoring
-spurious correlations.}
+\centerline{\includegraphics[width=1\textwidth]{figs/mapk.png}}
+\caption{A: The causal inference workflow demonstrated on the MAPK cascade.
+In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from measurements.
+Step 2 reduces the correlation graph a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).
+Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).
+While step 1 has little requirements, step 2 requires replicates, and step 3 requires interventions (e.g.
+inhibitors).
+B, C, and D feature an experiment simulated from the Huang-Ferrell computational model of the phosphorylation cascade signaling mechanism.
+B: Pairwise plots of concentration values of phosphorylated (doubly phosphorylated for Mek and Erk) forms of each protein, and observed Spearman correlation values.
+The Raf -- Erk correlation is high, despite the fact Raf does not directly regulate Erk.
+C: Raf values vs Erk, where instances of high Mek values are highlighted in red.
+In the right panel of B, Mek is fixed at high levels, and the observed association between Raf and Erk disappear, indicating Raf and Erk are conditionally independent given Mek.
+D: After Raf is inhibited, the observed association between Raf and Mek remains while the association between Mek and Erk disappear.
+Given Mek is conditionally dependent on Raf and on Erk, the inhibition reveals causal flow from Raf to Mek to Erk.}
+\label{mapk}
+\end{figure}
+
+Figure 2, panel A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.
+Figure 2 panel B shows 2-way plots of the concentration values.
+We use Spearman correlation as a measure of association, the top-right half of the plot-matrix displays  the correlation values for each of the 3 pairs.
+These high values would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
+The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.
+What about the noncausal Raf--Erk edge?
+Despite the high Raf--Erk correlation, there is no direct causal mechanism between them (aside from the one via Mek, which is already accounted for via the Raf-->Mek and Mek-->Erk edges).
+In causal  inference, we can eliminate this "nuisance" edge.
+How is this done?
+
+First, some terminology.
+When the biological variables  - eg.
+abundances of specific proteins - have values that move together due to a common biologic mechanism, they are dependent.
+Like the mechanism itself, whether or not dependence between two variables exists is typically not known, measures of statistical association (eg  observed Pearson or Spearman correlation, or mutual information) demonstrate empirical evidence of the presence of dependencies between variables.
+The problem is that only some  dependencies reflect actual causal relationships, the majority of dependencies reflect indirect relationships One case of dependence through an indirect relationship is when intermediaries exist between an upstream and downstream variable, for example Raf -- Erk dependence comes from their indirect relationship through Mek.
+Another such case is one where two variables have a common cause, for example two proteins that don't causally affect one another are dependent if they have the same upstream regulator.
+However, when the behavior of the intermediaries or the shared regulators is controlled for, dependency from indirect relationships disappears.
+This is called conditional independence.
+Causal inference methods attempt to parse causal dependence from indirect dependence by finding empirical evidence of conditional independence.
+
+Let's see how this applies here, by examining the dependence between Raf and Erk.
+Figure 2 panel C compares Raf to Erk, highlighting measurements where Mek values are high (top quartile) in red.
+Note that when we subset the data to only the red measurements, in other words when we condition on Mek being high, we can no longer detect the dependence between Raf and Erk.
+Based on the causal mechanism, Erk is determined by Mek, so when we fix Mek at a certain level of activity, Erk follows Mek irrespective of Raf levels.
+This phenomenon -- wherein the dependence between Raf and Erk disappears upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.
+Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure 1 panel A.
+
+When we observe all the key variables in a system, variables become conditionally independent of everything except variables with which it shares a direct causal relationship; continuing with the protein signaling example this is a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.
+So after the statistical methods of causal inference pare down a set of observed correlations to a sparse set of putative causal conditional dependence relationships, we rely on interventions to determine the direction of causality.
+Figure 2 panel D illustrates the results of an intervention that targeted Mek with an inhibitor.
+The concentration of Mek is unaffected but its ability to phosphorylate other proteins is blocked.
+After this intervention is introduced, the Raf--Mek relationship is unchanged, while Erk drops to a low level.
+From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.
+With the intervention, we can finally move from the conditional dependence graph in panel A - step 3, to the causal graph in panel A - step 3.
+
+Computational methods for  causal inference follow the workflow  we describe in panel A for Raf-Mek-Erk, scaling it for experiments characterizing multiple intercorrelated variables.
+They (1) reduce the dense network of pair-wise correlations (or other measures of association) to a sparse network of putative conditional dependencies using empirical evidence of  conditional independence between pairs of variables,  then (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
+See Koller-Friedman for a detailed description of these methods and their theoretical underpinnings.
+
+\section{II. Inference with large numbers of variables}
+
+In the context of large-scale experiments, the principal challenge is spurious statistical associations.
+These differ from the "nuisance" associations that arise from non-causal indirect albeit actual dependence.
+Spurious associations are purely random artifacts of the data, reflecting no actual direct or indirect dependence.
+
+The opportunity for spurious association grows with the number of variables simultaneously measured relative to the number of replicates.
+Indeed, the typical large-scale experiments quantify a large number of analytes, and have a small number of replicates.
+Measuring more analytes means more analyte pairs where spurious correlation can appear, and having fewer replicates makes it more likely two unrelated analytes will have measurements that align by mere coincidence.
+The high amount of spurious associations in these experiments confound causal inference methods' task of evaluating true associations to the point of infeasibility.
+
+A simulation illustrates this problem.
+The simulation is inspired  by Fan et.
+al.
+but is translated to our context.
+We simulated an experiment, we collect 100 replicates with abundance measurements for 20 proteins.
+We then simulate a second experiment where we increase the number of proteins to 500.
+In both experiments, each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
+Because of this, the measurements  for a given protein are completely independent from those of any other, meaning any correlation we find will be completely spurious.
+After repeating the experiments 500 times, we plot the histograms of the highest Pearson correlation value found in each case; the light histogram for the 20-protein experiments, the dark histogram for the 500-protein experiments.
+The 500-protein histogram is to the left of the 20-protein histogram, demonstrating that increasing the number of proteins measured  results in spurious correlations with higher values.
+This is a clearly problem when high correlation is used as evidence of a true dependency.
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=1\textwidth]{figs/spurious_corr.png}}
+\caption{In a simulated experiment inspired by Fan et.
+al., we collect 100 replicates with abundance measurements for 20 proteins.
+We then simulate a second experiment where we increase the number of proteins to 500.
+In both experiments, each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
+Because of this, the measurements  for a given protein are completely independent from those of any other, meaning any correlation we find will be completely spurious.
+After repeating the experiments 500 times, we plot the histograms of the highest Pearson correlation value found in each case; the light histogram for the 20 protein experiments, the dark histogram for the 500 protein experiments.
+The 500-protein histogram is to the left of the 20-protein histogram, demonstrating that iIncreasing the number of proteins measured  results in spurious correlations with higher values, which is a problem when high correlation is considered evidence of a true relationship.}
 \label{spur_corr}
@@
 
-The increased incidence of spurious correlation means increased false
-positives in detecting conditional dependence relationships. To
-illustrate, we repeat the previous simulation, except now expanding from
-20 to only 100 features. In each of the 500 instances, we apply a search
-algorithm that iteratively performs conditional independence tests,
-returning a count of detected conditional dependence relationships. As
-before, since we simulated independent features, any conditional
-dependence relationship reported by this algorithm is a false positive,
-i.e.~it doesn't actually exist in the mechanism that generated the data.
+The increased incidence of spurious correlation impedes the performance of causal inference methods’ evaluation of conditional independence in the data.
+Specifically they result in more false positives in detection of putative causal conditional dependence relationships.
+To illustrate, we repeated the previous simulation, again starting with 20 proteins with 100 samples each, but this time expanding to only 100 proteins.
+Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003, which performs a series of conditional independence tests between the sets of proteins.
+We obtain a count value for the number of detected conditional dependence relationships.
+As before, since we randomly draw protein abundance measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm is a false positive, i.e.
+it doesn't actually exist in the mechanism that generated the data.
+We again repeat these experiments 500 times and generate histograms of the count values in the 20 protein and 100 protein case.
 
@@
 \centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_dep.png}}
-\caption{Fig3. Increasing the number of measured features means increasing the
-false positive detection of conditional dependence.}
+\caption{Results of experiments similar to those demonstrated in figure 2.
+We simulate an experiment with 100 replicates of abundance measurements for 20 proteins.
+We then simulate a second experiment where we increase the number of proteins to 100.
+We apply a causal inference-related algorithm described in Dimitris 2003, which evaluates evidence for conditional independence between sets of proteins, and count the number of pairwise conditional dependence relationships detected by the algorithm.
+In both 20-protein and 100-protein experiments, each protein in each replicate was assigned a value randomly drawn from a Gaussian distribution, thus the proteins are completely independent and any detection of a conditional dependence relationship is a false positive, i.e.
+it doesn't actually exist in the mechanism that generated the data.
+We again repeat these experiments 500 times and generate histograms of the conditional dependence detection counts in the 20-protein (light histogram) and 100 protein (dark histogram) case.
+The counts of false positive detections in the 20-protein experiments were all less than 20, but in the 100-protein experiments, they were typically greater than 100.
+The results demonstrate that increased incidence of spurious correlation resulting from increasing the number of measured proteins, leads to increased false positives in detection of putative causal conditional dependence relationships.}
 \label{spur_dep}
@@
 
-As you increase the number of features you measure, the number of false
-positive detection of conditional dependence explodes. What this means
-is that the computational methods for causal inference will fail for the
-typical large-scale experiment, because when their input data contains a
-large amount of features relative to sample size, they cannot reliably
-detect the conditional dependence relationships they need to infer
-causality.
+These results demonstrate that the increased incidence of spurious correlation resulting from increasing the number of measured proteins, leads to increased false positives in detection of putative causal conditional dependence relationships.
+This means that the computational methods for causal inference will fail for the typical large-scale experiment, because when their input data contains a large amount of proteins relative to sample size, they cannot reliably detect the conditional dependence relationships they need to infer causality.
 
-The second problem is the requirement for interventions. As the number
-of features grows, the number of interventions needed to infer causality
-grows, and performing sufficient perturbation experiments becomes
-infeasible.
+There is a another problem with having the small number of replicates typical of a large scale experiment.
+Step 1 of the workflow indicated in figure 2 panel A, evaluating the set of pairwise associations, will work with only a small amount of replicates.
+However, the methods for evaluating conditional independence used in step 2 require the statistical degrees of freedom provided by more replicates.
+While sparse approaches working with few replicates are available, quality of results decreases drastically as the number of replicates fall.
+Note the experiments demonstrated in figure 3 had 100 replicates, a great deal relative to the typical large scale experiments, and still problems emerged.
 
-\subsection{III. Inferring causality from omics experiments}
+Step 3 of the work flow requires interventions to infer causal relationships from evidence of conditional dependence.
+Measuring many proteins introduces a  challenge for the step 3 as well.
+As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.
+We showed with Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships, and indeed when measuring k variables a set of interventions on < k variables may be sufficient to infer causality.
+But along with having too few replicates, the typical large-scale experiment has far too few interventions.
+Also, if the step 2 results in too many false positives for conditional dependence, this will adversely affect the results of step 3 regardless of the size of the set of interventions.
 
-The problems outlined in section II paint a grim picture for causal
-learning in large datasets. Fortunately, these can be overcome, and
-effective causal learning can be a reality for large scale datasets. We
-proffer the approaches listed in \textit{The Causality Inference in Omics Tool List}
-below.
+\section{III. Approaches for inferring causality from omics experiments}
+
+The problems outlined in section II paint a grim picture for causal learning in large datasets.
+Fortunately, these can be overcome, and effective causal learning can be a reality for large scale datasets.
+We describe best practices  in the Causality Inference in Omics tools below.
 
 \begin{enumerate}
-\item \textit{Refine the biological problem}, thus limiting the number of features. If
-the broader biological system is well-studied, it may be possible design
-an experiment that focuses on a specific part of the system of interest,
-then ask more specific questions of the data, such as whether a
-particular edge or pathway is present. The more specific the question,
-the less data is needed overall to make solid statistical inferences.
-\item \textit{Measure more samples}. If feasible, high throughput measurements that
-measure many samples (not just many features) provide the statistical
-power to tell true associations from spurious associations. This
-intuition motivates the use technologies that have lower throughput but
-enable measurement of more samples (or gather more data points per
-sample), for causal modeling. An example is single cell mass cytometry,
-where many thousands of cells per sample provide ample statistical
-power.
-\item \textit{Use prior knowledge}, not just from experts but also from noisier
-sources, to improve the process of searching for conditional
-independence. The MAPK pathway for instance, mentioned earlier, is well
-established and canonical. Assuming it, and other well established
-connections a priori can help reduce the number of possible
-associations, and enable more effective use of data. Knowledge about
-canonical causal relationships can be sourced from pathway databases
-such as KEGG. Another example is contextual information, such as spatial
-relations between phosphoproteins in the cell -- causal inference
-algorithms can be formulated to weigh evidence of conditional dependence
-differently depending on whether phosphoproteins are from the same or
-different spatial compartments. Employing prior knowledge that provides
-evidence for the regulatory network reduces the number of connections
-and causal arcs under consideration, which allows the available data to
-make reliable statistical inferences.
-\item \textit{Employ targeted interventions selectively}. Targeted interventions
-perturb individual components of the system. An example is small
-molecule inhibitors which block the causal influence of a specific
-phoshoprotein on downstream components. Since it is usually prohibitive
-to perturb every target component of a system, it is strategic to
-prioritize application of targeted interventions to parts of the system
-that have the most potential for new discoveries. Selection of these
-perturbation targets can be based on prior knowledge - e.g.~knowledge of
-which components are crucial players in the system of interest - or they
-can be applied iteratively, after an initial statistical analysis has
-revealed areas of the network in which causal inferences are not
-possible based on existing data. For instance, a resulting model graph
-with undirected edges can be inspected to reveal which nodes have
-potential to reveal the most causality if perturbed.
-\item \textit{Consider broad-scale interventions}. Traditionally, experiments consider
-a biological response to a particular stimulus, possibly in the context
-of an inhibitor. In contrast, broad-scale interventions sacrifice
-specificity to perturb many features in the system simultaneously. The
-advantage of this approach is that it can enable elucidation of
-causality across the entire system. Just as one intervention was
-sufficient to infer causality between 3 proteins Raf, Mek, and Erk,
-perturbing many things at once creates cascades of causal direction
-orientation across the broader network. This includes varying
-experimental conditions to activate multiple pathways. For example,
-signals from endocrine, paracrine, and autocrine ligands elicit various
-signaling responses in hepatocytes, thus interventions that cover this
-range of signals gives the best picture of the broader causal network of
-hepatocyte signaling. Similarly, interventions that go beyond
-receptor-level and perturb multiple components of the system bring
-cascading causal direct orientation deeper into the network. In contrast
-to targeted interventions, while it is difficult to know in advance what
-information will provided by broad-scale interventions, they provide
-more causal inference bang for your intervention buck.
+\item \textit{Refine the biological problem}, thus limiting the number of variables.
+The length of the list of identified and quantified analytes is not as important as the focus of the investigation and the quantity of measurements on important parts of the system.
+If the broader biological system is well-studied, it may be possible design to an experiment that focuses on a specific part of the system of interest, then ask more distinct questions of the data, such as whether a particular edge or pathway is present.
+The more specific the question, the less data needed overall to make solid statistical inferences.
+\item \textit{Measure more samples}.
+If feasible, high throughput measurements that measure many samples (not just many proteins) provide the statistical power to tell true associations from spurious associations.
+This intuition motivates the use of technologies that measure fewer variables but enable measurement of more samples (or gather more data points per sample), for causal modeling.
+A related  example is  use of single cell data for inference (e.g.
+from single cell  mass cytometry), where  many thousands of cells per sample provide ample statistical power.
+See Sachs et al 2005 for an in depth case study.
+\item \textit{Use prior knowledge}, not just from experts but also from noisier sources, to improve the process of searching for conditional independence.
+The MAPK pathway for instance, mentioned earlier, is canonical and well established.
+The process of inferring causal connections is helped by assuming a priori that such canonical connections exist.
+This is because knowledge of parts of the underlying regulatory network reduces the remaining number of possible associations, and enables more effective use of data.
+The number of total associations examined goes down, enabling increased confidence in statistical associations that are found.
+Knowledge about canonical causal relationships can be sourced from pathway databases such as KEGG.
+Another example of prior knowledge is contextual information, such as spatial relations between  signaling proteins in the cell -- causal inference algorithms can be formulated to weigh evidence of conditional dependence differently depending on whether proteins are from the same or different spatial compartments.
+\item \textit{Employ targeted interventions selectively}.
+Targeted interventions perturb individual components of the system.
+An example is small molecule inhibitors which block the causal influence of a specific protein on downstream components.
+Since it is usually prohibitive to perturb every target component of a system, it is strategic to prioritize application of targeted interventions to parts of the system that have the most potential for new discoveries.
+Selection of these perturbation targets can be based on prior knowledge - e.g.
+knowledge of which components are crucial players in the system of interest - or they can be applied iteratively, after an initial statistical analysis has revealed areas of the network in which causal inferences are not possible based on existing data.
+For instance, a resulting model graph with undirected edges can be inspected to reveal which nodes have potential to reveal the most causality if perturbed.
+\item \textit{Consider broad-scale interventions}.
+Traditionally, experiments characterize a biological response to a particular stimulus, possibly in the context of an inhibitor.
+In contrast, broad-scale interventions sacrifice specificity to perturb many features in the system simultaneously.
+The advantage of this approach is that it may enable elucidation of  causality across the entire system.
+Just as one intervention was sufficient to infer causality between three proteins (Raf, Mek, and Erk), perturbing many things at once creates cascades of causal direction orientation across the broader network.
+This includes varying experimental conditions to activate multiple pathways.
+For example, signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling.
+Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.
+Although they do not provide specific information about the downstream effects of stimulation,  broad-scale interventions can provide more causal inference bang for your intervention buck.
+
 \end{enumerate}
@@
 \#3, prior biological knowledge, can be used to prioritize what to
-target with that limited set of interventions.
+target with that limited set of interventions.  Causal inference becomes possible when using these tools in combination to design experiments.
+
+\section{Acknowledgement}
+We acknowledge the participants of Dagstuhl seminar 15351 "Computational Mass Spectrometry" (December 2015) (http://www.dagstuhl.de/de/programm/kalender/semhp/?semnr=15351) for their contributions to the discussion on computational manuscripts.
+In particular, we thank the participants of the Correlation vs Causality discussion.
+We are grateful for the support of Schloss Dagstuhl - Leibniz-Zentrum fur Informatik for supporting the seminar that led to this work.
 

### 2015-09-24 scripts_for_figures.R
prior to olga edits

@@ -12,10 +12,10 @@ for(i in 1:500){
   diag(cor_y) <- 0
   set_500[i] <- max(cor_y)
 }
-hist(set_500, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL, xlim = c(min(set_20) , max(set_500) + .1 ), 
-     main = "Highest Spurious Correlation")
+hist(set_500, main = NULL, col = rgb(0.1,0.1,0.1,0.5), xlim = c(min(set_20) , max(set_500) + .1 ),
+     xlab = "Highest measured correlation between protein pairs")
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("topr", legend=c("20 features", "500 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+legend("topr", legend=c("20 proteins", "500 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
 
 ## conditional dependence figure
 library(bnlearn)
@@ -28,7 +28,7 @@ for(i in 1:length(set_20)){
   y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
   set_100[i] <- narcs(gs(y, undirected = TRUE))
 }
-hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = NULL,  xlim = c(min(set_20), max(set_100)), 
-     main = "Spurious Conditional Dependence Relationships")
+hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = "Count of false positive detections of conditional dependence",  xlim = c(min(set_20), max(set_100)), 
+     main = NULL)
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("top", legend=c("20 features", "100 features"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+legend("top", legend=c("20 proteins", "100 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))

### 2015-09-25 write_up.tex
prior to olga edits
post olga edit

@@
 \newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+\usepackage{color}
+\def\todo#1{{\color{red}[TODO: #1]}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \author{Robert O. Ness}
 \email{nessr@purdue.edu}
-\affiliation[Purdue University]{Department of Statistics, Purdue University, West Lafayette}
+\affiliation[Purdue University]{Purdue University Department of Statistics, West Lafayette}
+\alsoaffiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
+\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
 \author{Karen Sachs}
 \affiliation[Stanford University]{School of Medicine, Stanford University, Palo Alto}
-\email{karens1@stanford.edu}
 \author{Olga Vitek}
-\email{o.vitek@neu.edu}
-\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Palo Alto}
+\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
 
@@
  \title[]
-  {From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments}
+   {From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations}
 
@@
 \begin{abstract}
-  Causal inference -- the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics experiments in molecular biology studies.
+  Causal inference -- the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of many high throughput investigations.
 Statistical associations between quantitative measurements can reveal an enticing number of putative causal interactions, but when do such associations reflect the underlying causal biomolecular mechanisms?
-The goal of this perspective is to provide suggestions in non-technical terms for causal inference in large scale experiments, resulting from high throughput omics technologies such as MS-based proteomics.
-We describe the pitfalls of inference in large datasets, and suggest methods to reliably find causal associations despite these pitfalls.
+The goal of this perspective is to provide suggestions for causal inference in large scale experiments, resulting from high throughput technologies such as mass spectrometry-based proteomics.
+We describe the pitfalls of inference in large datasets, and suggest methods to overcome these pitfalls and reliably find regulatory associations, in non-technical terms that are accessible to life scientists with a limited statistical background.
 \end{abstract}
@@
 \section{Introduction}
-One of the tasks of causal inference methods is to distinguish statistical association resulting from causal relationships from association resulting from indirect relationships.
-High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample.
-Paradoxically, the large amount of data generated by these experiments  make this separation task more difficult, because, when many associations are examined, spurious associations -- associations arising purely from random chance -- may appear as strong as the true associations that result from actual indirect and causal relationships.
-In other words, large-scale experiments are more prone to hiding true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.
-This problem can be addressed by improving experimental design and/or refining the biological question.
-Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.
 
-\section{I. Statistical inference of causal relationships: Conditional independence and Interventions}
+Causal inference \cite{pearl2009causality} elucidates statistical associations, which result from the underlying biomolecular mechanistic relationships.
+Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of such biomolecular systems on a large scale.
+Paradoxically, the large amount of data generated by these experiments make the task of causal inference more difficult.
+When many associations are examined, spurious associations -- associations arising purely from random chance -- may appear as strong, and the true signal may be obfuscated, leading to increased false discoveries of  putative causal events.
+This problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
+Below, we describe the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale datasets.
 
-A biomolecular mechanism creates a causal relationship between molecular species, which may manifest as statistical association between omics measurements of these components.
-The goal of causal inference performed on these measurements is to elucidate these associations, and determine which of these reflect causal relationships.
+\section{I. Small-scale statistical inference of causal relationships: conditional independence and interventions}
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/egfr.png}}
-\caption{EGFR MAPK signaling pathway, an example of a pathway containing the phosphorylation cascade from Raf to Mek to Erk.
-The binding of ligand EGF to EGFR initiates a signal that leads to the cascade, which leads to gene transcription.
-This cascade, a component of several pathways, implies two causal relationships Raf --> MEK, Mek --> Erk.
-Raf and Erk have an indirect relationship, through Mek.}
-\label{mapk}
+\centerline{\includegraphics[width=0.4\textwidth]{figs/egfr.png}}
+\caption{EGFR MAPK signaling pathway \cite{holbro2004erbb}, an example of a pathway containing the phosphorylation cascade from Raf to Mek to Erk.
+The binding of ligand EGF to EGFR initiates a signal that leads to the cascade, which in turn regulates transcription.
+This cascade implies two direct causal relationships, namely Raf --> MEK, and Mek --> Erk.
+Raf and Erk have an indirect causal relationship, through Mek.\label{mapk}}
 \end{figure}
 
-Take for example the MAPK signaling cascade, part of several signaling pathways such as the EGFR MAPK pathway in figure 1.
-Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk.
-Imagine these causal relationships were unknown: could they be detected from measurements of these proteins?
+Consider, e.g.
+the MAPK signaling cascade, which is part of several signaling pathways such as the EGFR MAPK pathway in Figure \ref{mapk}.
+In this cascade Raf causally affects the level of active (i.e., phosphorylated) Mek, while Mek causally affects Erk.
+Imagine these causal relationships were unknown: could they be detected from quantitative measurements on these phosphoproteins?
 
-To illustrate causal inference in this context, we use a computational model of the phosphorylation cascade mechanism behind the causal and indirect relationships between Raf, Mek, and Erk.
-The Huang-Ferrell computational model of MAPK signaling models the key binding, phosphorylation, and dephosphorylation reactions of the cascade with mass action kinetics, and replicates MAPK key signaling behavior observed in nature.
-With this model we simulated an experiment with 100 simultaneous measurements of concentrations of phosphorylated Raf, and doubly phosphorylated Mek and Erk.
+To illustrate the process of causal inference in this context, we simulate artificial data using the computational Huang-Ferrell model of this cascade.
+The model represents the key binding, phosphorylation, and dephosphorylation reactions of the cascade with mass action kinetics, and replicates the MAPK key signaling behavior observed in nature.
+With this model we simulated an experiment with 100 replicate biological samples, and measurements of concentration (umol) of phosphorylated Raf, and doubly phosphorylated Mek and Erk in each sample.
 
@@
 \centerline{\includegraphics[width=1\textwidth]{figs/mapk.png}}
-\caption{A: The causal inference workflow demonstrated on the MAPK cascade.
-In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from measurements.
-Step 2 reduces the correlation graph a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).
-Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).
-While step 1 has little requirements, step 2 requires replicates, and step 3 requires interventions (e.g.
-inhibitors).
-B, C, and D feature an experiment simulated from the Huang-Ferrell computational model of the phosphorylation cascade signaling mechanism.
+\caption{A: Overview of the 3 steps of causal inference, illustrated for the MAPK signaling cascade.
+B, C, and D feature an experiment simulated from the Huang-Ferrell computational model of the phosphorylation cascade.
 B: Pairwise plots of concentration values of phosphorylated (doubly phosphorylated for Mek and Erk) forms of each protein, and observed Spearman correlation values.
 The Raf -- Erk correlation is high, despite the fact Raf does not directly regulate Erk.
-C: Raf values vs Erk, where instances of high Mek values are highlighted in red.
-In the right panel of B, Mek is fixed at high levels, and the observed association between Raf and Erk disappear, indicating Raf and Erk are conditionally independent given Mek.
+C: Concentrations of Raf versus Erk, where samples corresponding to high Mek are highlighted in red.
+In the right panel of B, only points with high levels of Mek are display (i.e.
+condition on Mek being high), and the observed association between Raf and Erk disappear, indicating Raf and Erk are conditionally independent given Mek.
 D: After Raf is inhibited, the observed association between Raf and Mek remains while the association between Mek and Erk disappear.
 Given Mek is conditionally dependent on Raf and on Erk, the inhibition reveals causal flow from Raf to Mek to Erk.}
-\label{mapk}
+\label{mapkInference}
 \end{figure}
 
-Figure 2, panel A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.
-Figure 2 panel B shows 2-way plots of the concentration values.
-We use Spearman correlation as a measure of association, the top-right half of the plot-matrix displays  the correlation values for each of the 3 pairs.
-These high values would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
+Figure \ref{mapkInference}A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.
+In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.
+Step 2 reduces the correlation graph to a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).
+Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).
+While step 1 has little requirements, step 2 requires replicate biological samples, and step 3 requires systematic interventions (e.g.
+with protein inhibitors).
+
+Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the quantified protein concentrations, and Spearman correlations to quantify the extent of the associations.
+The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
 The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.
@@
 Despite the high Raf--Erk correlation, there is no direct causal mechanism between them (aside from the one via Mek, which is already accounted for via the Raf-->Mek and Mek-->Erk edges).
-In causal  inference, we can eliminate this "nuisance" edge.
+In causal  inference, our goal is to eliminate this "nuisance" edge.
 How is this done?
 
-First, some terminology.
-When the biological variables  - eg.
-abundances of specific proteins - have values that move together due to a common biologic mechanism, they are dependent.
-Like the mechanism itself, whether or not dependence between two variables exists is typically not known, measures of statistical association (eg  observed Pearson or Spearman correlation, or mutual information) demonstrate empirical evidence of the presence of dependencies between variables.
-The problem is that only some  dependencies reflect actual causal relationships, the majority of dependencies reflect indirect relationships One case of dependence through an indirect relationship is when intermediaries exist between an upstream and downstream variable, for example Raf -- Erk dependence comes from their indirect relationship through Mek.
-Another such case is one where two variables have a common cause, for example two proteins that don't causally affect one another are dependent if they have the same upstream regulator.
-However, when the behavior of the intermediaries or the shared regulators is controlled for, dependency from indirect relationships disappears.
-This is called conditional independence.
-Causal inference methods attempt to parse causal dependence from indirect dependence by finding empirical evidence of conditional independence.
+To describe Step 2 of causal inference, we need some terminology.
+In statistical language, the quantified proteins are called {\it variables}.
+If the values of concentrations of two proteins vary between the biological samples in a coordinated manner, which is due to a common biological mechanism, the proteins are {\it statistically dependent}.
+Just like the underlying biological mechanism is typically unknown, the presence of statistical dependence between two or more proteins is also unknown.
+Measures of statistical association (e.g.,  the Spearman correlation in Figure \ref{mapkInference}B or, alternatively, Pearson correlation or mutual information) demonstrate empirical evidence of the presence of such dependence.
+However, high associations are not a proof of direct causal relationships.
+In fact, the majority of statistical dependencies observed in experimental data reflect indirect associations.
+For example, a statistical association between an upstream and an indirect downstream protein can be due to the presence of intermediate proteins, such as the statistical association between Raf -- Erk is due to their indirect relationship through Mek.
+Another such example is a case where two proteins have a common regulator.
+In this case the two proteins are statistically associated through the regulator, without causally affecting one another.
 
-Let's see how this applies here, by examining the dependence between Raf and Erk.
-Figure 2 panel C compares Raf to Erk, highlighting measurements where Mek values are high (top quartile) in red.
-Note that when we subset the data to only the red measurements, in other words when we condition on Mek being high, we can no longer detect the dependence between Raf and Erk.
-Based on the causal mechanism, Erk is determined by Mek, so when we fix Mek at a certain level of activity, Erk follows Mek irrespective of Raf levels.
-This phenomenon -- wherein the dependence between Raf and Erk disappears upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.
-Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure 1 panel A.
+The direct causal events can be distinguished from other undesirable types of associations by controlling for (hold constant) the intermediaries and the common regulators.
+The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.
+Causal inference searches for empirical evidence of conditional independence, to distinguish causal events from indirect associations.
 
-When we observe all the key variables in a system, variables become conditionally independent of everything except variables with which it shares a direct causal relationship; continuing with the protein signaling example this is a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.
-So after the statistical methods of causal inference pare down a set of observed correlations to a sparse set of putative causal conditional dependence relationships, we rely on interventions to determine the direction of causality.
-Figure 2 panel D illustrates the results of an intervention that targeted Mek with an inhibitor.
-The concentration of Mek is unaffected but its ability to phosphorylate other proteins is blocked.
-After this intervention is introduced, the Raf--Mek relationship is unchanged, while Erk drops to a low level.
+Let's see how this applies to the MAPK signaling cascade.
+We can examine the nature of statistical association between Raf and Erk.
+Figure \ref{mapkInference}C compares Raf to Erk, highlighting in red the biological samples where concentrations of Mek are high (here, set to the top quartile).
+Note that when we subset the data to only the red samples (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk.
+This has a mechanistic explanation.
+As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek.
+Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.
+This phenomenon -- the disappearence of the association between Raf and Erk upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.
+Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure \ref{mapk}A.
+
+When the experiment quantifies all the key variables in a biological system, Step 2 above elucidates conditional dependence-based associations between causally related variables.
+In the MAPK example, this corresponds to a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.
+However, at this step the direction of the regulation remains unknown.
+Inference of the direction of the chain of events requires that the experimental design involves interventions.
+Figure \ref{mapkInference}D illustrates the results of Step 3, in the case of an intervention that targeted Mek with an inhibitor.
+With this intervention the concentration of Mek is unaffected, however its ability to phosphorylate other proteins is blocked.
+After this intervention is introduced the Raf--Mek relationship is unchanged, while Erk drops to a low level.
 From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.
-With the intervention, we can finally move from the conditional dependence graph in panel A - step 3, to the causal graph in panel A - step 3.
+With the intervention, we can finally move from the conditional dependence graph in panel A - step 2 to the causal graph in panel A - step 3.
 
-Computational methods for  causal inference follow the workflow  we describe in panel A for Raf-Mek-Erk, scaling it for experiments characterizing multiple intercorrelated variables.
-They (1) reduce the dense network of pair-wise correlations (or other measures of association) to a sparse network of putative conditional dependencies using empirical evidence of  conditional independence between pairs of variables,  then (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.
-See Koller-Friedman for a detailed description of these methods and their theoretical underpinnings.
+In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related variables.
+Step 1 creates a dense network of pairwise associations.
+Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of  conditional independence between pairs of variables.
+Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these conditional dependencies as evidence for potential causal events.
+See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
+Several of these algorithms are implemented in the R package \textit{bnlearn} \cite{scutari2009learning}.
 
-\section{II. Inference with large numbers of variables}
+\section{II. Large-scale statistical inference of causal relationships: challenges of scaling up}
 
-In the context of large-scale experiments, the principal challenge is spurious statistical associations.
-These differ from the "nuisance" associations that arise from non-causal indirect albeit actual dependence.
-Spurious associations are purely random artifacts of the data, reflecting no actual direct or indirect dependence.
+A typical high-throughput experiments includes a small number of interventions, a small number of biological replicates, and a large number of analytes.
+This creates challenges in each of the 3 steps of causal inference above.
 
-The opportunity for spurious association grows with the number of variables simultaneously measured relative to the number of replicates.
-Indeed, the typical large-scale experiments quantify a large number of analytes, and have a small number of replicates.
-Measuring more analytes means more analyte pairs where spurious correlation can appear, and having fewer replicates makes it more likely two unrelated analytes will have measurements that align by mere coincidence.
-The high amount of spurious associations in these experiments confound causal inference methods' task of evaluating true associations to the point of infeasibility.
+In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes.
+A large number of proteins  leads to a large number of spurious statistical associations that arise without any biological justification, and purely as an artifact of random chance.
+Systematic pairwise relationships such as between Raf, Mek and Erk in the MAPK pathway will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.
 
-A simulation illustrates this problem.
-The simulation is inspired  by Fan et.
-al.
-but is translated to our context.
-We simulated an experiment, we collect 100 replicates with abundance measurements for 20 proteins.
-We then simulate a second experiment where we increase the number of proteins to 500.
-In both experiments, each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
-Because of this, the measurements  for a given protein are completely independent from those of any other, meaning any correlation we find will be completely spurious.
-After repeating the experiments 500 times, we plot the histograms of the highest Pearson correlation value found in each case; the light histogram for the 20-protein experiments, the dark histogram for the 500-protein experiments.
-The 500-protein histogram is to the left of the 20-protein histogram, demonstrating that increasing the number of proteins measured  results in spurious correlations with higher values.
-This is a clearly problem when high correlation is used as evidence of a true dependency.
+We illustrate this problem with a computer simulation, inspired by Fan et.
+al \cite{fan2014challenges} but translated to our context.
+First, we simulated an experiment that quantifies the abundances of 20 proteins in 100 biological samples.
+Second, we simulates another experiment where the number of proteins was increased to 500.
+In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
+In other words, we do not expect any biologically meaningful associations in these data.
+We repeated each of these simulations 500 times.
+Figure \ref{spur_corr} plots the histograms of the highest Pearson correlation found across any pair of proteins in the 500 simulations, separately for each experiment.
+As can be seen, the experiment with 500 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlations.
+This is a clearly problem when high correlation is used as an initial evidence of a biological function.
 
@@
 \centerline{\includegraphics[width=1\textwidth]{figs/spurious_corr.png}}
-\caption{In a simulated experiment inspired by Fan et.
-al., we collect 100 replicates with abundance measurements for 20 proteins.
-We then simulate a second experiment where we increase the number of proteins to 500.
-In both experiments, each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
-Because of this, the measurements  for a given protein are completely independent from those of any other, meaning any correlation we find will be completely spurious.
-After repeating the experiments 500 times, we plot the histograms of the highest Pearson correlation value found in each case; the light histogram for the 20 protein experiments, the dark histogram for the 500 protein experiments.
-The 500-protein histogram is to the left of the 20-protein histogram, demonstrating that iIncreasing the number of proteins measured  results in spurious correlations with higher values, which is a problem when high correlation is considered evidence of a true relationship.}
+\caption{Simulated experiments quantifying 20 (or 500) unrelated proteins in 100 biological samples, each repeated 500 times.
+The plots are the histograms of the highest Pearson correlation found across any pair of proteins, calculated over the 500 simulations.
+Increasing the number of proteins results in higher spurious associations.}
 \label{spur_corr}
@@
 
-The increased incidence of spurious correlation impedes the performance of causal inference methods’ evaluation of conditional independence in the data.
-Specifically they result in more false positives in detection of putative causal conditional dependence relationships.
-To illustrate, we repeated the previous simulation, again starting with 20 proteins with 100 samples each, but this time expanding to only 100 proteins.
-Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003, which performs a series of conditional independence tests between the sets of proteins.
-We obtain a count value for the number of detected conditional dependence relationships.
-As before, since we randomly draw protein abundance measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm is a false positive, i.e.
-it doesn't actually exist in the mechanism that generated the data.
-We again repeat these experiments 500 times and generate histograms of the count values in the 20 protein and 100 protein case.
+Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.
+The spurious correlations result in more false positives when detecting putative causal conditional dependence relationships.
+To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.
+Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003 \cite{margaritis2003learning}, which performs a series of conditional independence tests between the sets of proteins, and count the number of detected conditional dependence relationships.
+As before, since we randomly draw protein abundance measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm has no biological justification.
+We again repeat these experiments 500 times.
+Figure \ref{spur_dep} shows the histograms of the counts of false positive detections of conditional dependence.
+The results demonstrate that an increase in the number of proteins leads to an increase in false positive detection of putative causal conditional dependence relationships.
+This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably detect the conditional dependence relationships needed to infer causality.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_dep.png}}
-\caption{Results of experiments similar to those demonstrated in figure 2.
-We simulate an experiment with 100 replicates of abundance measurements for 20 proteins.
-We then simulate a second experiment where we increase the number of proteins to 100.
-We apply a causal inference-related algorithm described in Dimitris 2003, which evaluates evidence for conditional independence between sets of proteins, and count the number of pairwise conditional dependence relationships detected by the algorithm.
-In both 20-protein and 100-protein experiments, each protein in each replicate was assigned a value randomly drawn from a Gaussian distribution, thus the proteins are completely independent and any detection of a conditional dependence relationship is a false positive, i.e.
-it doesn't actually exist in the mechanism that generated the data.
-We again repeat these experiments 500 times and generate histograms of the conditional dependence detection counts in the 20-protein (light histogram) and 100 protein (dark histogram) case.
-The counts of false positive detections in the 20-protein experiments were all less than 20, but in the 100-protein experiments, they were typically greater than 100.
-The results demonstrate that increased incidence of spurious correlation resulting from increasing the number of measured proteins, leads to increased false positives in detection of putative causal conditional dependence relationships.}
+\centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
+\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positive detections of conditional dependence.}
 \label{spur_dep}
@@
 
-These results demonstrate that the increased incidence of spurious correlation resulting from increasing the number of measured proteins, leads to increased false positives in detection of putative causal conditional dependence relationships.
-This means that the computational methods for causal inference will fail for the typical large-scale experiment, because when their input data contains a large amount of proteins relative to sample size, they cannot reliably detect the conditional dependence relationships they need to infer causality.
+An additional problem for Step 2 is a typically small number of biological replicates that these experiments have as compared to the number of analytes.
+While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 requires that the number of replicates is equal or exceeds the number of analytes.
+Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins.
+This number of samples per analyte is much larger than in a typical high-throughput experiment.
+While sparse approaches working around ths problem are available, quality of their results decreases drastically as the number of replicates fall.
 
-There is a another problem with having the small number of replicates typical of a large scale experiment.
-Step 1 of the workflow indicated in figure 2 panel A, evaluating the set of pairwise associations, will work with only a small amount of replicates.
-However, the methods for evaluating conditional independence used in step 2 require the statistical degrees of freedom provided by more replicates.
-While sparse approaches working with few replicates are available, quality of results decreases drastically as the number of replicates fall.
-Note the experiments demonstrated in figure 3 had 100 replicates, a great deal relative to the typical large scale experiments, and still problems emerged.
-
-Step 3 of the work flow requires interventions to infer causal relationships from evidence of conditional dependence.
-Measuring many proteins introduces a  challenge for the step 3 as well.
+Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional dependence.
+Quantifying a large number of analytes introduces a  challenge for this step as well.
 As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.
-We showed with Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships, and indeed when measuring k variables a set of interventions on < k variables may be sufficient to infer causality.
+We showed with Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships, and indeed when measuring $k$ variables a set of interventions on $< k$ variables may be sufficient to infer causality.
 But along with having too few replicates, the typical large-scale experiment has far too few interventions.
@@
 
-\section{III. Approaches for inferring causality from omics experiments}
+\section{III. Approaches for inferring causality from high-throughput experiments}
 
@@
 Fortunately, these can be overcome, and effective causal learning can be a reality for large scale datasets.
-We describe best practices  in the Causality Inference in Omics tools below.
+We describe best practices  in the Causality Inference in Omics  below.
 
 \begin{enumerate}
-\item \textit{Refine the biological problem}, thus limiting the number of variables.
-The length of the list of identified and quantified analytes is not as important as the focus of the investigation and the quantity of measurements on important parts of the system.
-If the broader biological system is well-studied, it may be possible design to an experiment that focuses on a specific part of the system of interest, then ask more distinct questions of the data, such as whether a particular edge or pathway is present.
-The more specific the question, the less data needed overall to make solid statistical inferences.
+\item \textit{Limiting the number of analytes}, and use a small subset of biologically relevant and technologically accurate measurements.
+The length of the list of identified and quantified analytes is not as important as the focus of the investigation and the quality of measurements on important parts of the system.
+If the broader biological system is well understood, it may be possible design to a targeted experiment that focuses on a specific part of the system of interest, then ask more distinct questions of the data, such as whether a particular edge or pathway is present.
+The more specific the question, the less data needed overall to make solid causal inference.
 \item \textit{Measure more samples}.
-If feasible, high throughput measurements that measure many samples (not just many proteins) provide the statistical power to tell true associations from spurious associations.
+If feasible, high throughput measurements that measure many samples (not just many proteins) provide the statistical power to distinguish true associations from spurious associations.
 This intuition motivates the use of technologies that measure fewer variables but enable measurement of more samples (or gather more data points per sample), for causal modeling.
@@
 See Sachs et al 2005 for an in depth case study.
+\cite{sachs2005causal}
 \item \textit{Use prior knowledge}, not just from experts but also from noisier sources, to improve the process of searching for conditional independence.
@@
 \#3, prior biological knowledge, can be used to prioritize what to
-target with that limited set of interventions.  Causal inference becomes possible when using these tools in combination to design experiments.
+target with that limited set of interventions.  Causal inference becomes possible when using these tools in combination with a sound experimental design.
 
-\section{Acknowledgement}
+\begin{acknowledgement}
 We acknowledge the participants of Dagstuhl seminar 15351 "Computational Mass Spectrometry" (December 2015) (http://www.dagstuhl.de/de/programm/kalender/semhp/?semnr=15351) for their contributions to the discussion on computational manuscripts.
-In particular, we thank the participants of the Correlation vs Causality discussion.
-We are grateful for the support of Schloss Dagstuhl - Leibniz-Zentrum fur Informatik for supporting the seminar that led to this work.
-
+\end{acknowledgement}
+\bibliography{write_up}
 \end{document}

### 2015-09-25 scripts_for_figures.R

@@ -1,3 +1,30 @@
+.data <- read.delim("Downloads/holygrail.txt") %>%
+  select(X.P.MAPKKK., X.PP.MAPKK., X.PP.MAPK.) %>%
+  transmute(Raf = as.numeric(X.P.MAPKKK.), 
+            Mek = as.numeric(X.PP.MAPKK.), 
+            Erk = as.numeric(X.PP.MAPK.)) %>%
+  plot
+
+library(bnlearn)
+.data <- read.delim("Downloads/holygrail.txt") %>%
+  select(X.P.MAPKKK., X.PP.MAPKK., X.PP.MAPK.) %>%
+  transmute(Raf = as.numeric(X.P.MAPKKK.), 
+            Mek = as.numeric(X.PP.MAPKK.), 
+            Erk = as.numeric(X.PP.MAPK.)) %>%
+  plot
+  
+d.data <- discretize(.data, method = "hartemink", breaks = 10)
+.data$dMek <- d.data$Mek    
+cdplot(dMek ~ Raf, data = .data)
+
+  gs(undirected = TRUE, debug = TRUE) %>%
+  graphviz.plot
+  lapply(function(item) {
+    x <- jitter(item, 2000)
+    ifelse(x > 0 , x , 0)
+  }) %>%
+  as.data.frame 
+
 # R scripts for generating spurious correlation and conditional dependence histograms.
 ## correlation figure
 set_20 <- rep(0, 500)
@@ -32,3 +59,6 @@ hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = "Count of false positive detect
      main = NULL)
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
 legend("top", legend=c("20 proteins", "100 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+
+
+

### 2015-09-25 scripts_for_figures.R
changes to figure scripts

@@ -1,19 +1,34 @@
-.data <- read.delim("Downloads/holygrail.txt") %>%
-  select(X.P.MAPKKK., X.PP.MAPKK., X.PP.MAPK.) %>%
-  transmute(Raf = as.numeric(X.P.MAPKKK.), 
-            Mek = as.numeric(X.PP.MAPKK.), 
-            Erk = as.numeric(X.PP.MAPK.)) %>%
-  plot
+library(dplyr)
+library(bnlearn)
+panel.cor <- function(x, y, digits=3, prefix="",  ...){
+  usr <- par("usr"); on.exit(par(usr))
+  par(usr = c(0, 1, 0, 1))
+  r <- abs(cor(x, y, method = "spearman"))
+  txt <- format(c(r, 0.123456789), digits=digits)[1]
+  txt <- paste(prefix, txt, sep="")
+  text(0.5, 0.5, txt, cex = 2)
+}
+set.seed(17)
+mapk <- read.delim("Downloads/mapk3.txt") %>%
+  sample_n(100) %>%
+  select(Mek1.PP, Erk2.PP, Mos.P) %>%
+  transmute(Raf = Mos.P, Mek = Mek1.PP, Erk = Erk2.PP) %>%
+  lapply(function(item) {
+    x <- jitter(item, 2000)
+    ifelse(x > 0 , x , 0)
+  }) %>%
+  as.data.frame 
+pairs(mapk, upper.panel = panel.cor)
+
+mapk2 <- mutate(mapk, Mek = ifelse(Mek > quantile(Mek)[3], "high", "low"))
+ggplot(mapk2, aes(x=Raf, y=Erk, group = Mek)) +
+  geom_point(aes(shape=Mek), size = 4)      
+ggplot(filter(mapk2, Mek == "high"), aes(x=Raf, y=Erk)) +
+  geom_point(aes(shape="circle"), size = 4)
+
 
 library(bnlearn)
-.data <- read.delim("Downloads/holygrail.txt") %>%
-  select(X.P.MAPKKK., X.PP.MAPKK., X.PP.MAPK.) %>%
-  transmute(Raf = as.numeric(X.P.MAPKKK.), 
-            Mek = as.numeric(X.PP.MAPKK.), 
-            Erk = as.numeric(X.PP.MAPK.)) %>%
-  plot
-  
-d.data <- discretize(.data, method = "hartemink", breaks = 10)
+d.data <- discretize(mapk, method = "hartemink", breaks = 8)
[…]

### 2015-09-26 scripts_for_figures.R
added changes to figures

@@ -20,12 +20,17 @@ mapk <- read.delim("Downloads/mapk3.txt") %>%
   as.data.frame 
 pairs(mapk, upper.panel = panel.cor)
 
-mapk2 <- mutate(mapk, Mek = ifelse(Mek > quantile(Mek)[3], "high", "low"))
-ggplot(mapk2, aes(x=Raf, y=Erk, group = Mek)) +
-  geom_point(aes(shape=Mek), size = 4)      
-ggplot(filter(mapk2, Mek == "high"), aes(x=Raf, y=Erk)) +
-  geom_point(aes(shape="circle"), size = 4)
+library(ggplot2)
+library(gridExtra)
 
+mapk2 <- mutate(mapk, Mek = ifelse(Mek > quantile(Mek)[3], "high", "low"))
+p1 <- ggplot(mapk2, aes(x=Raf, y=Erk, group = Mek)) +
+  geom_point(aes(shape=Mek), size = 4) + 
+  guides(shape=FALSE)
+p2 <- ggplot(filter(mapk2, Mek == "high"), aes(x=Raf, y=Erk)) +
+  geom_point(aes(shape="circle"), size = 4) + 
+  guides(shape=FALSE)
+grid.arrange(p1, p2, ncol=2)
 
 library(bnlearn)
 d.data <- discretize(mapk, method = "hartemink", breaks = 8)

### 2015-09-26 write_up.tex
final olga comments

@@
 Statistical associations between quantitative measurements can reveal an enticing number of putative causal interactions, but when do such associations reflect the underlying causal biomolecular mechanisms?
-The goal of this perspective is to provide suggestions for causal inference in large scale experiments, resulting from high throughput technologies such as mass spectrometry-based proteomics.
-We describe the pitfalls of inference in large datasets, and suggest methods to overcome these pitfalls and reliably find regulatory associations, in non-technical terms that are accessible to life scientists with a limited statistical background.
+The goal of this perspective is to provide suggestions for causal inference in large scale experiments, which utilize high throughput technologies such as mass spectrometry-based proteomics.
+We describe in non-technical terms the pitfalls of inference in large datasets, and suggest methods to overcome these pitfalls and reliably find regulatory associations.
 \end{abstract}
@@
 Causal inference \cite{pearl2009causality} elucidates statistical associations, which result from the underlying biomolecular mechanistic relationships.
-Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of such biomolecular systems on a large scale.
+Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale.
 Paradoxically, the large amount of data generated by these experiments make the task of causal inference more difficult.
@@
 This problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
-Below, we describe the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale datasets.
+Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale datasets.
 
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.4\textwidth]{figs/egfr.png}}
-\caption{EGFR MAPK signaling pathway \cite{holbro2004erbb}, an example of a pathway containing the phosphorylation cascade from Raf to Mek to Erk.
+\centerline{\includegraphics[width=.5\textwidth]{figs/egfr.png}}
+\caption{EGFR MAPK signaling pathway, an example of a pathway containing the phosphorylation cascade from Raf to Mek to Erk.
 The binding of ligand EGF to EGFR initiates a signal that leads to the cascade, which in turn regulates transcription.
@@
 Consider, e.g.
-the MAPK signaling cascade, which is part of several signaling pathways such as the EGFR MAPK pathway in Figure \ref{mapk}.
+the MAPK signaling cascade in Figure \ref{mapk}, which is part of several signaling pathways such as the EGFR MAPK pathway\cite{holbro2004erbb}.
 In this cascade Raf causally affects the level of active (i.e., phosphorylated) Mek, while Mek causally affects Erk.
@@
 
-To illustrate the process of causal inference in this context, we simulate artificial data using the computational Huang-Ferrell model of this cascade.
+To illustrate the process of causal inference in this context, we simulated artificial data using the computational Huang-Ferrell model \cite{huang1996ultrasensitivity} of this cascade.
 The model represents the key binding, phosphorylation, and dephosphorylation reactions of the cascade with mass action kinetics, and replicates the MAPK key signaling behavior observed in nature.
-With this model we simulated an experiment with 100 replicate biological samples, and measurements of concentration (umol) of phosphorylated Raf, and doubly phosphorylated Mek and Erk in each sample.
+We used the model to simulate an experiment with 100 replicate biological samples, and measurements of concentration (umol) of phosphorylated Raf, and doubly phosphorylated Mek and Erk in each sample.
 
@@
 B: Pairwise plots of concentration values of phosphorylated (doubly phosphorylated for Mek and Erk) forms of each protein, and observed Spearman correlation values.
-The Raf -- Erk correlation is high, despite the fact Raf does not directly regulate Erk.
-C: Concentrations of Raf versus Erk, where samples corresponding to high Mek are highlighted in red.
-In the right panel of B, only points with high levels of Mek are display (i.e.
-condition on Mek being high), and the observed association between Raf and Erk disappear, indicating Raf and Erk are conditionally independent given Mek.
-D: After Raf is inhibited, the observed association between Raf and Mek remains while the association between Mek and Erk disappear.
-Given Mek is conditionally dependent on Raf and on Erk, the inhibition reveals causal flow from Raf to Mek to Erk.}
-\label{mapkInference}
+The Raf -- Erk correlation is high, despite the fact that Raf does not directly regulate Erk.
+C: Concentrations of Raf versus Erk, where samples corresponding to high Mek are highlighted in circles.
+The right panel of B shows the subset of samples with high Mek (i.e.
+conditional on Mek being high).
+In these samples the association between Raf and Erk disappears, indicating that Raf and Erk are {\it conditionally independent} given Mek.
+D: After Raf is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears.
+A second simulation, which is similar to the previous one but incorporates the inhibition of Mek reveals the causal flow from Raf to Mek to Erk.
+\label{mapkInference}}
 \end{figure}
@@
 
-Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the quantified protein concentrations, and Spearman correlations to quantify the extent of the associations.
+Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the protein concentrations across the biological samples, and Spearman correlations to quantify the extent of the associations.
 The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
@@
 
-To describe Step 2 of causal inference, we need some terminology.
-In statistical language, the quantified proteins are called {\it variables}.
-If the values of concentrations of two proteins vary between the biological samples in a coordinated manner, which is due to a common biological mechanism, the proteins are {\it statistically dependent}.
-Just like the underlying biological mechanism is typically unknown, the presence of statistical dependence between two or more proteins is also unknown.
-Measures of statistical association (e.g.,  the Spearman correlation in Figure \ref{mapkInference}B or, alternatively, Pearson correlation or mutual information) demonstrate empirical evidence of the presence of such dependence.
-However, high associations are not a proof of direct causal relationships.
-In fact, the majority of statistical dependencies observed in experimental data reflect indirect associations.
+To describe Step 2 of causal inference, we introduce some terminology.
+In statistical language the quantified proteins are called {\it variables}.
+If the values of concentrations of two proteins vary between the biological samples in a coordinated manner, due to a common biological mechanism, the proteins are called {\it statistically dependent}.
+Since the underlying biological mechanism is unknown, the presence of statistical dependence between two or more proteins is also typically unknown.
+Measures of statistical association (e.g.,  the Spearman correlation in Figure \ref{mapkInference}B or, alternatively, Pearson correlation or mutual information) quantify the empirical evidence of presence of such dependence.
+However, high correlations are not proofs of direct causal relationships.
+In fact the majority of statistical dependencies observed in experimental data reflect indirect associations.
 For example, a statistical association between an upstream and an indirect downstream protein can be due to the presence of intermediate proteins, such as the statistical association between Raf -- Erk is due to their indirect relationship through Mek.
@@
 
-The direct causal events can be distinguished from other undesirable types of associations by controlling for (hold constant) the intermediaries and the common regulators.
+The direct causal events can be distinguished from other undesirable types of associations by controlling for (i.e., by holding constant) the intermediaries and the common regulators.
 The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.
@@
 We can examine the nature of statistical association between Raf and Erk.
-Figure \ref{mapkInference}C compares Raf to Erk, highlighting in red the biological samples where concentrations of Mek are high (here, set to the top quartile).
-Note that when we subset the data to only the red samples (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk.
-This has a mechanistic explanation.
+Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).
+Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk.
+This pattern has a mechanistic explanation.
 As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek.
@@
 
-When the experiment quantifies all the key variables in a biological system, Step 2 above elucidates conditional dependence-based associations between causally related variables.
+When the experiment quantifies all the key variables in a biological system, Step 2 above elucidates conditional dependence between causally related variables.
 In the MAPK example, this corresponds to a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.
 However, at this step the direction of the regulation remains unknown.
-Inference of the direction of the chain of events requires that the experimental design involves interventions.
-Figure \ref{mapkInference}D illustrates the results of Step 3, in the case of an intervention that targeted Mek with an inhibitor.
-With this intervention the concentration of Mek is unaffected, however its ability to phosphorylate other proteins is blocked.
-After this intervention is introduced the Raf--Mek relationship is unchanged, while Erk drops to a low level.
+Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses.
+Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor.
+The intervention does not affect the concentration of Mek, however it blocks its ability to phosphorylate other proteins.
+After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.
 From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.
@@
 See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
-Several of these algorithms are implemented in the R package \textit{bnlearn} \cite{scutari2009learning}.
+Numerous implementations of these algorithms are available, e.g.
+in the R package {\tt bnlearn} \cite{scutari2009learning}.
 
@@
 
-A typical high-throughput experiments includes a small number of interventions, a small number of biological replicates, and a large number of analytes.
+A typical high-throughput experiment includes a small number of interventions, a small number of biological replicates, and quantifies a large number of analytes such as proteins.
 This creates challenges in each of the 3 steps of causal inference above.
 
-In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes.
-A large number of proteins  leads to a large number of spurious statistical associations that arise without any biological justification, and purely as an artifact of random chance.
+In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes across the biological samples.
+A large number of analytes  yields a large number of spurious statistical associations, which arise without any biological justification, and are purely an artifact of random chance.
 Systematic pairwise relationships such as between Raf, Mek and Erk in the MAPK pathway will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.
@@
 We repeated each of these simulations 500 times.
-Figure \ref{spur_corr} plots the histograms of the highest Pearson correlation found across any pair of proteins in the 500 simulations, separately for each experiment.
+Figure \ref{spur_corr} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation.
 As can be seen, the experiment with 500 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlations.
-This is a clearly problem when high correlation is used as an initial evidence of a biological function.
+This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.
 
@@
 \centerline{\includegraphics[width=1\textwidth]{figs/spurious_corr.png}}
-\caption{Simulated experiments quantifying 20 (or 500) unrelated proteins in 100 biological samples, each repeated 500 times.
-The plots are the histograms of the highest Pearson correlation found across any pair of proteins, calculated over the 500 simulations.
-Increasing the number of proteins results in higher spurious associations.}
+\caption{Simulated experiments quantifying 20 (or 500) unrelated proteins in 100 biological samples.
+The histograms display the highest Pearson correlation across any pair of proteins, calculated over 500 repetitions of the simulation.
+Increasing the number of proteins results in higher values of Pearson correlation without any biological justification.}
 \label{spur_corr}
@@
 To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.
-Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003 \cite{margaritis2003learning}, which performs a series of conditional independence tests between the sets of proteins, and count the number of detected conditional dependence relationships.
-As before, since we randomly draw protein abundance measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm has no biological justification.
+Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003 \cite{margaritis2003learning} (which performs a series of conditional independence tests between the sets of proteins), and count the number of detected conditional dependence relationships.
+As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm has no biological justification.
 We again repeat these experiments 500 times.
@@
 The results demonstrate that an increase in the number of proteins leads to an increase in false positive detection of putative causal conditional dependence relationships.
-This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably detect the conditional dependence relationships needed to infer causality.
+This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the conditional dependence relationships from noise.
 
@@
 
-An additional problem for Step 2 is a typically small number of biological replicates that these experiments have as compared to the number of analytes.
-While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 requires that the number of replicates is equal or exceeds the number of analytes.
+An additional problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.
+While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 requires that the number of replicates equals or exceeds the number of analytes.
 Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins.
-This number of samples per analyte is much larger than in a typical high-throughput experiment.
-While sparse approaches working around ths problem are available, quality of their results decreases drastically as the number of replicates fall.
+While sparse approaches working around this problem are available, quality of their results decreases drastically as the numbers of replicates fall.
 
@@
 As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.
-We showed with Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships, and indeed when measuring $k$ variables a set of interventions on $< k$ variables may be sufficient to infer causality.
-But along with having too few replicates, the typical large-scale experiment has far too few interventions.
-Also, if the step 2 results in too many false positives for conditional dependence, this will adversely affect the results of step 3 regardless of the size of the set of interventions.
+We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships.
+Indeed, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes.
+Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.
+Also, if the step 2 results in too many false positives for conditional dependence, this will adversely affect the results of step 3 regardless of the size of the dataset.
 
@@
 
-The problems outlined in section II paint a grim picture for causal learning in large datasets.
-Fortunately, these can be overcome, and effective causal learning can be a reality for large scale datasets.
-We describe best practices  in the Causality Inference in Omics  below.
+The problems outlined in section II paint a grim picture for causal inference in large datasets.
+Fortunately, these can be overcome, and effective causal inference can be a reality for large scale datasets.
+We provide suggestions for the best practices below.
 
 \begin{enumerate}
-\item \textit{Limiting the number of analytes}, and use a small subset of biologically relevant and technologically accurate measurements.
-The length of the list of identified and quantified analytes is not as important as the focus of the investigation and the quality of measurements on important parts of the system.
-If the broader biological system is well understood, it may be possible design to a targeted experiment that focuses on a specific part of the system of interest, then ask more distinct questions of the data, such as whether a particular edge or pathway is present.
-The more specific the question, the less data needed overall to make solid causal inference.
-\item \textit{Measure more samples}.
-If feasible, high throughput measurements that measure many samples (not just many proteins) provide the statistical power to distinguish true associations from spurious associations.
-This intuition motivates the use of technologies that measure fewer variables but enable measurement of more samples (or gather more data points per sample), for causal modeling.
-A related  example is  use of single cell data for inference (e.g.
-from single cell  mass cytometry), where  many thousands of cells per sample provide ample statistical power.
-See Sachs et al 2005 for an in depth case study.
+\item \textit{Limit the number of analytes.} Even though a list of analytes quantifiable with high-throughput technologies grows larger, only use a subset of measurements that are both biologically relevant and technologically accurate.
+The length of the list is not as important as the quality of measurements on the key parts of the system.
+If the broader biological system is well understood, it may be possible to design a targeted experiment that focuses on a specific network or pathway, and ask more specific questions of the data, such as the presence of a particular regulatory event.
+The more specific the question, the less data are needed to make solid causal conclusions.
+
+\item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.
+This fact gives advantage to technologies that quantify fewer analytes but have a higher sample throughput.
+Examples of such technologies are targeted mass spectrometry, and single cell mass cytometry, where  many thousands of cells per sample provide ample statistical power.
+See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset.
 \cite{sachs2005causal}
-\item \textit{Use prior knowledge}, not just from experts but also from noisier sources, to improve the process of searching for conditional independence.
-The MAPK pathway for instance, mentioned earlier, is canonical and well established.
-The process of inferring causal connections is helped by assuming a priori that such canonical connections exist.
-This is because knowledge of parts of the underlying regulatory network reduces the remaining number of possible associations, and enables more effective use of data.
-The number of total associations examined goes down, enabling increased confidence in statistical associations that are found.
-Knowledge about canonical causal relationships can be sourced from pathway databases such as KEGG.
-Another example of prior knowledge is contextual information, such as spatial relations between  signaling proteins in the cell -- causal inference algorithms can be formulated to weigh evidence of conditional dependence differently depending on whether proteins are from the same or different spatial compartments.
-\item \textit{Employ targeted interventions selectively}.
-Targeted interventions perturb individual components of the system.
-An example is small molecule inhibitors which block the causal influence of a specific protein on downstream components.
-Since it is usually prohibitive to perturb every target component of a system, it is strategic to prioritize application of targeted interventions to parts of the system that have the most potential for new discoveries.
-Selection of these perturbation targets can be based on prior knowledge - e.g.
-knowledge of which components are crucial players in the system of interest - or they can be applied iteratively, after an initial statistical analysis has revealed areas of the network in which causal inferences are not possible based on existing data.
-For instance, a resulting model graph with undirected edges can be inspected to reveal which nodes have potential to reveal the most causality if perturbed.
+
+\item \textit{Use prior knowledge.} The prior knowledge improves the search for conditional independence.
+The prior knowledge can be in form of known canonical networks, extracted, e.g.
+from pathway databases such as KEGG.
+One example of such prior information is the MAPK pathway.
+The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.
+Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell.
+The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets.
+The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share a same or temporal context.
+
+\item \textit{Select targeted interventions wisely.} Targeted interventions perturb individual components of the biological system.
+An example is a small molecule inhibitor, which blocks the causal influence of a specific protein on its downstream components.
+Although effective, such targeted interventions are limited in number.
+Therefore, a strategic experimental design would use prior information, prioritize the interventions and the targets, and apply them to parts of the biological system that have most potential for new discovery of regulatory events.
+For example, a graph with undirected edges can be inspected, to reveal which nodes have potential to reveal the most causality if perturbed.
+Such targeted perturbations can be applied iteratively, after an initial statistical analysis revealed areas of the network where causal inference would benefit from extra measurements and data.
+
 \item \textit{Consider broad-scale interventions}.
-Traditionally, experiments characterize a biological response to a particular stimulus, possibly in the context of an inhibitor.
-In contrast, broad-scale interventions sacrifice specificity to perturb many features in the system simultaneously.
-The advantage of this approach is that it may enable elucidation of  causality across the entire system.
-Just as one intervention was sufficient to infer causality between three proteins (Raf, Mek, and Erk), perturbing many things at once creates cascades of causal direction orientation across the broader network.
-This includes varying experimental conditions to activate multiple pathways.
-For example, signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling.
+Broad-scale interventions sacrifice specificity of the target to simultaneously perturb many variables in a biological system.
+One example of broad-scale interventions is varying experimental conditions, in order to activate multiple pathways.
+Signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling \cite{alexopoulos2010networks}.
 Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.
-Although they do not provide specific information about the downstream effects of stimulation,  broad-scale interventions can provide more causal inference bang for your intervention buck.
+Although they do not provide specific information about the downstream effects of stimulation,  broad-scale interventions can provide more causal insight.
+Therefore, the advantage of this approach is that it may enable elucidation of  causality across the entire system.
 
@@
 
-\textit{The Causality in Omics Tool List} above provides impactful approaches
-that can drastically improve causal inference from omics datasets, by
-constraining the inference task, and thus allowing for accurate
+This list suggests impactful approaches that can drastically improve causal inference from high-thoughput experiments, by constraining the inference task, and thus allowing for accurate
 statistical inferences. For instance, the task of assessing which of all
@@
 error-prone than the task of assessing which of all possible
-combinations of my measured features might form a biological pathway.
+combinations of my measured variables might form a biological pathway.
 
@@
 combination, and in fact the lines between them are somewhat arbitrary
-and frequently blurred. For instance, using tool \#1 and tool \#2 in
+and frequently blurred. For instance, using item \#1 and item \#2 in
 concert can be thought of as reducing the breadth and increasing the
-depth of the investigation. Tools \#4 and \#5 call for use of
+depth of the investigation. Items \#4 and \#5 call for use of
 interventions, but this task itself is complicated by measuring many
-things, since we have more features to expose to intervention. Tool
+analytes. Item
 \#3, prior biological knowledge, can be used to prioritize what to

### 2015-09-27 write_up.tex
added last set of edits prior to submission

@@
 Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).
-While step 1 has little requirements, step 2 requires replicate biological samples, and step 3 requires systematic interventions (e.g.
+While step 1 has little requirements, step 2 requires multiple samples or replicates, and step 3 requires systematic interventions (e.g.
 with protein inhibitors).
@@
 
-The direct causal events can be distinguished from other undesirable types of associations by controlling for (i.e., by holding constant) the intermediaries and the common regulators.
+The direct causal events can be distinguished from other undesirable types of associations by controlling for the intermediaries and the common regulators (e.g, by selecting measurements where they have certain values).
 The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.
@@
 Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).
-Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk.
+{\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk}.
 This pattern has a mechanistic explanation.
@@
 
-\item \textit{Use prior knowledge.} The prior knowledge improves the search for conditional independence.
+\item \textit{Use prior knowledge.} Prior knowledge improves the search for conditional independence and helps to determine causality.
 The prior knowledge can be in form of known canonical networks, extracted, e.g.

### 2015-11-30 manuscript/scripts_for_figures.R
changed the cond ind figure generator to work with multiple algos

@@ -36,7 +36,6 @@ library(bnlearn)
 d.data <- discretize(mapk, method = "hartemink", breaks = 8)
 .data$dMek <- d.data$Mek    
 cdplot(dMek ~ Raf, data = .data)
-
   gs(undirected = TRUE, debug = TRUE) %>%
   graphviz.plot
   lapply(function(item) {
@@ -87,15 +86,20 @@ legend("topr", legend=c("20 proteins", "500 proteins"), cex = c(.5),  lwd=c(4,4)
 
 ## conditional dependence figure
 library(bnlearn)
-set_20 <- rep(0, 500)
-set_100 <- rep(0, 500)
-for(i in 1:length(set_20)){
-  print(i)
-  x <- as.data.frame(matrix(rnorm(100 * 20), ncol = 20))
-  set_20[i] <- narcs(gs(x, undirected = TRUE))
-  y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
-  set_100[i] <- narcs(gs(y, undirected = TRUE))
+do_sim <- function(learning_func){
+  set_20 <- rep(0, 500)
+  set_100 <- rep(0, 500)
+  for(i in 1:length(set_20)){
+    print(i)
+    x <- as.data.frame(matrix(rnorm(100 * 20), ncol = 20))
+    set_20[i] <- narcs(learning_func(x, undirected = TRUE))
+    y <- as.data.frame(matrix(rnorm(100 * 100), ncol = 100))
+    set_100[i] <- narcs(learning_func(y, undirected = TRUE))
+  }
+  return(list(set_20 = set_20, set_100 = set_100))
 }
+results <- lapply(c(gs, fast.iamb, inter.iamb), do_sim)
+names(results) <- c("gs", "fast.iamb", "inter.iamb")
 hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = "Count of false positive detections of conditional dependence",  xlim = c(min(set_20), max(set_100)), 
      main = NULL)
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)

### 2015-12-02 response2review.tex
heroku master

\documentclass{letter}[11]

\usepackage{fullpage}
\usepackage{setspace} 
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}

\usepackage{graphicx}
%\usepackage{subfig} 
\usepackage{multirow} %for multiple rows in table
\usepackage{rotating}  % for vertical text
\usepackage{array}  % for vertical alignment in a cell

\usepackage{hyperref}
\usepackage{verbatim} %comment over multiple lines

\usepackage[table]{xcolor}
\usepackage{multirow} %for multiple rows in table
\usepackage{rotating}  % for vertical text
\usepackage{color}

\usepackage{pdfpages}

\def\todo#1{{\color{red}[For Robert to finish: #1]}}
\def\ov#1{{\color{magenta}#1}}

\def\r#1{{\begin{quote}\textsf{\color{blue} #1}\end{quote}}}
\begin{document}
{
\address{
{\rm Olga Vitek} \\
{\rm Sy and Laurie Sternberg Associate Professor} \\
{\rm College of Science} \\
{\rm College of Computer and Information Science} \\
{\rm Northeastern University} \\
{\tt o.vitek@neu.edu}}
\signature{Olga Vitek}

\begin{letter}{
 		{\em Journal of Proteome Research}\\
		Attn: Susan Weintraub \\
		Editorial Office
}

\opening{Dear Dr. Weintraub:}   
\thispagestyle{empty}

We received the reviewers' comments on our manuscript {\it From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations}, and would like to thank them for the positive feedback.

On the following pages of this letter we discuss each reviewer's concerns (\textsf{\color{blue} reproduced in blue in sans serif font}), and the associated response. We hope that the reviewers will find the responses adequate, and look forward to hearing from you soon.\\ \\

\begin{tabular}{cc}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ & Sincerely,\\
\hspace{10cm} &  \\
& Olga Vitek, PhD 
\end{tabular}
\vfill

\end{letter}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\normalcolor
\noindent {\Large \bf Reviewer 1} 
\vspace{3mm} \\

\r{This paper discusses some of the difficulties of inferring networks and pathways with high-throughput data.  It is pedagogical --- methods and their short-comings are illustrated through simulation results. No new methods are proposed. The paper is well-written and should be useful for the proteomics community.}

We thank the reviewer for the positive feedback.

\r{The authors suggest single cell analysis as a means of increasing sample size. Single cell ``omics" in which all the cells come from a single biological replicate are useful for learning pathways because the cells are generally in different states, and thus provide information about dependencies within the individual. However, any study in which a single individual is observed, provides information only about that individual --- inference to a broader population must then be made either on purely theoretical grounds, or based on follow-up studies which confirm that the same relationship exists in others.}

We wholeheartedly agree with the reviewer regarding the importance of true biological replication. Indeed, this is currently insufficiently emphasized in the network inference literature. The revised manuscript now mentions \todo{in section XXX, page XXX ... rewrite in in terms of the true bioogical replication, and the fact that single sells are (heterogeneous) instances of subsampling}of the fact that while single cell data provides more data inputs to computational method, it is not sufficient for population level inference if the cells were not acquired from a sufficiently large sample of subjects from that population.

\r{p.11 line 35 ``depending on whether the analytes share a same or temporal context.'' There appears to be some missing text.}

Changed to ``depending on whether the analyses share the same spatial or temporal context''.

\r{p.12 line 38: ``my'' seems out of place}

Corrected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\normalcolor
\noindent {\Large \bf Reviewer 2} 
\vspace{3mm} \\

\r{p.1 line 7. not sure what are ``statistical associations" and that ``causal inference" deals with statistics only.}

We clarified \todo{In section XXX}  ``statistical associations'' with examples of statistical measures of association familiar to biologists, and that while ''causal inference'' spans multiple domains, we only address the specific case of causal inference through the statistical design of experiments.

\r{p.1 line 15 ``large amount of data generated by these experiments make the task of causal inference more difficult".  It would be great either to refer a publication/story where it was demonstrated, or to explain why authors think so and what they put into the word ``difficult" .}

\todo{The reviewer refers to the shortcoming of our paper, which we previously discussed, that the paper does not cite examples of network inference from high-throughput proteomic/mass spectrometry (or other types of) data. Here I would startb by adding this brief literature review, and then comment using these specific examples on the nature of difficulty. Some were successful by chance. Others had to simplify the problem. Explain why there are not many examples. Be specific when describing the difficulty. }

One of the goals of this manuscript is to demonstrate this difficulty exists and therefore argue that due to this difficulty high content low sampling throughput experiments are unsuitable for causal inference.  We agree that as stated, it seems as though we take this difficulty for granted.  We clarified the language and also added references to key concepts in statistical causal inference when they are introduced later in the manuscript.

\r{p.1 line 17 ``associations arising purely from random chance". On the other hand, this can happen (with a small probability) in an individual experiment as well, if the associations are random.}

We agree, and we modified the language to show that while even in the simplest single experiment with a single treatment and univariate response, a strong statistical association between treatment and response could have arisen spuriously.  We tied this in with the p-value, a statistical concept that addresses the problem of spurious results and which is familiar to biological investigators.

\todo{Not sure we need p-values here. I would rather say that the probability of random association is much higher in large-scale experiments}

\r{p.4 figure C. Using more different symbols in C (rather than solid triangles / circles) would certainly improve readability of the figure.}

We changed the triangles to squares, to increase readability.

\r{p.4 figure C. It seems like there is a circle with Raf=0.7, Erk=0.5 missing in C, left plot.}

The bounds on the plots were corrected.

\r{p.4 line 52. Why does the lack of correlation for samples with high Mek indicate conditional independence?}

We agree that the link between low correlation and conditional independence needs to be made more explicit.  To this end, we added to this example a statistical test for conditional independence between Raf and Erk given Mek.  The null hypothesis is conditional independence between Raf and Erk, and thus conditional correlation between Raf and Erk is a test statistic.  So when conditional correlation is lower than some statistically significant threshold, the null hypothesis cannot be rejected.  This ties the  intuition of the plots to a familiar statistical testing workflow for making decisions on the presence of conditional dependence given data.

\todo{It's a different problem. The reviewer did not understand that while the figure only shows a subset of the data as an illustration, the actual calculation involves full conditional probability distributions. Do not add any tests, but better explain the meaning of conditional independence, and how it is used.}

\r{p.4 line 52. Can it be the case that the scatterplots B and C look as in this example, but there is a direct dependence from Raf to Erk?  p.5 line 18. ``If the values of concentrations of two proteins vary between the biological samples in a coordinated manner, due to a common biological mechanism, the proteins are called statistically dependent."  Am I right that this def of the stat dependence objects your example of independence between Ref and Erk? Also, not clear to me: the mechanism of dependence between Ref and Erk is pretty ``biological" and even known (through Mek). So why can't we call Ref and Erk statistically dependent?}

We clarified the language to highlight the fact that Raf and Erk are simultaneously dependent and conditionally independent.  We also clarified the concept of a ``common biological mechanism" to mean that two proteins are affected by a common regulatory cause.  Finally, we made sure that statistical association and probabilistic dependence are clearly distinguished, and eliminated all references to ``statistical dependence" --- a potential source of confusion.

\todo{Right. The reviewer is not familiar with conditional independence and simply needs more non-technical explanation}
\todo{Spell out the definitions here - the reviewer may not read the revised manuscript, but will read the letter}

\r{p.5 line 30.  What are indirect statistical associations.}

We clarified this language to ``statistical associations resulting from indirect regulatory relationships".

\todo{Spell out the definitions here - the reviewer may not read the revised manuscript, but will read the letter}

\r{p.5 line 42 re: ``direct causal events can be distinguished from other undesirable types of associations by controlling for the intermediaries and the common regulators":  The interesting Q: when they can be distingushed? E.g. if the function from Erk is identical to Mek then it's not possible to distinguished between association Ref-Mek and Ref-Erk.}

This is true, and to address this we briefly previous work by co-author Karen Sachs with gives protein signaling examples of the conditions required to disentangle conditional dependence.

\r{p.5 line 50 ``The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called conditional independence." Hmm.}

Indeed this statement, as it was phrased, is not accurate.  Independence and conditional independence are not defined in terms of statistical associations in the data, and as mentioned above, strong statistical associations can arise even when there is independence/conditional independence.  We corrected the language here emphasized this point here and elsewhere else in the manuscript.

\todo{Do not say many words, but instead provide the correct definition right here}

\r{p.6 18 ``This phenomenon --- the disappearance of the association between Raf and Erk upon conditioning on Mek --- is evidence that Raf and Erk are conditionally independent."  Why is it evidence?}

We use the statistical test for conditional independence, described above in our response to the comment on p.4 line 52, to describe how to make decisions on the presence of conditional dependence  given data.

\todo{Do you rely on this test for network inference, or do you only do this because of the reviewer? If it's for the reviewer then do not worry about that. Just explain again in layman terms.}

\r{p.8 Figure 3.  Hm, not very convincing since the values are still pretty low, lower than 0.8 illustrated in the figure 2. Maybe increase the number of proteins to 1,000,000?}

We ran this simulation at 1 million and generated a new plot.  This resulted in higher spurious correlation values.

\r{p.9 figure 4. Hm, maybe the Margaritis algorithm is bad?}

Indeed, we agree that it is important to point out the various algorithms for conducting tests of conditional independence have different approaches, and therefore could have different results.  We regenerated the graph and overlaid the results of the Margaritis algorithm with 2 competing algorithms.

\r{p.9 line 48. At this point, it'd be great to have a figure similar to Figure 3 but that would illustrate that increasing the number of replicates helps to reduce the probability of finding a spurious dependence. p.10 line 38: ``Limit the number of analytes." Sure enough, nobody wants reducing the coverage. Can you please convince readers by providing a sort of rule of thumb? For example, ``Considering half the analytes increases the confidence in 2 times."}

To answer these points we conducted a simulation were we varied the sampling and content throughput of the input dataset, and evaluated detection of conditional dependence relationships while controlling for type 1 error.  We then generated a plot where the X axis is the ratio of sample size to coverage size, and the Y access is true positive rate.  We overlaid two curves, one where the Gaussian assumption was applied, and one where the data is assumed nonlinear.  We then highlighted ranges on the X axis that could be covered by common proteomics technologies.  This will give investigators a clear guide of what they can expect from a given technology when doing such an experiment.

\r{p.9 line 48.  What would be great as well to have a hint what to do with the replicates? Ok, I put efforts collecting data for replicates. What's next to get my results more credible? Experimental design? Which methods?  p.10 line 53. Again, it's probably not just replicates, but also the stat methods that use them in a clever way. Which methods?}

We included a remark \todo{In sec XXX} that single cells, while providing more experimental units, do not substitute for having multiple biological replicates in order to make population inferences. \todo{Again speak in terms of biological replication and subsampling}

\r{p.10 line 9 ``Indeed, a set of interventions on < k variables may be sufficient to infer causal events between k analytes" Sounds like you have a reference for this for this fact :)}

We included references on optimal experimental design for causal inference.

\r{p.11 line 38 ``4: Select targeted interventions wisely." p.12 line 6 ``5: Consider more broad scale interventions.": Do I understand correctly that 4 contradicts 5?}

\todo{Clarify that broad interventions mean single perturbations that simultaneously affect multiple nodes}
We clarified the language here, and referred to the specific classes of perturbagens that are available to investigators of cell signaling eg.  stimuli such as ligands of cytokines, genetic knockdowns, and siRNA or small molecule inhibitors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2015-12-02 manuscript/write_up.tex
heroku master

--- 2015-09-27 write_up.tex
@@
 
-Causal inference \cite{pearl2009causality} elucidates statistical associations, which result from the underlying biomolecular mechanistic relationships.
+Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations (e.g.
+Pearson correlation, Spearman correlation, mutual information), which result from the underlying biomolecular mechanistic relationships.
 Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale.
-Paradoxically, the large amount of data generated by these experiments make the task of causal inference more difficult.
-When many associations are examined, spurious associations -- associations arising purely from random chance -- may appear as strong, and the true signal may be obfuscated, leading to increased false discoveries of  putative causal events.
+We will demonstrate the paradoxical result that the large amount of data generated by these experiments make the task of causal inference more difficult.
+If only two biologically unrelated analytes were measured simultaneously, there is some probability of a spurious association -- an association between the analytes arising purely from random chance -- though the chance of this happening is low and gets lower as sample size increases.
+The difficulty of high-throughput data stems from the fact that the more analytes that are measured, the more pairwise associations between analytes are generated, resulting in a much greater chance for the appearance of spurious associations that are just as strong as associations that represent actual biological relationships.
+This obfuscates the true signal, leading to increased false discoveries of putative causal events.
 This problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
@@
 The model represents the key binding, phosphorylation, and dephosphorylation reactions of the cascade with mass action kinetics, and replicates the MAPK key signaling behavior observed in nature.
-We used the model to simulate an experiment with 100 replicate biological samples, and measurements of concentration (umol) of phosphorylated Raf, and doubly phosphorylated Mek and Erk in each sample.
+We used the model to simulate an experiment with 50 replicate biological samples, and measurements of concentration (umol) of phosphorylated Raf, and doubly phosphorylated Mek and Erk in each sample.
 
@@
 To describe Step 2 of causal inference, we introduce some terminology.
-In statistical language the quantified proteins are called {\it variables}.
-If the values of concentrations of two proteins vary between the biological samples in a coordinated manner, due to a common biological mechanism, the proteins are called {\it statistically dependent}.
-Since the underlying biological mechanism is unknown, the presence of statistical dependence between two or more proteins is also typically unknown.
-Measures of statistical association (e.g.,  the Spearman correlation in Figure \ref{mapkInference}B or, alternatively, Pearson correlation or mutual information) quantify the empirical evidence of presence of such dependence.
-However, high correlations are not proofs of direct causal relationships.
-In fact the majority of statistical dependencies observed in experimental data reflect indirect associations.
-For example, a statistical association between an upstream and an indirect downstream protein can be due to the presence of intermediate proteins, such as the statistical association between Raf -- Erk is due to their indirect relationship through Mek.
-Another such example is a case where two proteins have a common regulator.
-In this case the two proteins are statistically associated through the regulator, without causally affecting one another.
+In language of probability and statistics the quantified proteins are called {\it variables}.
+When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.
+If there is no such connection, they are called \{it independent}.
+{\it Conditional independence} is a special case that is important to causal inference.
+Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.
 
-The direct causal events can be distinguished from other undesirable types of associations by controlling for the intermediaries and the common regulators (e.g, by selecting measurements where they have certain values).
-The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.
-Causal inference searches for empirical evidence of conditional independence, to distinguish causal events from indirect associations.
+If one protein has direct causal influence on another, they are dependent, even after conditioning.
+So if we know two proteins are not conditionally independent, we can investigate whether or not they have a causal relation.
+However, we do not know the actual dependence relations between proteins, and must use statistical tests for conditional independence to infer them.
+These tests take as input the quantifications of statistical association derived in step 1, which in our example is the Spearman correlations as in Figure \ref{mapkInference}B.
+Note {\it dependence} is a property of the underlying biology, and {\it statistical association} is a property of the data generated by the biology, they are not the same.
 
@@
 Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).
-{\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk}.
-This pattern has a mechanistic explanation.
+{\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we {\it condition} on Mek being high), we can no longer visually detect the association between Raf and Erk}.
+
+Thinking in terms of a statistical test, our null hypothesis is conditional independence between Raf and Erk given Mek.
+We take the typical significance threshold of .05 --- p-values must be below this value for us to reject the null.
+We discretized the protein values into 3 levels; low, medium, and high.
+We then calculate the Pearson chi-squared test statistic -- a value that quantifies the statistical association between Raf and Erk after accounting for Mek's association with both variables (though we use Spearman correlation in the figure, chi-squared is preferred here because using the former in a conditional independence test is non-trivial).
+In this case the value of the test statistic is 10.24 with 12 degrees of freedom, which has p-value of .595 -- which doesn't meet our threshold.
+Thus we decide in favor of the null hypothesis -- the high p-value tells us we can't be confident the statistical association between Raf and Erk, beyond what is explained by Mek, did not arise spuriously.
+Tests for conditional independence between Raf and Mek given Erk, and Mek and Erk given Raf both lead to our rejecting the null (p = .034, p = .027).
+Given our decisions, the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A, a graph of putative conditional dependence relationships.
+
+This result has a mechanistic explanation.
 As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek.
 Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.
-This phenomenon -- the disappearence of the association between Raf and Erk upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.
-Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure \ref{mapk}A.
 
@@
 Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of  conditional independence between pairs of variables.
+
 Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these conditional dependencies as evidence for potential causal events.
@@
 \item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.
-This fact gives advantage to technologies that quantify fewer analytes but have a higher sample throughput.
-Examples of such technologies are targeted mass spectrometry, and single cell mass cytometry, where  many thousands of cells per sample provide ample statistical power.
+This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.
+Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population.
 See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset.
-\cite{sachs2005causal}
+\cite{sachs2005causal}.
 
@@
 The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets.
-The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share a same or temporal context.
+The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share the same spatial or temporal context.
 

### 2015-12-02 manuscript/scripts_for_figures.R

@@ -1,5 +1,24 @@
 library(dplyr)
 library(bnlearn)
+
+# Calculate MI
+
+get_mi <- function(x, y, p, base = p, ...){
+  disc <- discretize(data.frame(x, y), 
+                     method="quantile", 
+                     breaks = p) 
+  joint <- prop.table(table(disc) + .000001) 
+  x_mar <- prop.table(table(disc$x))
+  y_mar <- prop.table(table(disc$y))
+  joint %>%
+    apply(1, function(row) row / y_mar) %>%
+    apply(1, function(row) row / x_mar) %>%
+    log(base = p) %>%
+    {joint * .} %>%
+    sum
+}
+
+# Function for generating correlation panels
 panel.cor <- function(x, y, digits=3, prefix="",  ...){
   usr <- par("usr"); on.exit(par(usr))
   par(usr = c(0, 1, 0, 1))
@@ -8,13 +27,14 @@ panel.cor <- function(x, y, digits=3, prefix="",  ...){
   txt <- paste(prefix, txt, sep="")
   text(0.5, 0.5, txt, cex = 2)
 }
-set.seed(17)
+set.seed(67)
+# Generating the pairs table
 mapk <- read.delim("Downloads/mapk3.txt") %>%
-  sample_n(100) %>%
+  sample_n(60) %>%
   select(Mek1.PP, Erk2.PP, Mos.P) %>%
   transmute(Raf = Mos.P, Mek = Mek1.PP, Erk = Erk2.PP) %>%
   lapply(function(item) {
-    x <- jitter(item, 2000)
+    x <- jitter(item, 2500)
     ifelse(x > 0 , x , 0)
   }) %>%
   as.data.frame 
@@ -22,15 +42,28 @@ pairs(mapk, upper.panel = panel.cor)
 
 library(ggplot2)
 library(gridExtra)
+# Zooming in on Erk and Raf
+mapk2 <- mutate(mapk, 
+                Mek = ifelse(Mek > quantile(Mek)[4], "high", "low"),
+                Mek = factor(Mek, levels = c("derp", "low", "high")),
+                shape = ifelse(Mek == "low", "circle", "square"))
 
-mapk2 <- mutate(mapk, Mek = ifelse(Mek > quantile(Mek)[3], "high", "low"))
 p1 <- ggplot(mapk2, aes(x=Raf, y=Erk, group = Mek)) +
[…]

### 2015-12-04 manuscript/write_up.tex
incorporated changes from review

@@
 \def\todo#1{{\color{red}[TODO: #1]}}
+\def\added#1{{\color{magenta}[Added or edited: #1]}}
 
@@
 
-Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations (e.g.
-Pearson correlation, Spearman correlation, mutual information), which result from the underlying biomolecular mechanistic relationships.
+Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations \added{(e.g.
+Pearson correlation, Spearman correlation, mutual information)}, which result from the underlying biomolecular mechanistic relationships.
 Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale.
-We will demonstrate the paradoxical result that the large amount of data generated by these experiments make the task of causal inference more difficult.
+\added{Thus, the opportunity exists for the proteomics community to apply computational methods for causal inference developed for gene microarray data to proteomics data \cite{bensimon2012mass}.}
+
+We will demonstrate the paradoxical result that the large amount of data generated by high-throughput proteomics experiments make the task of causal inference more difficult.
 If only two biologically unrelated analytes were measured simultaneously, there is some probability of a spurious association -- an association between the analytes arising purely from random chance -- though the chance of this happening is low and gets lower as sample size increases.
@@
 This obfuscates the true signal, leading to increased false discoveries of putative causal events.
-This problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
-Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale datasets.
+\added{While the gene network inference community has developed approaches to dealing with this issue\cite{bensimon2012mass} , the solutions specific to the experimental conditionals and biological questions relevant to proteomics are still in early stages.} The problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
+Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale \added{proteomics} datasets.
 
@@
 To describe Step 2 of causal inference, we introduce some terminology.
-In language of probability and statistics the quantified proteins are called {\it variables}.
-When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.
-If there is no such connection, they are called \{it independent}.
+In language of \added{probability} the quantified proteins are called {\it variables}.
+\added{When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.
+If there is no such connection, they are called {\it independent}.
 {\it Conditional independence} is a special case that is important to causal inference.
-Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.
+Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.}
 
-If one protein has direct causal influence on another, they are dependent, even after conditioning.
-So if we know two proteins are not conditionally independent, we can investigate whether or not they have a causal relation.
-However, we do not know the actual dependence relations between proteins, and must use statistical tests for conditional independence to infer them.
-These tests take as input the quantifications of statistical association derived in step 1, which in our example is the Spearman correlations as in Figure \ref{mapkInference}B.
+\added{Causation and dependence are linked as follows;
+\begin{itemize}
+\item For two variables, if one is the direct cause of the other, they are dependent, even after conditioning on other variables in the system.
+\item If the two variables that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other variables in the system.
+\end{itemize}
 Note {\it dependence} is a property of the underlying biology, and {\it statistical association} is a property of the data generated by the biology, they are not the same.
+{\it Under the right measurement conditions} dependence manifests itself as statistical association in the data and application of statistical analysis can reveal cases of conditional independence (when those conditions breakdown is beyond the scope of this perspective, but Sachs and Itani et al.
+2013 \cite{sachs2013single} explores this in detail within the phospho-proteomics context).
+So if we had an oracle that revealed that two proteins are still dependent after conditioning on all others, then we'd know they were one of them was a direct causal of the other.
+Without an oracle, we must rely on statistical inference of cases where pairs of proteins were conditional independent, ignore those cases and focus on the remaining pairs as potential causal relations.
+For a more rigorous treatment of the link between mechanism, probabilistic dependence, and statistics, consult Pearl's seminal work in this area\cite{pearl2009causality}.}
 
 Let's see how this applies to the MAPK signaling cascade.
-We can examine the nature of statistical association between Raf and Erk.
-Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).
-{\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we {\it condition} on Mek being high), we can no longer visually detect the association between Raf and Erk}.
+\added{We know from the biology that Raf regulates the abundance of Erk though Mek.
+Thus Raf and Erk are dependent.
+However Erk abundance is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.
+Thus Raf and Erk are conditionally independent given Mek.
+This biology is manifests itself in the data.}  Figure \ref{mapkInference}C compares Raf to Erk, indicating in \added{solid circles} the biological samples where concentrations of Mek are high (here, set to the top quartile).
+Note that when we subset the data to only the samples with high Mek, we can no longer visually detect the association between Raf and Erk.
+\added{To understand why we'd interpret this as evidence for the presence of conditional independence between Raf and Mek, we can build intuition from a formal statistical test of conditional independence -- the null hypothesis is conditional independence between Raf and Erk given Mek, the quantification of association between Raf and Erk {\it after subsetting Mek} is the test statistic, and we decide in favor of the null hypothesis if the p-value for the test statistic (the probability an association at least as high as the one observed could have emerged spuriously) is too high.
+Indeed, the class of causal inference algorithms called constraint-based algorithms repeatedly performs this type of test over multiple pairs of variables \cite{spirtes2000causation}.
+Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.}
 
-Thinking in terms of a statistical test, our null hypothesis is conditional independence between Raf and Erk given Mek.
-We take the typical significance threshold of .05 --- p-values must be below this value for us to reject the null.
-We discretized the protein values into 3 levels; low, medium, and high.
-We then calculate the Pearson chi-squared test statistic -- a value that quantifies the statistical association between Raf and Erk after accounting for Mek's association with both variables (though we use Spearman correlation in the figure, chi-squared is preferred here because using the former in a conditional independence test is non-trivial).
-In this case the value of the test statistic is 10.24 with 12 degrees of freedom, which has p-value of .595 -- which doesn't meet our threshold.
-Thus we decide in favor of the null hypothesis -- the high p-value tells us we can't be confident the statistical association between Raf and Erk, beyond what is explained by Mek, did not arise spuriously.
-Tests for conditional independence between Raf and Mek given Erk, and Mek and Erk given Raf both lead to our rejecting the null (p = .034, p = .027).
-Given our decisions, the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A, a graph of putative conditional dependence relationships.
-
-This result has a mechanistic explanation.
-As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek.
-Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.
-
-When the experiment quantifies all the key variables in a biological system, Step 2 above elucidates conditional dependence between causally related variables.
-In the MAPK example, this corresponds to a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.
-However, at this step the direction of the regulation remains unknown.
+At Step 2 in the figure, the direction of the regulation remains unknown.
 Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses.
@@
 From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.
-With the intervention, we can finally move from the conditional dependence graph in panel A - step 2 to the causal graph in panel A - step 3.
+With the intervention, we can finally move from the undirected graph in panel A - step 2 to the causal graph in panel A - step 3.
 
@@
 Step 1 creates a dense network of pairwise associations.
-Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of  conditional independence between pairs of variables.
-
-Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these conditional dependencies as evidence for potential causal events.
+\added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.
 See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
-Numerous implementations of these algorithms are available, e.g.
+Numerous implementations of these algorithms are available in statistical software, e.g.
 in the R package {\tt bnlearn} \cite{scutari2009learning}.
@@
 Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.
-The spurious correlations result in more false positives when detecting putative causal conditional dependence relationships.
+If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}.
 To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.
-Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003 \cite{margaritis2003learning} (which performs a series of conditional independence tests between the sets of proteins), and count the number of detected conditional dependence relationships.
-As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm has no biological justification.
+Instead of finding the highest spurious correlation between pairs of proteins, we apply an example of type of causal inference algorithm that performs a series of conditional independence tests between the sets of proteins (constraint-based algorithms\cite{spirtes2000causation}), and count the number of pairs proteins \added{where the null hypothesis of conditional independence was rejected}.
+As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, \added{no test should ever be rejected, and any instance of rejection by this algorithm would have biological justification}.
 We again repeat these experiments 500 times.
-Figure \ref{spur_dep} shows the histograms of the counts of false positive detections of conditional dependence.
-The results demonstrate that an increase in the number of proteins leads to an increase in false positive detection of putative causal conditional dependence relationships.
-This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the conditional dependence relationships from noise.
+Figure \ref{spur_dep} shows the histograms of the counts of false positives.
+The figure graphs the results of the Grow-Shrink algorithm \cite{margaritis2003learning}, \added{but we also repeated the simulation with the IAMB algorithm\cite{tsamardinos2003algorithms} and Inter-IAMB algorithm\cite{yaramakala2005speculative} and achieved similar results}.
+The results demonstrate that an increase in the number of proteins leads to an increase in false positives.
+This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.
 
@@
 \centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
-\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positive detections of conditional dependence.}
+\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positives (rejecting the null hypothesis of conditional independence.}
 \label{spur_dep}
@@
 
-An additional problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.
-While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 requires that the number of replicates equals or exceeds the number of analytes.
-Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins.
-While sparse approaches working around this problem are available, quality of their results decreases drastically as the numbers of replicates fall.
+A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.
+While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 relies on having many replicates relative to the number analytes.
+Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high.
+The quality of causal inference approaches decreases drastically as the numbers of replicates fall.
 
-Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional dependence.
+Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional independence.
 Quantifying a large number of analytes introduces a  challenge for this step as well.
@@
 We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships.
-Indeed, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes.
+\added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes .
 Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.
-Also, if the step 2 results in too many false positives for conditional dependence, this will adversely affect the results of step 3 regardless of the size of the dataset.
+Also, if the step 2 results in too many false positives for potential causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.
 
@@
 This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.
-Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population.
+\added{A review by Twerve and Saez-Rodriguez provides a contrast of technologies by their content and sampling throughput\cite{SaezRodriguez:2012kx}.}
+
+\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}.
 See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset.
@@
 The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.
-Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell.
+\added{Work by Saez-Rodriguez, Lauffenburger, and Sorger\cite{SaezRodriguez:2009hb}, as well as Terfve and Saez-Rodriguez\cite{Terfve:2015kw} use prior network knowledge to build logic models that reflect causal relationships between signaling proteins using proteomics measurements, the former using antibody/bead-based XMAP technology from Luminex, and the latter using LC-MS/MS.} Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell.
 The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets.
-The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share the same spatial or temporal context.
+The causal inference algorithms can be extended to weigh evidence of conditional independence, depending on whether the analytes share the same spatial or temporal context.
 
 \item \textit{Select targeted interventions wisely.} Targeted interventions perturb individual components of the biological system.
-An example is a small molecule inhibitor, which blocks the causal influence of a specific protein on its downstream components.
+An example is \added{siRNA knockdowns, as well as} small molecule inhibitors, which block the causal influence of a specific protein on its downstream components.
 Although effective, such targeted interventions are limited in number.
@@
 \item \textit{Consider broad-scale interventions}.
-Broad-scale interventions sacrifice specificity of the target to simultaneously perturb many variables in a biological system.
+Broad-scale interventions sacrifice specificity of targeted interventions to simultaneously perturb many variables in a biological system.
 One example of broad-scale interventions is varying experimental conditions, in order to activate multiple pathways.

### 2015-12-09 manuscript/write_up.tex
olga todos

@@
 \def\todo#1{{\color{red}[TODO: #1]}}
-\def\added#1{{\color{magenta}[Added or edited: #1]}}
+\def\added#1{{\color{blue} #1}}
+\def\removed#1{{\color{magenta}[Removed: #1]}}
+\def\note#1{{\color{cyan}[Note on edit: #1]}}
 
@@
   Causal inference -- the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of many high throughput investigations.
-Statistical associations between quantitative measurements can reveal an enticing number of putative causal interactions, but when do such associations reflect the underlying causal biomolecular mechanisms?
+Statistical associations between quantitative measurements can suggest an enticing number of hypotheses regarding the underlying causal interactions.
+But when do such associations reflect the underlying causal biomolecular mechanisms?
 The goal of this perspective is to provide suggestions for causal inference in large scale experiments, which utilize high throughput technologies such as mass spectrometry-based proteomics.
@@
 
-Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations \added{(e.g.
-Pearson correlation, Spearman correlation, mutual information)}, which result from the underlying biomolecular mechanistic relationships.
-Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale.
-\added{Thus, the opportunity exists for the proteomics community to apply computational methods for causal inference developed for gene microarray data to proteomics data \cite{bensimon2012mass}.}
+%\note{Olga's comment: Modern experiments generate large scale data
+%Correlation. Not enough. Causal inference elucidates.... or causal relationships  Examples are XXX However this is difficult because XXX}
 
-We will demonstrate the paradoxical result that the large amount of data generated by high-throughput proteomics experiments make the task of causal inference more difficult.
-If only two biologically unrelated analytes were measured simultaneously, there is some probability of a spurious association -- an association between the analytes arising purely from random chance -- though the chance of this happening is low and gets lower as sample size increases.
-The difficulty of high-throughput data stems from the fact that the more analytes that are measured, the more pairwise associations between analytes are generated, resulting in a much greater chance for the appearance of spurious associations that are just as strong as associations that represent actual biological relationships.
-This obfuscates the true signal, leading to increased false discoveries of putative causal events.
-\added{While the gene network inference community has developed approaches to dealing with this issue\cite{bensimon2012mass} , the solutions specific to the experimental conditionals and biological questions relevant to proteomics are still in early stages.} The problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
-Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale \added{proteomics} datasets.
+\added{Modern high-throughput technologies such as mass spectrometry simultaneously quantify hundreds or even thousands of biomolecular analytes.
+Statistical associations (e.g.
+Pearson correlation, Spearman correlation, and mutual information) between the analytes  can suggest an enticing number of hypotheses regarding the underlying causal biomolecular mechanism.
+However, associations do not imply causation.
+Formal methods of causal inference \cite{pearl2009causality} \todo{Could you also reference http://www.biomedcentral.com/1471-2105/8/S6/S5} are required to probe these statistical associations for causality, and infer the underlying regulatory mechanisms.
+}
+
+\added{There is an increasing interest in causal inference in proteome research.
+It has previously been used to learn the directed structure of signal transduction networks, e.g.
+in flow cytometry investigations in Sachs et al.
+2005 \cite{sachs2013single}, antibody/bead-based XMAP technology from Luminex in Saez-Rodriguez et al.
+2009\cite{saez2009discrete} and others \todo{what do you mean by 'others'?
+Other technologies?
+Other authors?
+What did they do?} \cite{prill2011crowdsourcing}.
+However, the successful examples of causal inference in proteomics are very few \cite{bensimon2012mass}.
+In this perspective we demonstrate that this is not an accident, because large numbers of quantified analytes can make the task of causal inference more difficult.
+The difficulty stems from the fact that large numbers of quantified analytes  lead to more spurious pairwise associations, obfuscate the true signal, and increase the false discoveries of putative causal events.
+Below we describe in non-technical terms the process of elucidating causal associations from high-throughput data, and suggest practical approaches for causal inference in large scale proteomic datasets.
+Specifically, we suggest that the task of causal inference can be facilitated by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses.
+}
 
@@
 The Raf -- Erk correlation is high, despite the fact that Raf does not directly regulate Erk.
-C: Concentrations of Raf versus Erk, where samples corresponding to high Mek are highlighted in circles.
-The right panel of B shows the subset of samples with high Mek (i.e.
+C: Concentrations of Raf versus Erk, where samples corresponding to high Mek (here, set to the top quartile) are highlighted \added{with filled circles}.
+The right panel of C shows the subset of samples with high Mek (i.e.
 conditional on Mek being high).
-In these samples the association between Raf and Erk disappears, indicating that Raf and Erk are {\it conditionally independent} given Mek.
+In these samples the association between Raf and Erk disappears, \added{and we infer} that Raf and Erk are {\it conditionally independent} given Mek.
 D: After Raf is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears.
 A second simulation, which is similar to the previous one but incorporates the inhibition of Mek reveals the causal flow from Raf to Mek to Erk.
+\todo{Where is this last simulation shown?
+If it's not shown, add (not shown) at the end of the sentence.}
 \label{mapkInference}}
@@
 In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.
-Step 2 reduces the correlation graph to a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).
-Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).
+Step 2 reduces the correlation graph to a sparse graph of \added{inferred} underlying conditional dependencies (Raf--Mek, and Mek--Erk).
+Step 3 interrogates this graph to find putative directions of causal relationships (Raf-->Mek, and Mek-->Erk).
 While step 1 has little requirements, step 2 requires multiple samples or replicates, and step 3 requires systematic interventions (e.g.
@@
 The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.
-What about the noncausal Raf--Erk edge?
+What about the non-causal Raf--Erk edge?
 Despite the high Raf--Erk correlation, there is no direct causal mechanism between them (aside from the one via Mek, which is already accounted for via the Raf-->Mek and Mek-->Erk edges).
-In causal  inference, our goal is to eliminate this "nuisance" edge.
+In causal  inference, our goal is to eliminate this ``nuisance" edge.
 How is this done?
 
-To describe Step 2 of causal inference, we introduce some terminology.
-In language of \added{probability} the quantified proteins are called {\it variables}.
-\added{When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.
-If there is no such connection, they are called {\it independent}.
+\added{To describe Step 2 of causal inference, we introduce some terminology.
+If concentrations of two proteins vary between the biological samples in a coordinated manner, such that knowing the behavior of one protein provides information on the other, the two proteins are called {\it dependent}.
+Otherwise they are called {\it independent}.
 {\it Conditional independence} is a special case that is important to causal inference.
-Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.}
+Two {\it dependent} proteins are {\it conditionally independent} if, after knowing (in probability language, {\it conditioning on}) the behavior of third-party biomolecules, the two proteins become independent.
+In other words, knowing the concentration of one protein provides no information about the other after considering the information from the third party.}
 
-\added{Causation and dependence are linked as follows;
-\begin{itemize}
-\item For two variables, if one is the direct cause of the other, they are dependent, even after conditioning on other variables in the system.
-\item If the two variables that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other variables in the system.
-\end{itemize}
-Note {\it dependence} is a property of the underlying biology, and {\it statistical association} is a property of the data generated by the biology, they are not the same.
-{\it Under the right measurement conditions} dependence manifests itself as statistical association in the data and application of statistical analysis can reveal cases of conditional independence (when those conditions breakdown is beyond the scope of this perspective, but Sachs and Itani et al.
-2013 \cite{sachs2013single} explores this in detail within the phospho-proteomics context).
-So if we had an oracle that revealed that two proteins are still dependent after conditioning on all others, then we'd know they were one of them was a direct causal of the other.
-Without an oracle, we must rely on statistical inference of cases where pairs of proteins were conditional independent, ignore those cases and focus on the remaining pairs as potential causal relations.
-For a more rigorous treatment of the link between mechanism, probabilistic dependence, and statistics, consult Pearl's seminal work in this area\cite{pearl2009causality}.}
+\added{While statistical associations and correlations are properties of the observed data, dependence and conditional independence are properties of the underlying processes that generate the data.
+Step 2 relies on statistical inference\cite{pearl2009causality} to infer from the observed data pairs of proteins that are conditionally independent.
+The conditionally independent pairs are ignored, and the remaining pairs are kept as potential causal relations.}
 
 Let's see how this applies to the MAPK signaling cascade.
-\added{We know from the biology that Raf regulates the abundance of Erk though Mek.
-Thus Raf and Erk are dependent.
-However Erk abundance is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.
+\added{Since Raf regulates the concentration of Erk, Raf and Erk are dependent.
+However, since the concentration of Erk is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.
 Thus Raf and Erk are conditionally independent given Mek.
-This biology is manifests itself in the data.}  Figure \ref{mapkInference}C compares Raf to Erk, indicating in \added{solid circles} the biological samples where concentrations of Mek are high (here, set to the top quartile).
-Note that when we subset the data to only the samples with high Mek, we can no longer visually detect the association between Raf and Erk.
-\added{To understand why we'd interpret this as evidence for the presence of conditional independence between Raf and Mek, we can build intuition from a formal statistical test of conditional independence -- the null hypothesis is conditional independence between Raf and Erk given Mek, the quantification of association between Raf and Erk {\it after subsetting Mek} is the test statistic, and we decide in favor of the null hypothesis if the p-value for the test statistic (the probability an association at least as high as the one observed could have emerged spuriously) is too high.
-Indeed, the class of causal inference algorithms called constraint-based algorithms repeatedly performs this type of test over multiple pairs of variables \cite{spirtes2000causation}.
-Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.}
+}  Figure \ref{mapkInference}C illustrates the process of statistical inference by comparing the concentrations of Raf and Erk.
+When we subset the measurements to only the samples with high Mek, we can no longer see the association between Raf and Erk.
+Formally, the algorithms repeatedly test the null hypothesis of conditional independence between Raf and Erk given Mek, and makes the decision based on p-values\cite{spirtes2000causation}.
+Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.
 
-At Step 2 in the figure, the direction of the regulation remains unknown.
+\added{At Step 2 in the figure,} the direction of the regulation remains unknown.
 Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses.
@@
 
-In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related variables.
+\added{Causation and dependence are linked as follows;
+\begin{itemize}
+\item For two proteins, if one is the direct cause of the other, they are dependent, even after conditioning on other proteins in the system.
+\item If the two proteins that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other proteins in the system.
+\end{itemize}}
+
+In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related proteins.
 Step 1 creates a dense network of pairwise associations.
-\added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.
+\removed{Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of conditional independence between pairs of proteins.} \added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.
 See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{II. Large-scale statistical inference of causal relationships: challenges of scaling up}
@@
 al \cite{fan2014challenges} but translated to our context.
-First, we simulated an experiment that quantifies the abundances of 20 proteins in 100 biological samples.
+First, we simulated an experiment that quantifies the concentrations of 20 proteins in 100 biological samples.
 Second, we simulates another experiment where the number of proteins was increased to 500.
@@
 We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships.
-\added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes .
+\added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ proteins may be sufficient to infer causal events between $k$ analytes .
 Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.
@@
 \item \textit{Consider broad-scale interventions}.
-Broad-scale interventions sacrifice specificity of targeted interventions to simultaneously perturb many variables in a biological system.
+Broad-scale interventions sacrifice specificity of targeted interventions to simultaneously perturb many proteins in a biological system.
 One example of broad-scale interventions is varying experimental conditions, in order to activate multiple pathways.
@@
 error-prone than the task of assessing which of all possible
-combinations of my measured variables might form a biological pathway.
+combinations of my measured proteins might form a biological pathway.
 

### 2015-12-09 manuscript/scripts_for_figures.R

@@ -116,7 +116,7 @@ for(i in 1:m){
 hist(set_500, main = NULL, col = rgb(0.1,0.1,0.1,0.5), xlim = c(min(set_20) , max(set_500) + .1 ),
      xlab = "Highest measured correlation between protein pairs")
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("topr", legend=c("20 proteins", "500 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+legend("topright", legend=c("20 proteins", "500 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
 
 ## conditional dependence figure
 library(bnlearn)
@@ -132,12 +132,132 @@ do_sim <- function(learning_func){
   }
   return(list(set_20 = set_20, set_100 = set_100))
 }
-results <- lapply(c(gs, fast.iamb, inter.iamb), do_sim)
-names(results) <- c("gs", "fast.iamb", "inter.iamb")
-hist(set_100, col = rgb(0.1,0.1,0.1,0.5), xlab = "Count of false positive detections of conditional dependence",  xlim = c(min(set_20), max(set_100)), 
-     main = NULL)
-hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("top", legend=c("20 proteins", "100 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+#results <- lapply(c(gs, fast.iamb, inter.iamb), do_sim)
+#names(results) <- c("gs", "fast.iamb", "inter.iamb")
+results1 <- list(gs = do_sim(gs))
+hist(results1[["gs"]][["set_100"]], 
+     col = rgb(0.1,0.1,0.1,0.5), 
+     xlab = "Count of false positive detections of conditional dependence",  
+     xlim = c(0, 140),
+     ylim = c(0, 130),
+     main = NULL, freq = T)
+hist(results1[["gs"]][["set_20"]], 
+     col = rgb(0.8,0.8,0.8,0.5), 
+     add = TRUE,
+     freq = T)
[…]

### 2015-12-17 manuscript/write_up.tex
revamped simulation

@@
 \author{Robert O. Ness}
-\email{nessr@purdue.edu}
 \affiliation[Purdue University]{Purdue University Department of Statistics, West Lafayette}
@@
 \author{Olga Vitek}
+\email{o.vitek@neu.edu}
 \affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
@@
 However, associations do not imply causation.
-Formal methods of causal inference \cite{pearl2009causality} \todo{Could you also reference http://www.biomedcentral.com/1471-2105/8/S6/S5} are required to probe these statistical associations for causality, and infer the underlying regulatory mechanisms.
+Formal methods of causal inference \cite{pearl2009causality,markowetz2007inferring} are required to probe these statistical associations for causality, and infer the underlying regulatory mechanisms.
 }
@@
 2005 \cite{sachs2013single}, antibody/bead-based XMAP technology from Luminex in Saez-Rodriguez et al.
-2009\cite{saez2009discrete} and others \todo{what do you mean by 'others'?
-Other technologies?
-Other authors?
-What did they do?} \cite{prill2011crowdsourcing}.
+2009\cite{saez2009discrete} and participants in the DREAM4 Predictive Signaling Network Challenge\cite{prill2011crowdsourcing}.
 However, the successful examples of causal inference in proteomics are very few \cite{bensimon2012mass}.
-In this perspective we demonstrate that this is not an accident, because large numbers of quantified analytes can make the task of causal inference more difficult.
-The difficulty stems from the fact that large numbers of quantified analytes  lead to more spurious pairwise associations, obfuscate the true signal, and increase the false discoveries of putative causal events.
+In this perspective we demonstrate that this is not an accident.
+The difficulty stems from the fact that large numbers of quantified analytes, combined with small sample size, lead to more spurious pairwise associations, obfuscate the true signal, and increase the false discoveries of putative causal events.
 Below we describe in non-technical terms the process of elucidating causal associations from high-throughput data, and suggest practical approaches for causal inference in large scale proteomic datasets.
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=1\textwidth]{figs/mapk.png}}
+\centerline{\includegraphics[width=.95\textwidth]{figs/mapk.png}}
 \caption{A: Overview of the 3 steps of causal inference, illustrated for the MAPK signaling cascade.
@@
 In these samples the association between Raf and Erk disappears, \added{and we infer} that Raf and Erk are {\it conditionally independent} given Mek.
-D: After Raf is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears.
-A second simulation, which is similar to the previous one but incorporates the inhibition of Mek reveals the causal flow from Raf to Mek to Erk.
-\todo{Where is this last simulation shown?
-If it's not shown, add (not shown) at the end of the sentence.}
+D:  After Mek is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears.
+This reveals the causal flow from Raf to Mek.
 \label{mapkInference}}
@@
 In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.
-Step 2 reduces the correlation graph to a sparse graph of \added{inferred} underlying conditional dependencies (Raf--Mek, and Mek--Erk).
+Step 2 reduces the correlation graph to a \added{sparse graph of inferred} underlying conditional dependencies (Raf--Mek, and Mek--Erk).
 Step 3 interrogates this graph to find putative directions of causal relationships (Raf-->Mek, and Mek-->Erk).
@@
 Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the protein concentrations across the biological samples, and Spearman correlations to quantify the extent of the associations.
-The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
+\added{We chose Spearman correlation instead of Pearson correlation because it best quantifies nonlinear relationships.}  The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.
 The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.
@@
 \added{To describe Step 2 of causal inference, we introduce some terminology.
-If concentrations of two proteins vary between the biological samples in a coordinated manner, such that knowing the behavior of one protein provides information on the other, the two proteins are called {\it dependent}.
+If concentrations of two proteins vary between the biological samples in a coordinated manner, such that knowing the concentration of one protein provides information on the concentration of the other, the two proteins are called {\it dependent}.
 Otherwise they are called {\it independent}.
 {\it Conditional independence} is a special case that is important to causal inference.
-Two {\it dependent} proteins are {\it conditionally independent} if, after knowing (in probability language, {\it conditioning on}) the behavior of third-party biomolecules, the two proteins become independent.
-In other words, knowing the concentration of one protein provides no information about the other after considering the information from the third party.}
+Two {\it dependent} proteins are {\it conditionally independent} if, after knowing (in probability language, {\it conditioning on}) the concentration of third-party biomolecules, the two proteins become independent.
+In other words, after considering the information from the third party, the behavior of one protein provides no additional information on the behavior of the other.}.
 
 \added{While statistical associations and correlations are properties of the observed data, dependence and conditional independence are properties of the underlying processes that generate the data.
-Step 2 relies on statistical inference\cite{pearl2009causality} to infer from the observed data pairs of proteins that are conditionally independent.
-The conditionally independent pairs are ignored, and the remaining pairs are kept as potential causal relations.}
+Step 2 relies on statistical inference\cite{pearl2009causality,markowetz2007inferring} to infer from the observed data pairs of proteins that are conditionally independent.
+The conditionally independent pairs are ignored, and the remaining pairs are kept as hypothesized causal relations.
+This sparse inferred conditional independence graph is often desirable, as it reduces the many pairwise associations to a smaller number of hypothesized causal relations.}
 
 Let's see how this applies to the MAPK signaling cascade.
-\added{Since Raf regulates the concentration of Erk, Raf and Erk are dependent.
-However, since the concentration of Erk is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.
-Thus Raf and Erk are conditionally independent given Mek.
+\added{Since Raf regulates the concentration of Erk by way of regulating Mek, Raf and Erk are dependent.
+However, if we know the concentration of Mek, then the concentration of Raf provides no additional information about the concentration of Erk.
+Therefore, even though Raf and Erk are dependent, they are also conditionally independent given Mek.
 }  Figure \ref{mapkInference}C illustrates the process of statistical inference by comparing the concentrations of Raf and Erk.
 When we subset the measurements to only the samples with high Mek, we can no longer see the association between Raf and Erk.
-Formally, the algorithms repeatedly test the null hypothesis of conditional independence between Raf and Erk given Mek, and makes the decision based on p-values\cite{spirtes2000causation}.
-Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.
+\added{Formally, the algorithm tests the null hypothesis of conditional independence between Raf and Erk given the full range of values of Mek (and not just high values of Mek, as was shown for the purposes of illustration), and evaluates evidence against the null hypothesis\cite{spirtes2000causation}.
+In this example the test did not reject the null hypothesis, and resulted in removing the edge between Raf and Erk as in the middle graph of Figure \ref{mapk}A.
+}
 
-\added{At Step 2 in the figure,} the direction of the regulation remains unknown.
-Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses.
+At Step 2 the direction of the regulation remains unknown.
+Inference of the direction of the chain of events requires the experimental design, which involves external interventions or stresses.
 Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor.
@@
 After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.
-From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.
-With the intervention, we can finally move from the undirected graph in panel A - step 2 to the causal graph in panel A - step 3.
-
-\added{Causation and dependence are linked as follows;
-\begin{itemize}
-\item For two proteins, if one is the direct cause of the other, they are dependent, even after conditioning on other proteins in the system.
-\item If the two proteins that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other proteins in the system.
-\end{itemize}}
+From this we can infer that Mek has causal influence on Erk.
+Since Raf was unaffected by the intervention, Mek does not have a causal influence on Raf, and therefore the direction of causal influence in this edge goes from Raf to Mek.
+This intervention is required to transform the undirected graph in panel A - Step 2 to the causal graph in panel A - Step 3.
 
@@
 Step 1 creates a dense network of pairwise associations.
-\removed{Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of conditional independence between pairs of proteins.} \added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.
-See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
-Numerous implementations of these algorithms are available in statistical software, e.g.
+\added{Step 2 identifies cases of conditional independence, and removes the corresponding edges to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.
+See Koller and Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings.
+Numerous implementations are available in statistical software, e.g.
 in the R package {\tt bnlearn} \cite{scutari2009learning}.
+\added{Depending on the biological system and on the experimental setting, the strength of causal evidence may vary.
+For example, Sachs, Itani {\it et al.} 2013\cite{sachs2013single} highlight conditions in phosho-proteomic experiments  where it may be infeasible to disentangle causality using perturbations.}
 
@@
 
-A typical high-throughput experiment includes a small number of interventions, a small number of biological replicates, and quantifies a large number of analytes such as proteins.
-This creates challenges in each of the 3 steps of causal inference above.
+\added{Experimentalists can choose to quantify more or less analytes, and collect more or less replicates such as proteins.} A typical high-throughput experiment includes \removed{a small number of interventions,} a small number of biological replicates, and quantifies a large number of analytes \removed{such as proteins}.
+\removed{This creates challenges in each of the 3 steps of causal inference above.}  \added{This experimental approach is an obstacle to causal inference.}
+
+\added{To understand why, recall that we are interested in paring down the correlation graph in step one to the sparse conditional independence graph in step 2.
+A {\it true positive} is the case when we correctly keep an edge from step 1 to step 2 in cases where the proteins are not conditionally independent.
+Incorrectly keeping an edge when the proteins are actually conditionally independent is a case of a false positive.
+Since the edges we preserve imply a regulatory relationship and motivate the expense of a validation experiment, false positives are costly and should be minimized.
+However, the number of false positives increases as an exponential rate as one increases the number of proteins, much faster than the number of true positives.}
 
@@
 al \cite{fan2014challenges} but translated to our context.
-First, we simulated an experiment that quantifies the concentrations of 20 proteins in 100 biological samples.
-Second, we simulates another experiment where the number of proteins was increased to 500.
-In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
-In other words, we do not expect any biologically meaningful associations in these data.
-We repeated each of these simulations 500 times.
-Figure \ref{spur_corr} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation.
-As can be seen, the experiment with 500 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlations.
-This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.
+We simulate data from a conditional independence network of 60 proteins and evaluate performance of structure inference algorithms in recovering this structure.
+We then simulate a second experiment where the number of proteins is increased 99, to show how the increased protein coverage affects the performance of structure inference.
+We varied depth as well as breadth -- In order to evaluate the effect of increasing sample size we start with 100 replicates, then repeat the two experiments with 200 replicates.
+We evaluate performance by counting the total number of false positives (reject conditional independence for two conditionally independent proteins) and true positives (correct rejections).
+
+The simplest case of for testing conditional independence is one of three variables -- two to compare, and at one variable upon which to condition the comparison.
+Thus in the simulate we generate networks composed of three-protein subnetworks comprised of protein A, B, and C.
+C's measurements come a univariate Gaussian distribution with 0 mean and variance 1.
+Proteins A and B also have 0 mean and variance of 1, but their dependence with C is quantified in the multivariate Gaussian covariance matrix, each has a covariance with C of .3.
+Most importantly, A and B are conditionally independent of each other given C.
+So each subnetwork has two edges, from A to C and from B to C, and no edge from A to B is missing.
+We created a network of 60 proteins within 20 disjoint subnetworks, 40 edges in total, and simulated multivariate Gaussian data with the covariance structure determined by this network.
+
+We examined two cases for expanding the number of measured proteins.
+The first describes the case of measuring proteins that provide no additional information about the biology.
+We model this scenario by adding 39 unconnected proteins with independent Gaussian values for the second experiment.
+Any inferred edge in the condition independence network linking these two proteins would be a false positive.
+We then look at the case where more proteins are added, but the added proteins are informative about the biology.
+We model this by increasing the number of triplets from 20 to 33, again to 99 proteins in total.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=1\textwidth]{figs/spurious_corr.png}}
-\caption{Simulated experiments quantifying 20 (or 500) unrelated proteins in 100 biological samples.
-The histograms display the highest Pearson correlation across any pair of proteins, calculated over 500 repetitions of the simulation.
-Increasing the number of proteins results in higher values of Pearson correlation without any biological justification.}
+\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr2.png}}
+\caption{Simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.
+This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected.
+The histograms display the distributions of the maximum correlation values.Parts A and B indicates the case of randomly adding completely independent proteins to the panel, parts C and D illustrated informed addition of proteins -- the added proteins contain interdependencies.
+The left column (A and C) show results for 30 replicates, the right (B and D) for 60 replicates.
+Comparing A and B to C and D demonstrate increasing the number of proteins on the panel creates more spurious correlation, and this is unaffected by whether or not the addition of proteins are added in an informative way.
+This is a problem because correlation is used as initial evidence for a biological relationship.
+Comparing the left and right columns show increasing the number of replicates can mitigates this somewhat.}
 \label{spur_corr}
@@
 
-Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.
-If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}.
-To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.
-Instead of finding the highest spurious correlation between pairs of proteins, we apply an example of type of causal inference algorithm that performs a series of conditional independence tests between the sets of proteins (constraint-based algorithms\cite{spirtes2000causation}), and count the number of pairs proteins \added{where the null hypothesis of conditional independence was rejected}.
-As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, \added{no test should ever be rejected, and any instance of rejection by this algorithm would have biological justification}.
-We again repeat these experiments 500 times.
-Figure \ref{spur_dep} shows the histograms of the counts of false positives.
-The figure graphs the results of the Grow-Shrink algorithm \cite{margaritis2003learning}, \added{but we also repeated the simulation with the IAMB algorithm\cite{tsamardinos2003algorithms} and Inter-IAMB algorithm\cite{yaramakala2005speculative} and achieved similar results}.
-The results demonstrate that an increase in the number of proteins leads to an increase in false positives.
-This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.
+Figure \ref{spur_corr} shows illustrates simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.
+This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected.
+The histograms display the distributions of the maximum correlation values.
+The top row (figure \ref{spur_corr} parts A and B) indicates the case if randomly adding proteins to the panel -- the added proteins where completely independent of one another protein in the network.
+The bottom row (C and D) indicates the case of informed addition of proteins to the panel -- the added proteins contain interdependencies.
+The first column (A and C) show results for 30 replicates, the second (B and D) for 60 replicates.
+Parts A and B show that an increase in proteins leads to more high correlations, however since the added proteins were completely independent, the higher correlations were completely spurious.
+Further, comparing parts C and D to A and B show that even when addition of proteins is informed by biology, the results are the same.
+Spurious correlation increases with the number of proteins whether or not the addition of proteins is informed by biological prior knowledge.
+This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.
+Comparing the left and right columns show that increasing the number of replicates can mitigate this situation someone, but as we will see next, false positives will remain a problem.
+
+This increased incidence of spurious correlations impedes Step 2, which \added{infers} conditional independences in the data.
+To illustrate this, we repeated the previous simulation, but this time in each iteration we \added{performed Step 1 by calculating pairwise \added{Pearson} correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.}   \added{These algorithms have stricter data requirements that correlation calculations, so we changed settings to 60/100 samples, and 60/99 proteins to meet these requirements.}
+
+Figure \ref{spur_dep} shows the counts of false positive edges in the resulting conditional independence graph, i.e.
+the number of protein pairs \added{where the null hypothesis of conditional independence was unduly rejected}.
+The results demonstrate that an increase in the number of proteins leads to an increase in false positive rejections of conditional independence, and to a denser conditional independence graph.
+\added{Again, comparing the top and bottom rows in figure \ref{spur_dep} show that it doesn't matter if the addition of proteins to the panel is informed or not.
+Increasing the number of proteins results in an exponential increase in pairwise comparisons -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of pairs.
+So though the algorithm controls the rate of false positives, the absolute number of false positives increases exponentially.
+Increasing the number of replicates does not help -- paradoxically in this case it hurt, most likely because the additional degrees of freedom enabled more tests, some of which added to the false positive count.} In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.
+\added{ To verify that this pattern is not the artifact of a particular protein inference algorithm we repeated the simulation with two comparable algorithms\cite{tsamardinos2003algorithms,yaramakala2005speculative}, and arrived to the same conclusion (results not shown).}
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
-\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positives (rejecting the null hypothesis of conditional independence.}
+\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep.png}}
+\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples.
+The histograms display the counts of false positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation.
+The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.
+The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.
+Increasing the number of proteins results in more false positives without any biological justification.
+Informed addition and added replicates does not help.}
 \label{spur_dep}
@@
 
-A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.
+\added{These results lead to obvious question.
+What exactly is the use of having more replicates?
+How does an informed selection of proteins for the panel make a difference relative to random selection?
+These factors come to play in detection of true positives.
+Figure\ref{true positives} is the true positive counterpart to figure\ref{true_pos}.
+Given the described subnetwork approach we described above, the 60 protein network is comprised of 20 subnetworks of 3 proteins, each with 2 edges each for a total of 40 relationships that the algorithm is tasked with detecting.
+That upper bound is illustrated with the dashed line in figure\ref{true_pos} parts A and B.
+Uninformative addition of proteins adds no new relationships.
+In parts C and D, informed addition of proteins adds 26 new relationships, indicated by the double-dashed line.
+Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.
+Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.
+More importantly, the scale of the increase in total true positives gained from informed addition of proteins to the panel was less than that of the increase of false positives.
+The costs of spurious correlation and resulting false positives quickly outweighs the benefits of higher throughput}
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=.8\textwidth]{figs/true_pos.png}}
+\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples.
+The histograms display the counts of true positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation.
+The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.
+The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.
+Dashed lines indicate the total amount of positives the algorithm is tasked with discovering.
+Increasing the number of replicates increases true positives and is indeed essential when increasing the number of proteins increases the number of relationships to be detected.}
+\label{spur_dep}
+\end{figure}
+
+\removed{A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.
+Once again the purpose of step 2 is to test conditional independence, and detect relationships between proteins that remain even when conditioning on other proteins.
+While measuring too many analytes causes an explosion in false positives, collecting too few samples results in insufficient power to detect these relationships.
 While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 relies on having many replicates relative to the number analytes.
-Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high.
-The quality of causal inference approaches decreases drastically as the numbers of replicates fall.
+Note that even when the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high.
+The quality of causal inference approaches decreases drastically as the numbers of replicates fall.}
 
-Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional independence.
-Quantifying a large number of analytes introduces a  challenge for this step as well.
-As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.
-We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships.
-\added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ proteins may be sufficient to infer causal events between $k$ analytes .
-Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.
-Also, if the step 2 results in too many false positives for potential causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.
+\added{Finally, the challenge of completing step 3 in the workflow also increases with the number of proteins quantified.}  Step 3 requires interventions to infer causal relationships from evidence of conditional independence.
+\added{For a panel of k analytes, theoretical work by Eberhardt and Scheines {\it et al.} \cite{eberhardt2012number} show that in least ideal case of the only interventions available being interventions that target a single analyte (such as with small molecule inhibitors), k - 1 separate interventions would be needed.
+With application of interventions that simultaneously target multiple analytes, this number can be lowered significantly, but this depends on several different factors\cite{eberhardt2012number,hauser2012two,hyttinen2013experiment} beyond the scope of this discussion.} \removed{For example, in the Raf-Mek-Erk pathway one intervention on Mek was sufficient to infer the two causal relationships.
+However, scaling this approach to a large number of proteins also introduces.}  But generally speaking as the number of proteins grows, so does the number of interventions required to fully infer causality, and eventually the total number of required interventions may become infeasible.
+Moreover, if the step 2 results in too many false positive hypothesized causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.
 
@@
 \begin{enumerate}
-\item \textit{Limit the number of analytes.} Even though a list of analytes quantifiable with high-throughput technologies grows larger, only use a subset of measurements that are both biologically relevant and technologically accurate.
+\item \textit{Limit the number of analytes.} \added{Reducing the number of analytes dramatically reduces the number of spurious associations and resulting false positives.} Even though a list of analytes quantifiable with high-throughput technologies grows larger, only use a subset of measurements that are both biologically relevant and technologically accurate.
 The length of the list is not as important as the quality of measurements on the key parts of the system.
@@
 
-\item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.
+\item \textit{Profile more biological replicates.}  \added{While reducing the number of analytes reduces spurious associations (false positives), increasing the number of replicates increases statistical power (true positives).}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population.
 This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.
@@
 
-\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}.
+\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide much more data as an input to causal inference, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}.
 See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset.

### 2015-12-17 manuscript/scripts_for_figures.R

@@ -99,7 +99,7 @@ gs(mapk) %>%
 
 
 # R scripts for generating spurious correlation and conditional dependence histograms.
-## correlation figure
+## correlation figure at 20 reps
 m <- 1000
 set_20 <- rep(0, m)
 set_500 <- rep(0, m)
@@ -114,9 +114,40 @@ for(i in 1:m){
   set_500[i] <- max(cor_y)
 }
 hist(set_500, main = NULL, col = rgb(0.1,0.1,0.1,0.5), xlim = c(min(set_20) , max(set_500) + .1 ),
-     xlab = "Highest measured correlation between protein pairs")
+     xlab = "Highest observed correlation between unrelated protein pairs")
 hist(set_20, col = rgb(0.8,0.8,0.8,0.5), add = TRUE)
-legend("topright", legend=c("20 proteins", "500 proteins"), cex = c(.5),  lwd=c(4,4), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+legend("topright", legend=c("20 proteins", "500 proteins"), cex = c(.6),  lwd=c(8,8), col=c(rgb(0.8,0.8,0.8,0.5), rgb(0.1,0.1,0.1,0.5)))
+
+## Generalization of the above sim
+cor_sim <- function(n, p1, p2, m){
+  set_1 <- rep(0, m)
+  set_2 <- rep(0, m)
+  for(i in 1:m){
+    print(i)
+    x <- matrix(rnorm(n * p1), ncol = p1)
+    cor_x <- cor(x)
+    diag(cor_x) <- 0
+    set_1[i] <- max(cor_x)
+    y <- matrix(rnorm(n * p2), ncol = p2)
+    cor_y <- cor(y)
+    diag(cor_y) <- 0
+    set_2[i] <- max(cor_y)
+  }
+  list(set_1 = set_1, set_2 = set_2)
+}
+m <- 500
+n <- 50
+p1 <- 20
+p2 <- 5000
+r <- cor_sim(n, p1, p2, m)
+hist(r$set_2, main = NULL, col = rgb(0.1,0.1,0.1,0.5), xlim = c(min(r$set_1) , max(r$set_2) + .1 ),
+     freq = F,
+     xlab = "Highest observed correlation between unrelated protein pairs")
+hist(r$set_1, col = rgb(0.8,0.8,0.8,0.5), freq = F, add = TRUE)
[…]

### 2015-12-18 manuscript/write_up.tex
a bit late in the game

@@
 
-\added{To understand why, recall that we are interested in paring down the correlation graph in step one to the sparse conditional independence graph in step 2.
-A {\it true positive} is the case when we correctly keep an edge from step 1 to step 2 in cases where the proteins are not conditionally independent.
-Incorrectly keeping an edge when the proteins are actually conditionally independent is a case of a false positive.
-Since the edges we preserve imply a regulatory relationship and motivate the expense of a validation experiment, false positives are costly and should be minimized.
-However, the number of false positives increases as an exponential rate as one increases the number of proteins, much faster than the number of true positives.}
-
-In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes across the biological samples.
+\added{To understand why, recall that we are interested in paring down the correlation graph in step 1 to the sparse conditional independence graph in step 2.}  In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes across the biological samples.
 A large number of analytes yields a large number of spurious statistical associations, which arise without any biological justification, and are purely an artifact of random chance.
@@
 We illustrate this problem with a computer simulation, inspired by Fan et.
-al \cite{fan2014challenges} but translated to our context.
-We simulate data from a conditional independence network of 60 proteins and evaluate performance of structure inference algorithms in recovering this structure.
-We then simulate a second experiment where the number of proteins is increased 99, to show how the increased protein coverage affects the performance of structure inference.
-We varied depth as well as breadth -- In order to evaluate the effect of increasing sample size we start with 100 replicates, then repeat the two experiments with 200 replicates.
-We evaluate performance by counting the total number of false positives (reject conditional independence for two conditionally independent proteins) and true positives (correct rejections).
-
-The simplest case of for testing conditional independence is one of three variables -- two to compare, and at one variable upon which to condition the comparison.
-Thus in the simulate we generate networks composed of three-protein subnetworks comprised of protein A, B, and C.
-C's measurements come a univariate Gaussian distribution with 0 mean and variance 1.
-Proteins A and B also have 0 mean and variance of 1, but their dependence with C is quantified in the multivariate Gaussian covariance matrix, each has a covariance with C of .3.
-Most importantly, A and B are conditionally independent of each other given C.
-So each subnetwork has two edges, from A to C and from B to C, and no edge from A to B is missing.
-We created a network of 60 proteins within 20 disjoint subnetworks, 40 edges in total, and simulated multivariate Gaussian data with the covariance structure determined by this network.
-
-We examined two cases for expanding the number of measured proteins.
-The first describes the case of measuring proteins that provide no additional information about the biology.
-We model this scenario by adding 39 unconnected proteins with independent Gaussian values for the second experiment.
-Any inferred edge in the condition independence network linking these two proteins would be a false positive.
-We then look at the case where more proteins are added, but the added proteins are informative about the biology.
-We model this by increasing the number of triplets from 20 to 33, again to 99 proteins in total.
+al\cite{fan2014challenges} but translated to our context.
+First, we simulated an experiment that quantifies the abundances of 60 proteins in 60 biological samples.
+Second, we simulated another experiment where the number of proteins was increased from 60 to 6000.
+In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution.
+In other words, we do not expect any biologically meaningful associations in these data.
+We repeated each of these simulations 500 times.
+Figure \ref{spurious_corr.png} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation.
+As can be seen, the increase to 6000 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlation.
+This is clearly a problem when high Pearson correlation is used as an initial evidence of biological function.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr2.png}}
-\caption{Simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.
-This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected.
-The histograms display the distributions of the maximum correlation values.Parts A and B indicates the case of randomly adding completely independent proteins to the panel, parts C and D illustrated informed addition of proteins -- the added proteins contain interdependencies.
-The left column (A and C) show results for 30 replicates, the right (B and D) for 60 replicates.
-Comparing A and B to C and D demonstrate increasing the number of proteins on the panel creates more spurious correlation, and this is unaffected by whether or not the addition of proteins are added in an informative way.
-This is a problem because correlation is used as initial evidence for a biological relationship.
-Comparing the left and right columns show increasing the number of replicates can mitigates this somewhat.}
+\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr.png}}
+\caption{Histogram of maximum observed correlation values between unrelated protein pairs in a simulation of 500 experiments.
+In each iteration,  first 60 proteins are quantified, after which throughput is increased to 6000 proteins.
+Increasing the number of proteins increases spurious correlation.
+This is a problem because correlation is used as initial evidence for a biological relationship.}
 \label{spur_corr}
@@
 
-Figure \ref{spur_corr} shows illustrates simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.
-This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected.
-The histograms display the distributions of the maximum correlation values.
-The top row (figure \ref{spur_corr} parts A and B) indicates the case if randomly adding proteins to the panel -- the added proteins where completely independent of one another protein in the network.
-The bottom row (C and D) indicates the case of informed addition of proteins to the panel -- the added proteins contain interdependencies.
-The first column (A and C) show results for 30 replicates, the second (B and D) for 60 replicates.
-Parts A and B show that an increase in proteins leads to more high correlations, however since the added proteins were completely independent, the higher correlations were completely spurious.
-Further, comparing parts C and D to A and B show that even when addition of proteins is informed by biology, the results are the same.
-Spurious correlation increases with the number of proteins whether or not the addition of proteins is informed by biological prior knowledge.
-This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.
-Comparing the left and right columns show that increasing the number of replicates can mitigate this situation someone, but as we will see next, false positives will remain a problem.
+Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.
+If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}.
+To illustrate, \added{we performed a second simulation}, this time we included some dependence between proteins so we could evaluate the effect of dimensionality of the data on true positives (pairs of proteins that are not conditionally independent) as well as false positives.
 
-This increased incidence of spurious correlations impedes Step 2, which \added{infers} conditional independences in the data.
-To illustrate this, we repeated the previous simulation, but this time in each iteration we \added{performed Step 1 by calculating pairwise \added{Pearson} correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.}   \added{These algorithms have stricter data requirements that correlation calculations, so we changed settings to 60/100 samples, and 60/99 proteins to meet these requirements.}
+\added{To build dependence into the simulated data, we created multiple sets of 3-proteins "triplets" A, B, and C, where all three are dependent but A and B are conditionally independent given C.
+Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^th$ triplet to be independent.
+We made it such that the proteins in each triplet generate measurements with Gaussian values with mean 0 and variance 1, and that the covariance between A--C and B--C was .3.
+We simulated data from a multivariate Gaussian.
+}
+
+\added{Again we run the simulation 500 times, in each iteration we performed step 1 by calculating pairwise Pearson correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.
+To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative}.
+The results did not differ (results not shown).}
+
+\added{These algorithms have stricter data requirements that correlation calculations, so in this simulation we increase from 60 proteins to only 99.
+The starting 60 proteins contains 20 triplets, and thus perfect performance of the algorithm would yield 40 true positives (and of course 0 false positives).
+In a high-throughput experiment, the investigator may arbitrarily select proteins for quantification, hoping the more proteins are quantified, the higher the chance of uncovering biology.
+Alternatively, he or she may apply their prior knowledge about the system to select proteins that are strong candidates for being involved in the biology.
+We respectively term these approaches "uninformed addition" and "informed addition" of proteins.
+We simulate "uniformed addition" but adding 39 proteins to the original 60 that are completely independent.
+In this case perfect performance is still 40 true positives.
+In the second "informed addition" case, we add 13 new triplets, increasing the perfect performance standard from 40 to 66.
+In addition to comparing these two strategies, we compare the 60 replicate case to the 100 replicate case, to show what role sample size has in affecting performance.}
 
@@
 The results demonstrate that an increase in the number of proteins leads to an increase in false positive rejections of conditional independence, and to a denser conditional independence graph.
-\added{Again, comparing the top and bottom rows in figure \ref{spur_dep} show that it doesn't matter if the addition of proteins to the panel is informed or not.
-Increasing the number of proteins results in an exponential increase in pairwise comparisons -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of pairs.
-So though the algorithm controls the rate of false positives, the absolute number of false positives increases exponentially.
-Increasing the number of replicates does not help -- paradoxically in this case it hurt, most likely because the additional degrees of freedom enabled more tests, some of which added to the false positive count.} In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.
-\added{ To verify that this pattern is not the artifact of a particular protein inference algorithm we repeated the simulation with two comparable algorithms\cite{tsamardinos2003algorithms,yaramakala2005speculative}, and arrived to the same conclusion (results not shown).}
+\added{This occurs because the number of false positives depends on the total number of pairwise comparison -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of comparisons.
+In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.}
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep.png}}
-\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples.
-The histograms display the counts of false positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation.
-The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.
+\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep2.png}}
+\caption{Histograms of false positives detection of relationships between proteins (rejections of tests for conditional independence) over 500 simulated experiment.
+Simulated experiments start with quantifying 60 and then expand to 99.
+The initial 60 proteins contains 40 pairwise relationships that the algorithm is based with detection.
+In parts A and B, 39 proteins are added that have no relationship with any other protein.
+In parts C and D, 39 proteins are added that have 26 additional pairwise relationships.
 The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.
 Increasing the number of proteins results in more false positives without any biological justification.
-Informed addition and added replicates does not help.}
+Informed addition of proteins mitigates this only slightly.
+Adding replicates paradoxically increases false positives.
+In both cases there are less replicates than pairwise comparisons, so the extra replicates add degrees of freedom which enables more tests, some of which add to the total false positives.}
 \label{spur_dep}
@@
 
-\added{These results lead to obvious question.
-What exactly is the use of having more replicates?
-How does an informed selection of proteins for the panel make a difference relative to random selection?
-These factors come to play in detection of true positives.
-Figure\ref{true positives} is the true positive counterpart to figure\ref{true_pos}.
-Given the described subnetwork approach we described above, the 60 protein network is comprised of 20 subnetworks of 3 proteins, each with 2 edges each for a total of 40 relationships that the algorithm is tasked with detecting.
-That upper bound is illustrated with the dashed line in figure\ref{true_pos} parts A and B.
-Uninformative addition of proteins adds no new relationships.
-In parts C and D, informed addition of proteins adds 26 new relationships, indicated by the double-dashed line.
-Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.
-Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.
-More importantly, the scale of the increase in total true positives gained from informed addition of proteins to the panel was less than that of the increase of false positives.
-The costs of spurious correlation and resulting false positives quickly outweighs the benefits of higher throughput}
+\added{Comparing the panels figure \ref{spur_dep} reveals a noteworthy result; informed addition of proteins offers only a slight reduction in false positives on average.
+Increasing sample size doesn't help much either, paradoxically in this case it hurt, because even at 100 replicates there are still  more pairwise comparisons than replicates, so the algorithm uses every last degree of freedom at its disposal.
+Increasing the number of replicates from 60 to 100 therefore allowed more conditional independence tests to take place, and some of these additional tests added to the false positive count.}
 
@@
 Increasing the number of replicates increases true positives and is indeed essential when increasing the number of proteins increases the number of relationships to be detected.}
-\label{spur_dep}
+\label{true_pos}
 \end{figure}
+
+\added{So what exactly is the use of having more replicates?
+How does an informed selection of proteins for the panel make a difference relative to random selection?
+These factors come to play in detection of true positives.
+Figure \ref{true_pos} is the true positive counterpart to figure\ref{spur_dep}.
+Parts A and B illustrate the 40 true positives upper bound on performance with the dashed line, and in parts C and D, where "informed addition" of proteins is applied, the increase upper bound of 66 is pictured with the double-dashed line.
+Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.
+Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.
+However, when comparing parts C and D in figure \ref{true_pos} to C and D in figure \ref{spur_dep}, we see that the scale of the increase in total true positives gained from informed addition of proteins was less than that of the increase of false positives.
+The spurious correlation and resulting false positives in high throughput experiments quickly outweigh the benefit that informed selection of proteins and greater sample size has on discovery.}
 

### 2015-12-18 manuscript/write_up.tex
tweaked responses

@@
 We repeated each of these simulations 500 times.
-Figure \ref{spurious_corr.png} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation.
+Figure \ref{spur_corr} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation.
 As can be seen, the increase to 6000 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlation.
@@
 \added{To build dependence into the simulated data, we created multiple sets of 3-proteins "triplets" A, B, and C, where all three are dependent but A and B are conditionally independent given C.
-Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^th$ triplet to be independent.
+Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^{th}$ triplet to be independent.
 We made it such that the proteins in each triplet generate measurements with Gaussian values with mean 0 and variance 1, and that the covariance between A--C and B--C was .3.
@@
 \added{Again we run the simulation 500 times, in each iteration we performed step 1 by calculating pairwise Pearson correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.
-To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative}.
+To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative} .
 The results did not differ (results not shown).}

### 2015-12-18 manuscript/scripts_for_figures.R

@@ -4,10 +4,10 @@ library(bnlearn)
 # Calculate MI
 
 get_mi <- function(x, y, p, base = p, ...){
-  disc <- discretize(data.frame(x, y), 
-                     method="quantile", 
-                     breaks = p) 
-  joint <- prop.table(table(disc) + .000001) 
+  disc <- discretize(data.frame(x, y),
+                     method="quantile",
+                     breaks = p)
+  joint <- prop.table(table(disc) + .000001)
   x_mar <- prop.table(table(disc$x))
   y_mar <- prop.table(table(disc$y))
   joint %>%
@@ -37,29 +37,29 @@ mapk <- read.delim("Downloads/mapk3.txt") %>%
     x <- jitter(item, 2500)
     ifelse(x > 0 , x , 0)
   }) %>%
-  as.data.frame 
+  as.data.frame
 pairs(mapk, upper.panel = panel.cor)
 
 library(ggplot2)
 library(gridExtra)
 # Zooming in on Erk and Raf
-mapk2 <- mutate(mapk, 
+mapk2 <- mutate(mapk,
                 Mek = ifelse(Mek > quantile(Mek)[4], "high", "low"),
                 Mek = factor(Mek, levels = c("derp", "low", "high")),
                 shape = ifelse(Mek == "low", "circle", "square"))
 
 p1 <- ggplot(mapk2, aes(x=Raf, y=Erk, group = Mek)) +
-  geom_point(aes(shape=shape), size = 4) + 
+  geom_point(aes(shape=shape), size = 4) +
   scale_shape_manual(name = "Mek", labels = c("low", "high"), values = c(1, 16)) +
   guides(shape=FALSE)
 p2 <- ggplot(filter(mapk2, Mek == "high"), aes(x=Raf, y=Erk)) +
-  geom_point(aes(shape="circle"), size = 4) + 
+  geom_point(aes(shape="circle"), size = 4) +
   guides(shape=FALSE)
 grid.arrange(p1, p2, ncol=2)
 
 # CI Test for Erk and Raf
-.data <- discretize(mapk, 
-           method="quantile", 
+.data <- discretize(mapk,
+           method="quantile",
            breaks = 3)
 ci.test("Erk", "Raf", "Mek", data = .data, test = "x2", debug = T)
 ci.test("Raf", "Mek", "Erk", data = .data, test = "x2", debug = T)
[…]
